---
title: "2000HIV_Co-receptor_Forestplot_ORs"
author: "Kavita Mehta"
date: "2026-01-20"
output: html_document
---

#Setup chunk (runs once, shared by all later chunks)
```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
library(readxl)
library(dplyr)
library(tibble)
library(openxlsx)
library(ggplot2)
library(cowplot)
library(gridExtra)
library(purrr)
library(stringr)
library(tidyr)
library(scales)
library(reshape2)
})

# ---- Paths ----

file_path <- "S:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx"
base_out  <- "S:/hcrc/Personal folders/Kavita Mehta/Manuscripts/2000HIV_Tropism"

# ---- Read once ----

df_raw <- read_excel(file_path)

# ---- Common cohort filter functions ----

prep_r5x4 <- function(df) {
df %>%
filter(Tropism %in% c("R5","X4"),
CONTROLLER_1B != "EC_persistent") %>%
mutate(Tropism = factor(Tropism, levels = c("R5","X4")))
}


# ---- Derived columns (shared) ----

add_derived_cols <- function(df) {
df %>%
mutate(
lo_vs_rest = if_else(ViReTgroup == "LoViReT", 1, 0, missing = NA_real_),
hi_vs_rest = if_else(ViReTgroup == "HiViReT", 1, 0, missing = NA_real_),
hi_vs_lo   = case_when(
ViReTgroup == "HiViReT" ~ 1,
ViReTgroup == "LoViReT" ~ 0,
TRUE ~ NA_real_
),
CDC3 = if_else(HIV_STAGE_CDC == 2, 1L, 0L, missing = NA_integer_)
) %>%
mutate(
Steatosis_yesno = case_when(
is.na(Steatosis_yesno) ~ NA_character_,
trimws(Steatosis_yesno) %in% c("S1","S2","S3","S1-S3") ~ "S1-S3",
trimws(Steatosis_yesno) == "S0" ~ "S0",
TRUE ~ NA_character_
),
Steatosis_yesno = factor(Steatosis_yesno, levels = c("S0","S1-S3"))
)
}

# ---- Helper: coerce outcome to 0/1 when needed ----

coerce_binary01 <- function(x){
if (is.numeric(x)) return(x)
if (is.factor(x) && nlevels(x) == 2L) return(as.integer(x) - 1L)
if (is.logical(x)) return(as.integer(x))
if (is.character(x)) {
ux <- sort(unique(na.omit(x)))
if (length(ux) == 2L) return(as.integer(factor(x, levels = ux)) - 1L)
}
suppressWarnings(as.numeric(x))
}

fmt_p <- function(p){
ifelse(is.na(p), NA_character_,
ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}


```


#correlation heatmap of the variables
```{r}
# ---- Covariate correlation heatmap with BH-adjusted significance ----

# Variables to include
covars_heatmap <- c(
  "SMOKING_CURRENT",
  "CD4_NADIR",
  "CD4_PRECART",
  "SEX_BIRTH",
  "BMI_BASELINE",
  "TIMETOCART",
  "DM",
  "AGE",
  "CART_DURATION",
  "HT",
  "HC",
  "VL_ZENITH")

# Prepare data
df_cor <- df_raw %>%
  prep_r5x4() %>%
  add_derived_cols() %>%
  select(all_of(covars_heatmap))

# Convert variables to numeric for correlation
to_numeric_for_cor <- function(x) {
  if (is.numeric(x)) return(as.numeric(x))
  if (is.logical(x)) return(as.numeric(x))
  
  if (is.factor(x)) {
    if (nlevels(x) == 2) return(as.numeric(x) - 1)
    return(as.numeric(x))
  }
  
  if (is.character(x)) {
    x_trim <- trimws(x)
    
    if (all(na.omit(unique(x_trim)) %in% c("0", "1"))) {
      return(as.numeric(x_trim))
    }
    
    ux <- sort(unique(na.omit(x_trim)))
    if (length(ux) == 2) {
      return(as.numeric(factor(x_trim, levels = ux)) - 1)
    }
    
    x_num <- suppressWarnings(as.numeric(x_trim))
    if (!all(is.na(x_num))) return(x_num)
    
    return(as.numeric(factor(x_trim)))
  }
  
  suppressWarnings(as.numeric(x))
}

df_cor_num <- df_cor %>%
  mutate(across(everything(), to_numeric_for_cor))

# Convert to base numeric matrix
mat_cor <- as.matrix(df_cor_num)
storage.mode(mat_cor) <- "numeric"

# Spearman correlation matrix
cor_mat <- cor(mat_cor, use = "pairwise.complete.obs", method = "spearman")

# Function to compute raw p-values
cor_pmat <- function(mat) {
  n <- ncol(mat)
  p_mat <- matrix(NA_real_, n, n)
  colnames(p_mat) <- colnames(mat)
  rownames(p_mat) <- colnames(mat)
  
  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      x <- as.numeric(mat[, i])
      y <- as.numeric(mat[, j])
      keep <- complete.cases(x, y)
      
      if (sum(keep) >= 3) {
        p_mat[i, j] <- suppressWarnings(
          cor.test(x[keep], y[keep], method = "spearman", exact = FALSE)$p.value
        )
      } else {
        p_mat[i, j] <- NA_real_
      }
    }
  }
  p_mat
}

p_mat <- cor_pmat(mat_cor)

# ---- Adjust p-values using only unique pairs from one triangle ----
p_adj_mat <- matrix(NA_real_, nrow = nrow(p_mat), ncol = ncol(p_mat),
                    dimnames = dimnames(p_mat))

# lower triangle including diagonal
lt_idx <- lower.tri(p_mat, diag = TRUE)

# adjust only non-NA values from lower triangle
p_vals_lt <- p_mat[lt_idx]
non_na_lt <- !is.na(p_vals_lt)

p_adj_lt <- rep(NA_real_, length(p_vals_lt))
p_adj_lt[non_na_lt] <- p.adjust(p_vals_lt[non_na_lt], method = "BH")

# put adjusted values back into lower triangle
p_adj_mat[lt_idx] <- p_adj_lt

# mirror to upper triangle so lookup works everywhere
p_adj_mat[upper.tri(p_adj_mat)] <- t(p_adj_mat)[upper.tri(p_adj_mat)]

# Convert to long format
cor_long   <- reshape2::melt(cor_mat,   varnames = c("Var1", "Var2"), value.name = "r")
p_long     <- reshape2::melt(p_mat,     varnames = c("Var1", "Var2"), value.name = "p_raw")
padj_long  <- reshape2::melt(p_adj_mat, varnames = c("Var1", "Var2"), value.name = "p_adj")

df_plot <- cor_long %>%
  left_join(p_long,    by = c("Var1", "Var2")) %>%
  left_join(padj_long, by = c("Var1", "Var2"))

# Keep order
df_plot$Var1 <- factor(df_plot$Var1, levels = rownames(cor_mat))
df_plot$Var2 <- factor(df_plot$Var2, levels = colnames(cor_mat))

# Show labels only on one half; asterisk based on adjusted p-value
df_plot <- df_plot %>%
  mutate(
    x_id = as.numeric(Var2),
    y_id = as.numeric(Var1),
    label = ifelse(
      y_id >= x_id,
      paste0(
        sprintf("%.2f", r),
        ifelse(!is.na(p_adj) & p_adj < 0.05, "*", "")
      ),
      NA_character_
    )
  )

# Plot
p_cor <- ggplot(df_plot, aes(x = Var2, y = Var1, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = label), size = 3, na.rm = TRUE) +
  scale_fill_gradient2(
    low = "cyan2",
    mid = "white",
    high = "hotpink",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Corr"
  ) +
  coord_fixed() +
  labs(title = "Covariate correlation heatmap", x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    panel.grid = element_blank()
  )

# Display
p_cor

# Save PDF
ggsave(
  filename = file.path(base_out, "Covariate_Correlation_Heatmap.pdf"),
  plot = p_cor,
  width = 8,
  height = 7
)

# Save TIFF
ggsave(
  filename = file.path(base_out, "Covariate_Correlation_Heatmap.tiff"),
  plot = p_cor,
  width = 8,
  height = 7,
  dpi = 600,
  compression = "lzw"
)

# ---- Save correlation results to Excel ----

df_export <- df_plot %>%
  mutate(
    Var1 = as.character(Var1),
    Var2 = as.character(Var2)
  ) %>%
  select(Var1, Var2, r, p_raw, p_adj) %>%
  arrange(Var1, Var2)

df_export_lower <- df_export %>%
  mutate(
    x_id = as.numeric(factor(Var2, levels = colnames(cor_mat))),
    y_id = as.numeric(factor(Var1, levels = rownames(cor_mat)))
  ) %>%
  filter(y_id >= x_id) %>%
  select(-x_id, -y_id)

wb <- createWorkbook()

addWorksheet(wb, "All_pairs")
writeData(wb, "All_pairs", df_export)

addWorksheet(wb, "Lower_triangle")
writeData(wb, "Lower_triangle", df_export_lower)

addWorksheet(wb, "Correlation_matrix")
writeData(wb, "Correlation_matrix", round(cor_mat, 3), rowNames = TRUE)

addWorksheet(wb, "Raw_p_matrix")
writeData(wb, "Raw_p_matrix", signif(p_mat, 3), rowNames = TRUE)

addWorksheet(wb, "Adj_p_matrix")
writeData(wb, "Adj_p_matrix", signif(p_adj_mat, 3), rowNames = TRUE)

saveWorkbook(
  wb,
  file = file.path(base_out, "Covariate_Correlation_Results.xlsx"),
  overwrite = TRUE
)
```

#covariates correction for variables (different sets of covariates)
```{r}
setwd(file.path(base_out, "covariates"))

df <- df_raw %>% prep_r5x4() %>% add_derived_cols()

filter_suffix <- paste0(
"Tropism_",
paste(levels(df$Tropism), collapse = "_"),
"_Excl_EC"
)

is_binary_col <- function(x) {
u <- unique(na.omit(x))
all(as.character(u) %in% c("0","1")) && length(u) == 2
}
all_cols    <- names(df)
binary_cols <- all_cols[sapply(df, is_binary_col)]
df[binary_cols] <- lapply(df[binary_cols], function(x) as.numeric(as.character(x)))
continuous_cols <- names(df)[sapply(df, is.numeric) & !names(df) %in% binary_cols]

covariate_sets <- list(
NONE                = character(0),
AGE_SEX_TIMETOcART  = c("AGE","SEX_BIRTH","TIMETOCART")
)


for (cov_name in names(covariate_sets)) {
covs       <- covariate_sets[[cov_name]]
results_ls <- list()

for (col in c(binary_cols, continuous_cols)) {
vars_needed <- c(col, "Tropism", covs)
df_sub <- df %>%
select(all_of(vars_needed)) %>%
filter(!is.na(.data[[col]])) %>%
na.omit()

if (length(unique(df_sub$Tropism)) < 2) next

if (length(covs) > 0) {
  facs <- covs[sapply(df_sub[covs], \(x) is.factor(x) || is.character(x))]
  if (length(facs) > 0) {
    lvl_counts <- sapply(facs, \(v) length(unique(df_sub[[v]])))
    if (any(lvl_counts < 2)) next
  }
}

N_R5 <- sum(df_sub$Tropism == "R5")
N_X4 <- sum(df_sub$Tropism == "X4")

if (col %in% binary_cols) {
  n1_R5   <- sum(df_sub[[col]] == 1 & df_sub$Tropism == "R5", na.rm = TRUE)
  n1_X4   <- sum(df_sub[[col]] == 1 & df_sub$Tropism == "X4", na.rm = TRUE)
  pct1_R5 <- ifelse(N_R5 > 0, n1_R5 / N_R5 * 100, NA_real_)
  pct1_X4 <- ifelse(N_X4 > 0, n1_X4 / N_X4 * 100, NA_real_)
} else {
  n1_R5 <- n1_X4 <- NA_integer_
  pct1_R5 <- pct1_X4 <- NA_real_
}

resp <- paste0("`", col, "`")
rhs  <- paste(c("Tropism", covs), collapse = " + ")
fml  <- as.formula(paste(resp, "~", rhs))

if (col %in% binary_cols) {
  mod  <- suppressWarnings(glm(fml, data = df_sub, family = binomial))
  sm   <- summary(mod)
  est  <- coef(mod)["TropismX4"]
  se   <- sm$coefficients["TropismX4","Std. Error"]
  pval <- sm$coefficients["TropismX4","Pr(>|z|)"]
  estimate <- exp(est)
  ci_lower <- exp(est - 1.96*se)
  ci_upper <- exp(est + 1.96*se)
  effect <- "OR"
} else {
  mod  <- lm(fml, data = df_sub)
  sm   <- summary(mod)
  est  <- coef(mod)["TropismX4"]
  se   <- sm$coefficients["TropismX4","Std. Error"]
  pval <- sm$coefficients["TropismX4","Pr(>|t|)"]
  estimate <- est
  ci_lower <- est - 1.96*se
  ci_upper <- est + 1.96*se
  effect <- "Beta"
}

results_ls[[col]] <- tibble(
  Variable   = col,
  Type       = if_else(col %in% binary_cols, "binary", "continuous"),
  N_R5       = N_R5,
  N_X4       = N_X4,
  Count1_R5  = n1_R5,
  Pct1_R5    = pct1_R5,
  Count1_X4  = n1_X4,
  Pct1_X4    = pct1_X4,
  Estimate   = estimate,
  CI_Lower   = ci_lower,
  CI_Upper   = ci_upper,
  PValue_raw = pval,
  Effect     = effect
)


}

final_df <- bind_rows(results_ls) %>%
mutate(Signif_raw = case_when(
PValue_raw < 0.001 ~ "***",
PValue_raw < 0.01  ~ "**",
PValue_raw < 0.05  ~ "*",
TRUE               ~ ""
))

wb <- createWorkbook()
addWorksheet(wb, "Results")
writeData(wb, "Results", final_df)
saveWorkbook(wb,
paste0("Mixed_Models_", cov_name, "_", filter_suffix, ".xlsx"),
overwrite = TRUE)

p_bin <- final_df %>%
filter(Type == "binary" & PValue_raw < 0.05) %>%
ggplot(aes(x = Estimate, y = reorder(Variable, Estimate))) +
geom_point(size = 3) +
geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2) +
scale_x_log10() +
geom_vline(xintercept = 1, linetype = "dashed") +
labs(title = paste("Binary (OR) —", cov_name),
x = "Odds Ratio (log scale)", y = NULL) +
theme_minimal()

p_cont <- final_df %>%
filter(Type == "continuous" & PValue_raw < 0.05) %>%
ggplot(aes(x = Estimate, y = reorder(Variable, Estimate))) +
geom_point(size = 3) +
geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2) +
geom_vline(xintercept = 0, linetype = "dashed") +
labs(title = paste("Continuous (Beta) —", cov_name),
x = "Beta coefficient", y = NULL) +
theme_minimal()

ggsave(
paste0("ForestPlot_", cov_name, "_SigOnly_", filter_suffix, ".pdf"),
plot_grid(p_bin, p_cont, ncol = 2, rel_widths = c(0.5, 0.5)),
width = 12, height = 6
)
}

```
#ORs_Logistic regression and forest plot of selected clinical outcomes

This chunk runs unadjusted and adjusted logistic regression analyses for a predefined set of binary clinical outcomes, comparing X4 vs R5 tropism. The variables can be decided based on previous chunk or based on relevance for manuscript.

For each outcome:
 1) Fit an unadjusted model: outcome ~ Tropism
 2) Fit an adjusted model:  outcome ~ Tropism + outcome-specific covariates
 3) Extract ORs, 95% CIs, p-values, and event counts per group
 
The results are:
- Saved to an Excel file (one row per model per outcome)
- Summarized into a single forest plot showing raw and adjusted ORs
```{r}
setwd(file.path(base_out, "ORs"))

df <- df_raw %>% prep_r5x4() %>% add_derived_cols()

# ---- Outcomes -> covariates map ----

var_covs <- list(
"CDC3"                    = c("AGE","SEX_BIRTH" ,"TIMETOCART"),
"OPP_AIDS_DEF"            = c("AGE","SEX_BIRTH","TIMETOCART"),
"OTHER_AIDS_DEF"          = c("AGE","SEX_BIRTH","TIMETOCART"),
"INR_Willem2"              = c("AGE","SEX_BIRTH","TIMETOCART"),
"MH_PSY_D.Other"          = c("AGE","SEX_BIRTH","TIMETOCART"),
"MED_PSY.Antipsychotics"  = c("AGE","SEX_BIRTH","TIMETOCART"),
"MED_PSY.Benzodiazepines" = c("AGE","SEX_BIRTH","TIMETOCART"),
"MH_TR_END_D.DM2"         = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","SMOKING_CURRENT"),
"Plaque"                  = c("AGE","SEX_BIRTH","TIMETOCART","HT", "DM","HC","BMI_BASELINE","SMOKING_CURRENT"),
"MALIG_NONAIDS"           = c("AGE","SEX_BIRTH","TIMETOCART","SMOKING_CURRENT","ALCOHOL_UNITS_WEEK"),
"CVD_baseline"            = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","HT","HC","DM","SMOKING_CURRENT"),
"Steatosis_yesno"         = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","HT","HC","DM","SMOKING_CURRENT"),
"hi_vs_rest"              = c("AGE","SEX_BIRTH","TIMETOCART")
)

library(splines)

check_linearity_one <- function(df, y, covs, var = c("AGE","BMI_BASELINE"), df_spline = 3) {
  var <- match.arg(var)

  cols <- unique(c(y, "Tropism", covs))
  d <- df %>%
    select(any_of(cols)) %>%
    mutate("{y}" := coerce_binary01(.data[[y]])) %>%
    filter(!is.na(.data[[y]])) %>%
    na.omit()

  # Need variation + enough rows
  if (nrow(d) < 50) return(NULL)
  if (!all(sort(unique(d[[y]])) %in% c(0,1))) return(NULL)
  if (length(unique(d$Tropism)) < 2) return(NULL)
  if (!var %in% names(d)) return(NULL)
  if (all(is.na(d[[var]]))) return(NULL)
  if (length(unique(d[[var]])) < 5) return(NULL)

  # Linear model
  rhs_lin <- paste(c("Tropism", covs), collapse = " + ")
  f_lin   <- as.formula(sprintf("`%s` ~ %s", y, rhs_lin))
  m_lin   <- suppressWarnings(glm(f_lin, data = d, family = binomial))

  # Spline model: replace var with ns(var, df)
  covs_spl <- covs
  covs_spl[covs_spl == var] <- sprintf("ns(%s, df=%d)", var, df_spline)
  rhs_spl <- paste(c("Tropism", covs_spl), collapse = " + ")
  f_spl   <- as.formula(sprintf("`%s` ~ %s", y, rhs_spl))
  m_spl   <- suppressWarnings(glm(f_spl, data = d, family = binomial))

  # Likelihood ratio test (does spline improve fit?)
  lrt <- anova(m_lin, m_spl, test = "LRT")

  tibble(
    Outcome = y,
    Variable = var,
    LRT_p = lrt$`Pr(>Chi)`[2],
    AIC_linear = AIC(m_lin),
    AIC_spline = AIC(m_spl),
    N = nrow(d)
  )
}
# Check linearity for AGE and BMI_BASELINE in adjusted models (where present)
lin_checks <- purrr::imap_dfr(var_covs, function(covs, y) {
  bind_rows(
    if ("AGE" %in% covs) check_linearity_one(df, y, covs, var = "AGE") else NULL,
    if ("BMI_BASELINE" %in% covs) check_linearity_one(df, y, covs, var = "BMI_BASELINE") else NULL
  )
})

lin_checks <- lin_checks %>% arrange(LRT_p)

# Save for documentation
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
write.xlsx(lin_checks, paste0("Linearity_checks_splines_", timestamp, ".xlsx"))
lin_checks

# ---- Fit function: Unadjusted + Adjusted rows with counts ----
fit_unadj_adj <- function(y, covs) {

# Unadjusted

d_u <- df %>%
select(all_of(c(y, "Tropism"))) %>%
mutate("{y}" := coerce_binary01(.data[[y]])) %>%
filter(!is.na(.data[[y]])) %>%
na.omit()

if (!all(sort(unique(d_u[[y]])) %in% c(0,1))) return(NULL)
if (length(unique(d_u$Tropism)) < 2) return(NULL)

N_R5_u  <- sum(d_u$Tropism == "R5")
N_X4_u  <- sum(d_u$Tropism == "X4")
Ev_R5_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "R5")
Ev_X4_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "X4")

f0 <- as.formula(sprintf("`%s` ~ Tropism", y))
m0 <- suppressWarnings(glm(f0, data = d_u, family = binomial))
co0 <- summary(m0)$coefficients
if (!"TropismX4" %in% rownames(co0)) return(NULL)
b0  <- co0["TropismX4","Estimate"]
se0 <- co0["TropismX4","Std. Error"]
p0  <- co0["TropismX4","Pr(>|z|)"]

# Adjusted

cols_a <- unique(c(y, "Tropism", covs))
d_a <- df %>%
select(any_of(cols_a)) %>%
mutate("{y}" := coerce_binary01(.data[[y]])) %>%
filter(!is.na(.data[[y]])) %>%
na.omit()

if (length(unique(d_a$Tropism)) < 2) return(NULL)

N_R5_a  <- sum(d_a$Tropism == "R5")
N_X4_a  <- sum(d_a$Tropism == "X4")
Ev_R5_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "R5")
Ev_X4_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "X4")

rhs <- paste(c("Tropism", covs), collapse = " + ")
f1  <- as.formula(sprintf("`%s` ~ %s", y, rhs))
m1  <- suppressWarnings(glm(f1, data = d_a, family = binomial))
co1 <- summary(m1)$coefficients
if (!"TropismX4" %in% rownames(co1)) return(NULL)
b1  <- co1["TropismX4","Estimate"]
se1 <- co1["TropismX4","Std. Error"]
p1  <- co1["TropismX4","Pr(>|z|)"]

tibble(
Outcome   = y,
Model     = c("Unadjusted","Adjusted"),
OR        = exp(c(b0, b1)),
CI_Lower  = exp(c(b0 - 1.96*se0, b1 - 1.96*se1)),
CI_Upper  = exp(c(b0 + 1.96*se0, b1 + 1.96*se1)),
PValue    = c(p0, p1),
Covariates= c("none", paste(covs, collapse = ", ")),
N_R5      = c(N_R5_u, N_R5_a),
N_X4      = c(N_X4_u, N_X4_a),
Events_R5 = c(Ev_R5_u, Ev_R5_a),
Events_X4 = c(Ev_X4_u, Ev_X4_a)
)
}

# ---- Run models across outcomes ----

res <- purrr::imap_dfr(var_covs, ~fit_unadj_adj(.y, .x)) %>%
mutate(
Outcome = factor(Outcome, levels = names(var_covs)),
Pct_X4  = ifelse(N_X4 > 0, 100 * Events_X4 / N_X4, NA_real_),
Pct_R5  = ifelse(N_R5 > 0, 100 * Events_R5 / N_R5, NA_real_)
)

# ---- Save results to Excel ----

ts <- format(as.POSIXct(Sys.time(), tz = "Europe/Brussels"), "%Y%m%d_%H%M%S")
xlsx_name <- paste0("Outcomes_Unadj_vs_Adj_CustomCovs_risk_factors", ts, ".xlsx")

wb <- createWorkbook()
addWorksheet(wb, "Results_Unadj_Adj_withCounts")
writeData(wb, "Results_Unadj_Adj_withCounts", res)
saveWorkbook(wb, xlsx_name, overwrite = TRUE)

# ---- Build forest table ----

adj <- res %>%
filter(Model == "Adjusted") %>%
transmute(
Outcome,
adjOR = OR, adj_L = CI_Lower, adj_U = CI_Upper,
adj_p = PValue,
Covariates,
N_R5 = N_R5, N_X4 = N_X4, Ev_R5 = Events_R5, Ev_X4 = Events_X4
)

raw <- res %>%
filter(Model == "Unadjusted") %>%
transmute(
Outcome,
rawOR = OR, raw_L = CI_Lower, raw_U = CI_Upper,
raw_p = PValue
)

forest_df <- left_join(adj, raw, by = "Outcome") %>%
mutate(
Outcome   = factor(Outcome, levels = names(var_covs)),
raw_p_lab = fmt_p(raw_p),
adj_p_lab = fmt_p(adj_p),
Pct_X4    = ifelse(N_X4 > 0, 100 * Ev_X4 / N_X4, NA_real_),
Pct_R5    = ifelse(N_R5 > 0, 100 * Ev_R5 / N_R5, NA_real_)
) %>%
arrange(Outcome)

# ---- Right-side tables (REORDERED): Outcome | R5 | X4 | raw OR | raw p | adj OR | adj p | Covariates ----

# (Optional) wrap covariates if they are long
# forest_df <- forest_df %>% mutate(Covariates = stringr::str_wrap(Covariates, width = 45))

counts_tbl <- forest_df %>%
  transmute(
    Outcome   = as.character(Outcome),
    `R5 n(%)` = ifelse(is.na(Ev_R5) | is.na(Pct_R5), "",
                       paste0(Ev_R5, " (", round(Pct_R5), "%)")),
    `X4 n(%)` = ifelse(is.na(Ev_X4) | is.na(Pct_X4), "",
                       paste0(Ev_X4, " (", round(Pct_X4), "%)"))
  )

tg <- tableGrob(
  counts_tbl, rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

# raw OR + raw p
rg <- tableGrob(
  data.frame(
    `raw OR` = sprintf("%.2f", forest_df$rawOR),
    `raw p`  = forest_df$raw_p_lab,
    check.names = FALSE
  ),
  rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

# adj OR + adj p
og <- tableGrob(
  data.frame(
    `adj OR` = sprintf("%.2f", forest_df$adjOR),
    `adj p`  = forest_df$adj_p_lab,
    check.names = FALSE
  ),
  rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

# Covariates only
pg <- tableGrob(
  forest_df %>% transmute(Covariates),
  rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

# ---- Forest plot ----
fp <- ggplot(forest_df, aes(y = Outcome)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +

  geom_errorbar(aes(xmin = adj_L, xmax = adj_U),
                width = 0.22, linewidth = 0.8, colour = "brown",
                position = position_nudge(y = -0.10),
                orientation = "y") +
  geom_point(aes(x = adjOR),
             size = 3, colour = "brown",
             position = position_nudge(y = -0.10)) +

  geom_errorbar(aes(xmin = raw_L, xmax = raw_U),
                width = 0.22, linewidth = 0.8, colour = "black",
                position = position_nudge(y = 0.10),
                orientation = "y") +
  geom_point(aes(x = rawOR),
             size = 3, colour = "black",
             position = position_nudge(y = 0.10)) +

  scale_x_log10(breaks = c(0.2, 1, 2, 4, 8), limits = c(0.2, 8)) +
  scale_y_discrete(limits = rev(levels(forest_df$Outcome))) +
  labs(x = "OR (95% CI) for Tropism (X4 vs R5)", y = NULL) +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank())

# ---- Assemble and save (ORDERED) ----
combined <- plot_grid(
  tg, rg, og, fp, pg,
  ncol       = 5,
  rel_widths = c(0.06, 0.02, 0.04, 0.12, 0.1),
  align      = "v"
)

print(combined)

pdf_name <- paste0("Forest_Selected_Outcomes_Unadj_vs_Adj_RiskFactors", ts, ".pdf")

ggsave(
  filename = pdf_name,
  plot     = combined,
  width    = 28,
  height   = 10
)
```


#ORs_Logistic regression (with CD4_Nadir sensitivity)
```{r}

# ===================== SETUP =====================
setwd(file.path(base_out, "ORs"))

df <- df_raw %>% prep_r5x4() %>% add_derived_cols()

# ---- Outcomes -> covariates map (CD4_NADIR added to ALL outcomes) ----
var_covs <- list(
  "CDC3"                    = c("AGE","SEX_BIRTH","TIMETOCART"),
  "OPP_AIDS_DEF"            = c("AGE","SEX_BIRTH","TIMETOCART"),
  "OTHER_AIDS_DEF"          = c("AGE","SEX_BIRTH","TIMETOCART"),
  "INR_Willem"              = c("AGE","SEX_BIRTH","CD4_NADIR"),
  "MH_PSY_D.Other"          = c("AGE","SEX_BIRTH","TIMETOCART","CD4_NADIR"),
  "MED_PSY.Antipsychotics"  = c("AGE","SEX_BIRTH","TIMETOCART","CD4_NADIR"),
  "MED_PSY.Benzodiazepines" = c("AGE","SEX_BIRTH","TIMETOCART","CD4_NADIR"),
  "MH_TR_END_D.DM2"         = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","SMOKING_CURRENT","CD4_NADIR"),
  "Plaque"                  = c("AGE","HT","SEX_BIRTH","TIMETOCART","DM","HC","BMI_BASELINE","SMOKING_CURRENT","CD4_NADIR"),
  "MALIG_NONAIDS"           = c("AGE","SEX_BIRTH","TIMETOCART","SMOKING_CURRENT","ALCOHOL_UNITS_WEEK","CD4_NADIR"),
  "CVD_baseline"            = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","HT","HC","DM","SMOKING_CURRENT","CD4_NADIR"),
  "Steatosis_yesno"         = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","HT","HC","DM","SMOKING_CURRENT","CD4_NADIR"),
  "hi_vs_rest"              = c("AGE","SEX_BIRTH","TIMETOCART","CD4_NADIR")
)


# ===================== LINEARITY CHECKS =====================
library(splines)

check_linearity_one <- function(df, y, covs, var = c("AGE","BMI_BASELINE","CD4_NADIR"), df_spline = 3) {
  var <- match.arg(var)

  cols <- unique(c(y, "Tropism", covs))
  d <- df %>%
    select(any_of(cols)) %>%
    mutate("{y}" := coerce_binary01(.data[[y]])) %>%
    filter(!is.na(.data[[y]])) %>%
    na.omit()

  if (nrow(d) < 50) return(NULL)
  if (!all(sort(unique(d[[y]])) %in% c(0, 1))) return(NULL)
  if (length(unique(d$Tropism)) < 2) return(NULL)
  if (!var %in% names(d)) return(NULL)
  if (all(is.na(d[[var]]))) return(NULL)
  if (length(unique(d[[var]])) < 5) return(NULL)

  rhs_lin <- paste(c("Tropism", covs), collapse = " + ")
  f_lin   <- as.formula(sprintf("`%s` ~ %s", y, rhs_lin))
  m_lin   <- suppressWarnings(glm(f_lin, data = d, family = binomial))

  covs_spl <- covs
  covs_spl[covs_spl == var] <- sprintf("ns(%s, df=%d)", var, df_spline)
  rhs_spl <- paste(c("Tropism", covs_spl), collapse = " + ")
  f_spl   <- as.formula(sprintf("`%s` ~ %s", y, rhs_spl))
  m_spl   <- suppressWarnings(glm(f_spl, data = d, family = binomial))

  lrt <- anova(m_lin, m_spl, test = "LRT")

  tibble(
    Outcome    = y,
    Variable   = var,
    LRT_p      = lrt$`Pr(>Chi)`[2],
    AIC_linear = AIC(m_lin),
    AIC_spline = AIC(m_spl),
    N          = nrow(d)
  )
}

lin_checks <- purrr::imap_dfr(var_covs, function(covs, y) {
  bind_rows(
    if ("AGE" %in% covs)          check_linearity_one(df, y, covs, var = "AGE") else NULL,
    if ("BMI_BASELINE" %in% covs) check_linearity_one(df, y, covs, var = "BMI_BASELINE") else NULL,
    if ("CD4_NADIR" %in% covs)    check_linearity_one(df, y, covs, var = "CD4_NADIR") else NULL
  )
})

lin_checks <- lin_checks %>% arrange(LRT_p)

timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
write.xlsx(lin_checks, paste0("Linearity_checks_splines_CD4NADIR_all_", timestamp, ".xlsx"))

lin_checks

# ===================== LOGISTIC MODELS + FOREST PLOT =====================

# ---- Fit function: Unadjusted + Adjusted rows with counts ----
fit_unadj_adj <- function(y, covs) {

  # ---------------- Unadjusted ----------------
  d_u <- df %>%
    select(all_of(c(y, "Tropism"))) %>%
    mutate("{y}" := coerce_binary01(.data[[y]])) %>%
    filter(!is.na(.data[[y]])) %>%
    na.omit()

  if (!all(sort(unique(d_u[[y]])) %in% c(0, 1))) return(NULL)
  if (length(unique(d_u$Tropism)) < 2) return(NULL)

  N_R5_u  <- sum(d_u$Tropism == "R5")
  N_X4_u  <- sum(d_u$Tropism == "X4")
  Ev_R5_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "R5")
  Ev_X4_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "X4")

  f0 <- as.formula(sprintf("`%s` ~ Tropism", y))
  m0 <- suppressWarnings(glm(f0, data = d_u, family = binomial))
  co0 <- summary(m0)$coefficients
  if (!"TropismX4" %in% rownames(co0)) return(NULL)

  b0  <- co0["TropismX4", "Estimate"]
  se0 <- co0["TropismX4", "Std. Error"]
  p0  <- co0["TropismX4", "Pr(>|z|)"]

  # ---------------- Adjusted ----------------
  cols_a <- unique(c(y, "Tropism", covs))
  d_a <- df %>%
    select(any_of(cols_a)) %>%
    mutate("{y}" := coerce_binary01(.data[[y]])) %>%
    filter(!is.na(.data[[y]])) %>%
    na.omit()

  if (!all(sort(unique(d_a[[y]])) %in% c(0, 1))) return(NULL)
  if (length(unique(d_a$Tropism)) < 2) return(NULL)

  N_R5_a  <- sum(d_a$Tropism == "R5")
  N_X4_a  <- sum(d_a$Tropism == "X4")
  Ev_R5_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "R5")
  Ev_X4_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "X4")

  rhs <- paste(c("Tropism", covs), collapse = " + ")
  f1  <- as.formula(sprintf("`%s` ~ %s", y, rhs))
  m1  <- suppressWarnings(glm(f1, data = d_a, family = binomial))
  co1 <- summary(m1)$coefficients
  if (!"TropismX4" %in% rownames(co1)) return(NULL)

  b1  <- co1["TropismX4", "Estimate"]
  se1 <- co1["TropismX4", "Std. Error"]
  p1  <- co1["TropismX4", "Pr(>|z|)"]

  tibble(
    Outcome    = y,
    Model      = c("Unadjusted", "Adjusted"),
    OR         = exp(c(b0, b1)),
    CI_Lower   = exp(c(b0 - 1.96 * se0, b1 - 1.96 * se1)),
    CI_Upper   = exp(c(b0 + 1.96 * se0, b1 + 1.96 * se1)),
    PValue     = c(p0, p1),
    Covariates = c("none", paste(covs, collapse = ", ")),
    N_R5       = c(N_R5_u, N_R5_a),
    N_X4       = c(N_X4_u, N_X4_a),
    Events_R5  = c(Ev_R5_u, Ev_R5_a),
    Events_X4  = c(Ev_X4_u, Ev_X4_a)
  )
}

# ---- Run models across outcomes ----
res <- purrr::imap_dfr(var_covs, ~ fit_unadj_adj(.y, .x)) %>%
  mutate(
    Outcome = factor(Outcome, levels = names(var_covs)),
    Pct_X4  = ifelse(N_X4 > 0, 100 * Events_X4 / N_X4, NA_real_),
    Pct_R5  = ifelse(N_R5 > 0, 100 * Events_R5 / N_R5, NA_real_)
  )

# ---- Save results to Excel ----
ts <- format(as.POSIXct(Sys.time(), tz = "Europe/Brussels"), "%Y%m%d_%H%M%S")
xlsx_name <- paste0("Outcomes_Unadj_vs_Adj_CD4NADIR_all_", ts, ".xlsx")

wb <- createWorkbook()
addWorksheet(wb, "Results_Unadj_Adj_withCounts")
writeData(wb, "Results_Unadj_Adj_withCounts", res)
saveWorkbook(wb, xlsx_name, overwrite = TRUE)

# ---- Build forest table ----
adj <- res %>%
  filter(Model == "Adjusted") %>%
  transmute(
    Outcome,
    adjOR = OR, adj_L = CI_Lower, adj_U = CI_Upper,
    adj_p = PValue,
    Covariates,
    N_R5 = N_R5, N_X4 = N_X4, Ev_R5 = Events_R5, Ev_X4 = Events_X4
  )

raw <- res %>%
  filter(Model == "Unadjusted") %>%
  transmute(
    Outcome,
    rawOR = OR, raw_L = CI_Lower, raw_U = CI_Upper,
    raw_p = PValue
  )

forest_df <- left_join(adj, raw, by = "Outcome") %>%
  mutate(
    Outcome   = factor(Outcome, levels = names(var_covs)),
    raw_p_lab = fmt_p(raw_p),
    adj_p_lab = fmt_p(adj_p),
    Pct_X4    = ifelse(N_X4 > 0, 100 * Ev_X4 / N_X4, NA_real_),
    Pct_R5    = ifelse(N_R5 > 0, 100 * Ev_R5 / N_R5, NA_real_)
  ) %>%
  arrange(Outcome)

# ---- Tables ----
counts_tbl <- forest_df %>%
  transmute(
    Outcome   = as.character(Outcome),
    `R5 n(%)` = ifelse(is.na(Ev_R5) | is.na(Pct_R5), "",
                       paste0(Ev_R5, " (", round(Pct_R5), "%)")),
    `X4 n(%)` = ifelse(is.na(Ev_X4) | is.na(Pct_X4), "",
                       paste0(Ev_X4, " (", round(Pct_X4), "%)"))
  )

tg <- tableGrob(
  counts_tbl, rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

rg <- tableGrob(
  data.frame(
    `raw OR` = sprintf("%.2f", forest_df$rawOR),
    `raw p`  = forest_df$raw_p_lab,
    check.names = FALSE
  ),
  rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

og <- tableGrob(
  data.frame(
    `adj OR` = sprintf("%.2f", forest_df$adjOR),
    `adj p`  = forest_df$adj_p_lab,
    check.names = FALSE
  ),
  rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

pg <- tableGrob(
  forest_df %>% transmute(Covariates),
  rows = NULL,
  theme = ttheme_minimal(
    core    = list(fg_params = list(hjust = 0, x = 0)),
    colhead = list(fg_params = list(hjust = 0, x = 0))
  )
)

# ---- Forest plot ----
fp <- ggplot(forest_df, aes(y = Outcome)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbar(
    aes(xmin = adj_L, xmax = adj_U),
    width = 0.22, linewidth = 0.8, colour = "brown",
    position = position_nudge(y = -0.10),
    orientation = "y"
  ) +
  geom_point(
    aes(x = adjOR),
    size = 3, colour = "brown",
    position = position_nudge(y = -0.10)
  ) +
  geom_errorbar(
    aes(xmin = raw_L, xmax = raw_U),
    width = 0.22, linewidth = 0.8, colour = "black",
    position = position_nudge(y = 0.10),
    orientation = "y"
  ) +
  geom_point(
    aes(x = rawOR),
    size = 3, colour = "black",
    position = position_nudge(y = 0.10)
  ) +
  scale_x_log10(breaks = c(0.2, 1, 2, 4, 8), limits = c(0.2, 8)) +
  scale_y_discrete(limits = rev(levels(forest_df$Outcome))) +
  labs(x = "OR (95% CI) for Tropism (X4 vs R5)", y = NULL) +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor = element_blank())

# ---- Assemble and save ----
combined <- plot_grid(
  tg, rg, og, fp, pg,
  ncol       = 5,
  rel_widths = c(0.06, 0.02, 0.04, 0.12, 0.10),
  align      = "v"
)

print(combined)

pdf_name <- paste0("Forest_Selected_Outcomes_Unadj_vs_Adj_CD4NADIR_all_", ts, ".pdf")

ggsave(
  filename = pdf_name,
  plot     = combined,
  width    = 28,
  height   = 10
)
```
#interaction check CD4_nadir
```{r}
# ============================================================
# FULL SCRIPT: Tropism * CD4_NADIR interaction checks
# + Excel output with:
#   (1) Interaction test results
#   (2) Explicit main-effects + interaction model formulas per outcome
# ============================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(ggplot2)
  library(openxlsx)
})

# -----------------------------
# Helpers (assumed to exist in your pipeline)
# -----------------------------
# df_raw
# prep_r5x4()
# add_derived_cols()
# coerce_binary01()
# var_covs (named list: outcome -> vector of covariates)

# ============================================================
# Interaction check for ONE outcome: Tropism * interact_var
# Uses LRT (nested model) + Wald p-value for interaction term(s)
# ============================================================

interaction_check_one <- function(df, outcome, covs, interact_var,
                                  family = binomial, min_n = 50) {

  # Only test if interact_var is in covariates for that outcome
  if (!interact_var %in% covs) return(NULL)
  if (!outcome %in% names(df)) return(NULL)
  if (!all(c("Tropism", interact_var) %in% names(df))) return(NULL)

  cols_use <- unique(c(outcome, "Tropism", covs))

  d <- df %>%
    dplyr::select(dplyr::any_of(cols_use)) %>%
    dplyr::mutate("{outcome}" := coerce_binary01(.data[[outcome]])) %>%
    dplyr::filter(!is.na(.data[[outcome]])) %>%
    tidyr::drop_na()

  # basic checks
  if (nrow(d) < min_n) return(NULL)
  if (length(unique(d$Tropism)) < 2) return(NULL)
  if (!all(sort(unique(d[[outcome]])) %in% c(0, 1))) return(NULL)
  if (length(unique(d[[interact_var]])) < 2) return(NULL)

  # If interact_var is categorical, require >=2 levels after filtering
  if (is.character(d[[interact_var]]) || is.factor(d[[interact_var]])) {
    if (length(unique(d[[interact_var]])) < 2) return(NULL)
  }

  # -------------------------
  # Main-effects model: Outcome ~ Tropism + covariates
  # -------------------------
  rhs_main <- paste(c("Tropism", covs), collapse = " + ")
  f_main   <- as.formula(sprintf("`%s` ~ %s", outcome, rhs_main))
  m_main   <- suppressWarnings(glm(f_main, data = d, family = family))

  # -------------------------
  # Interaction model: replace interact_var with Tropism*interact_var
  # Outcome ~ Tropism + covariates + Tropism*interact_var
  # -------------------------
  covs_int <- covs
  covs_int[covs_int == interact_var] <- paste0("Tropism*", interact_var)

  rhs_int <- paste(c("Tropism", covs_int[covs_int != "Tropism"]), collapse = " + ")
  f_int   <- as.formula(sprintf("`%s` ~ %s", outcome, rhs_int))
  m_int   <- suppressWarnings(glm(f_int, data = d, family = family))

  # -------------------------
  # Likelihood ratio test: does interaction improve fit?
  # -------------------------
  lrt <- anova(m_main, m_int, test = "LRT")
  lrt_p <- lrt$`Pr(>Chi)`[2]

  # -------------------------
  # Wald p-value(s) for interaction term(s)
  # (handles common naming: TropismX4:CD4_NADIR or CD4_NADIR:TropismX4)
  # -------------------------
  co <- summary(m_int)$coefficients
  int_rows <- grep(
    paste0("^TropismX4:", interact_var, "|^", interact_var, ":TropismX4$"),
    rownames(co),
    value = TRUE
  )
  wald_p <- if (length(int_rows) == 0) NA_real_ else max(co[int_rows, "Pr(>|z|)"], na.rm = TRUE)

  tibble::tibble(
    Outcome        = outcome,
    InteractionVar = interact_var,
    N              = nrow(d),
    LRT_p          = lrt_p,
    Wald_p_max     = wald_p,
    AIC_main       = AIC(m_main),
    AIC_int        = AIC(m_int)
  )
}

# ============================================================
# Build a table of model formulas for each outcome
# ============================================================

make_model_formula_table <- function(var_covs, interact_var = "CD4_NADIR") {

  purrr::imap_dfr(var_covs, function(covs, outcome) {

    if (!interact_var %in% covs) return(NULL)

    covs <- unique(covs)

    # Main model formula text
    rhs_main <- paste(c("Tropism", covs), collapse = " + ")
    f_main_txt <- sprintf("%s ~ %s", outcome, rhs_main)

    # Interaction model formula text
    covs_int <- covs
    covs_int[covs_int == interact_var] <- paste0("Tropism*", interact_var)
    rhs_int <- paste(c("Tropism", covs_int[covs_int != "Tropism"]), collapse = " + ")
    f_int_txt <- sprintf("%s ~ %s", outcome, rhs_int)

    # Expected interaction coefficient name (useful for reviewer clarity)
    expected_term <- paste0("TropismX4:", interact_var)

    tibble::tibble(
      Outcome = outcome,
      InteractionVar = interact_var,
      MainEffectsModel = f_main_txt,
      InteractionModel = f_int_txt,
      ExpectedInteractionTerm = expected_term
    )
  })
}

# ============================================================
# Plot predicted probabilities for ONE outcome (if needed)
# ============================================================

plot_interaction_pred <- function(df, outcome, covs, interact_var = "CD4_NADIR") {

  cols_use <- unique(c(outcome, "Tropism", covs))
  d <- df %>%
    dplyr::select(dplyr::any_of(cols_use)) %>%
    dplyr::mutate("{outcome}" := coerce_binary01(.data[[outcome]])) %>%
    dplyr::filter(!is.na(.data[[outcome]])) %>%
    tidyr::drop_na()

  # Fit interaction model
  covs_int <- covs
  covs_int[covs_int == interact_var] <- paste0("Tropism*", interact_var)
  rhs_int <- paste(c("Tropism", covs_int[covs_int != "Tropism"]), collapse = " + ")
  f_int   <- as.formula(sprintf("`%s` ~ %s", outcome, rhs_int))
  m_int   <- suppressWarnings(glm(f_int, data = d, family = binomial))

  # Newdata grid: vary interact_var; hold other covariates at typical values
  grid <- expand.grid(
    Tropism = factor(c("R5", "X4"), levels = c("R5", "X4")),
    x       = seq(min(d[[interact_var]], na.rm = TRUE),
                  max(d[[interact_var]], na.rm = TRUE),
                  length.out = 100)
  )
  names(grid)[names(grid) == "x"] <- interact_var

  # Set other covariates to medians / most frequent levels
  other_covs <- setdiff(covs, interact_var)
  for (v in other_covs) {
    if (!v %in% names(d)) next
    if (is.numeric(d[[v]])) {
      grid[[v]] <- stats::median(d[[v]], na.rm = TRUE)
    } else {
      tab <- sort(table(d[[v]]), decreasing = TRUE)
      grid[[v]] <- names(tab)[1]
      if (is.factor(d[[v]])) {
        grid[[v]] <- factor(grid[[v]], levels = levels(d[[v]]))
      }
    }
  }

  grid$pred <- predict(m_int, newdata = grid, type = "response")

  ggplot2::ggplot(grid, ggplot2::aes(x = .data[[interact_var]], y = pred, color = Tropism)) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::theme_minimal() +
    ggplot2::labs(
      title = paste0("Predicted probability: ", outcome),
      x = interact_var,
      y = "Predicted probability"
    )
}

# ============================================================
# RUN
# ============================================================

# Prepare analysis dataframe
df <- df_raw %>% prep_r5x4() %>% add_derived_cols()

# Choose interaction variable
interact_var <- "CD4_NADIR"

# 1) Interaction tests across all outcomes
interaction_results <- purrr::imap_dfr(var_covs, function(covs, outcome) {
  interaction_check_one(df, outcome, covs, interact_var = interact_var)
}) %>%
  dplyr::arrange(LRT_p)

# 2) Table of explicit model formulas per outcome
model_formula_table <- make_model_formula_table(var_covs, interact_var = interact_var)

# Preview in R
print(interaction_results)
print(model_formula_table)

# ============================================================
# EXPORT TO EXCEL: two sheets in one workbook
# ============================================================

out_xlsx <- paste0(
  "InteractionChecks_TropismX_", interact_var, "_",
  format(Sys.time(), "%Y%m%d_%H%M%S"),
  ".xlsx"
)

wb <- openxlsx::createWorkbook()

openxlsx::addWorksheet(wb, "InteractionTests")
openxlsx::writeData(wb, "InteractionTests", interaction_results)

openxlsx::addWorksheet(wb, "ModelFormulas")
openxlsx::writeData(wb, "ModelFormulas", model_formula_table)

openxlsx::saveWorkbook(wb, out_xlsx, overwrite = TRUE)

message("Saved: ", out_xlsx)
```


#post2014 and infection date known subset_supplementary Figure 2(A)
```{r}
setwd(file.path(base_out, "ORs"))

# ---- Base data (from setup) ----
df_base <- df_raw %>% prep_r5x4() %>% add_derived_cols()

# ---- Subset makers ----
make_post2014 <- function(df) {
  if (!"HIV_DATE" %in% names(df)) stop("Column 'HIV_DATE' not found for POST2014.")
  if (inherits(df$HIV_DATE, "Date")) {
    df %>%
      mutate(HIV_YEAR = as.integer(format(HIV_DATE, "%Y"))) %>%
      filter(!is.na(HIV_YEAR) & HIV_YEAR > 2014)
  } else {
    df %>%
      mutate(HIV_YEAR = suppressWarnings(as.integer(HIV_DATE))) %>%
      filter(!is.na(HIV_YEAR) & HIV_YEAR > 2014)
  }
}

make_inf1y <- function(df) {
  if (!"INF_date_known_1y" %in% names(df)) stop("'INF_date_known_1y' not found.")
  df %>% filter(INF_date_known_1y == 1)
}

# ---- Outcomes & covariates per subset (DIFFERENT) ----
var_covs_by_subset <- list(
  POST2014 = list(
    "INR_Willem"      = c("AGE","SEX_BIRTH","TIMETOCART"),
    "MH_PSY_D.Other"  = c("AGE","SEX_BIRTH","TIMETOCART"),
    "hi_vs_rest"      = c("AGE","SEX_BIRTH","TIMETOCART")
  ),
  INF_1y = list(
    "CDC3"            = c("AGE","SEX_BIRTH","TIMETOCART"),
    "hi_vs_rest"      = c("AGE","SEX_BIRTH","TIMETOCART")
  )
)

# ---- Fit on a subset with its own outcomes; return results + forest_df (no plot yet) ----
fit_on_subset <- function(df_sub, subset_label, var_covs) {

  usable_vars <- names(var_covs)[names(var_covs) %in% names(df_sub)]
  if (length(usable_vars) == 0) return(list(results = tibble(), forest_df = NULL))

  fit_unadj_adj <- function(y, covs) {

    # Unadjusted
    d_u <- df_sub %>%
      select(all_of(c(y, "Tropism"))) %>%
      mutate("{y}" := coerce_binary01(.data[[y]])) %>%
      filter(!is.na(.data[[y]])) %>%
      na.omit()

    if (nrow(d_u) == 0 || !all(sort(unique(d_u[[y]])) %in% c(0,1))) return(NULL)
    if (length(unique(d_u$Tropism)) < 2) return(NULL)

    N_R5_u  <- sum(d_u$Tropism == "R5")
    N_X4_u  <- sum(d_u$Tropism == "X4")
    Ev_R5_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "R5")
    Ev_X4_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "X4")

    m0 <- suppressWarnings(glm(as.formula(sprintf("`%s` ~ Tropism", y)),
                               data = d_u, family = binomial))
    co0 <- summary(m0)$coefficients
    if (!"TropismX4" %in% rownames(co0)) return(NULL)
    b0  <- co0["TropismX4","Estimate"]
    se0 <- co0["TropismX4","Std. Error"]
    p0  <- co0["TropismX4","Pr(>|z|)"]

    # Adjusted
    base_covs <- covs %||% character(0)
    covs_use  <- setdiff(intersect(unique(base_covs), names(df_sub)), "Tropism")

    d_a <- df_sub %>%
      select(any_of(c(y, "Tropism", covs_use))) %>%
      mutate("{y}" := coerce_binary01(.data[[y]])) %>%
      filter(!is.na(.data[[y]])) %>%
      na.omit()

    if (nrow(d_a) == 0 || length(unique(d_a$Tropism)) < 2) return(NULL)

    N_R5_a  <- sum(d_a$Tropism == "R5")
    N_X4_a  <- sum(d_a$Tropism == "X4")
    Ev_R5_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "R5")
    Ev_X4_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "X4")

    rhs <- if (length(covs_use) > 0) {
      paste("Tropism", paste(covs_use, collapse = " + "), sep = " + ")
    } else "Tropism"

    m1 <- suppressWarnings(glm(as.formula(sprintf("`%s` ~ %s", y, rhs)),
                               data = d_a, family = binomial))
    co1 <- summary(m1)$coefficients
    if (!"TropismX4" %in% rownames(co1)) return(NULL)
    b1  <- co1["TropismX4","Estimate"]
    se1 <- co1["TropismX4","Std. Error"]
    p1  <- co1["TropismX4","Pr(>|z|)"]

    tibble(
      Subset    = subset_label,
      Outcome   = y,
      Model     = c("Unadjusted","Adjusted"),
      OR        = exp(c(b0, b1)),
      CI_Lower  = exp(c(b0 - 1.96*se0, b1 - 1.96*se1)),
      CI_Upper  = exp(c(b0 + 1.96*se0, b1 + 1.96*se1)),
      PValue    = c(p0, p1),
      Covariates= c("none", paste(covs_use, collapse = ", ")),
      N_R5      = c(N_R5_u, N_R5_a),
      N_X4      = c(N_X4_u, N_X4_a),
      Events_R5 = c(Ev_R5_u, Ev_R5_a),
      Events_X4 = c(Ev_X4_u, Ev_X4_a)
    )
  }

  res <- purrr::imap_dfr(var_covs[usable_vars], ~fit_unadj_adj(.y, .x))
  if (nrow(res) == 0) return(list(results = tibble(), forest_df = NULL))

  outcome_levels <- names(var_covs)[names(var_covs) %in% usable_vars]

  adj <- res %>%
    filter(Model == "Adjusted") %>%
    transmute(
      Outcome,
      adjOR = OR, adj_L = CI_Lower, adj_U = CI_Upper,
      adj_p = PValue,
      Covariates, N_R5, N_X4, Ev_R5 = Events_R5, Ev_X4 = Events_X4
    )

  raw <- res %>%
    filter(Model == "Unadjusted") %>%
    transmute(Outcome, rawOR = OR, raw_L = CI_Lower, raw_U = CI_Upper, raw_p = PValue)

  forest_df <- left_join(adj, raw, by = "Outcome") %>%
    mutate(
      Outcome   = factor(Outcome, levels = outcome_levels),
      raw_p_lab = fmt_p(raw_p),
      adj_p_lab = fmt_p(adj_p),
      Pct_X4    = ifelse(N_X4 > 0, 100 * Ev_X4 / N_X4, NA_real_),
      Pct_R5    = ifelse(N_R5 > 0, 100 * Ev_R5 / N_R5, NA_real_)
    ) %>%
    arrange(Outcome)

  list(results = res, forest_df = forest_df)
}

# ---- Build subsets ----
df_post2014 <- make_post2014(df_base)
df_inf1y    <- make_inf1y(df_base)

out_post2014 <- fit_on_subset(df_post2014, "POST2014", var_covs_by_subset$POST2014)
out_inf1y    <- fit_on_subset(df_inf1y,    "INF_1y",   var_covs_by_subset$INF_1y)

# ---- Compute COMMON auto x-range across both panels ----
get_limits_from_forest <- function(fdf) {
  if (is.null(fdf) || nrow(fdf) == 0) return(c(NA_real_, NA_real_))
  lo <- suppressWarnings(min(c(fdf$adj_L, fdf$raw_L), na.rm = TRUE))
  hi <- suppressWarnings(max(c(fdf$adj_U, fdf$raw_U), na.rm = TRUE))
  c(lo, hi)
}

lim1 <- get_limits_from_forest(out_post2014$forest_df)
lim2 <- get_limits_from_forest(out_inf1y$forest_df)
lims_all <- range(c(lim1, lim2), na.rm = TRUE)

if (is.finite(lims_all[1]) && is.finite(lims_all[2]) && lims_all[1] > 0) {
  pad <- 0.06
  log_lo <- log(lims_all[1]); log_hi <- log(lims_all[2])
  span <- log_hi - log_lo
  log_lo <- log_lo - pad * span
  log_hi <- log_hi + pad * span
  x_limits <- c(exp(log_lo), exp(log_hi))
  if (x_limits[1] >= 1) x_limits[1] <- 1/exp(0.5)
  if (x_limits[2] <= 1) x_limits[2] <- exp(0.5)
} else {
  x_limits <- c(0.2, 8)
}

x_breaks <- scales::log_breaks(n = 5)(x_limits)

# ---- Panel builder using SHARED limits/breaks ----
build_forest_panel <- function(forest_df, subset_label, x_limits, x_breaks) {
  if (is.null(forest_df) || nrow(forest_df) == 0) return(NULL)

  counts_tbl <- forest_df %>%
    transmute(
      Outcome = as.character(Outcome),
      `X4 n(%)` = ifelse(is.na(Ev_X4) | is.na(Pct_X4), "",
                        paste0(Ev_X4, " (", round(Pct_X4), "%)")),
      `R5 n(%)` = ifelse(is.na(Ev_R5) | is.na(Pct_R5), "",
                        paste0(Ev_R5, " (", round(Pct_R5), "%)"))
    )

  tg <- tableGrob(counts_tbl, rows = NULL,
                  theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                         colhead=list(fg_params=list(hjust=0, x=0))))
  og <- tableGrob(data.frame(`adj OR` = sprintf("%.2f", forest_df$adjOR), check.names = FALSE),
                  rows = NULL,
                  theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                         colhead=list(fg_params=list(hjust=0, x=0))))
  rg <- tableGrob(data.frame(`raw OR` = sprintf("%.2f", forest_df$rawOR),
                             `raw p`  = forest_df$raw_p_lab, check.names = FALSE),
                  rows = NULL,
                  theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                         colhead=list(fg_params=list(hjust=0, x=0))))
  pg <- tableGrob(forest_df %>% transmute(`adj p` = adj_p_lab, Covariates),
                  rows = NULL,
                  theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                         colhead=list(fg_params=list(hjust=0, x=0))))

  fp <- ggplot(forest_df, aes(y = Outcome)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
    geom_errorbar(aes(xmin = adj_L, xmax = adj_U),
                  width = 0.22, linewidth = 0.8, colour = "brown",
                  position = position_nudge(y = -0.15),
                  orientation = "y") +
    geom_point(aes(x = adjOR), size = 3, colour = "brown",
               position = position_nudge(y = -0.15)) +
    geom_errorbar(aes(xmin = raw_L, xmax = raw_U),
                  width = 0.22, linewidth = 0.8, colour = "black",
                  position = position_nudge(y = 0.15),
                  orientation = "y") +
    geom_point(aes(x = rawOR), size = 3, colour = "black",
               position = position_nudge(y = 0.15)) +
    scale_x_log10(breaks = x_breaks, limits = x_limits) +
    scale_y_discrete(limits = rev(levels(forest_df$Outcome))) +
    labs(x = paste0("OR (95% CI) X4 vs R5 — ", subset_label), y = NULL) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  plot_grid(tg, og, rg, fp, pg,
            ncol = 5, rel_widths = c(0.28, 0.10, 0.12, 0.35, 0.15),
            align = "v")
}

panel_post2014 <- build_forest_panel(out_post2014$forest_df, "POST2014", x_limits, x_breaks)
panel_inf1y    <- build_forest_panel(out_inf1y$forest_df,    "INF_1y",   x_limits, x_breaks)

# ---- ONE Excel (per-subset sheets + Combined) ----
wb <- createWorkbook()
if (!is.null(out_post2014$results) && nrow(out_post2014$results) > 0) {
  addWorksheet(wb, "POST2014"); writeData(wb, "POST2014", out_post2014$results)
}
if (!is.null(out_inf1y$results) && nrow(out_inf1y$results) > 0) {
  addWorksheet(wb, "INF_1y");   writeData(wb, "INF_1y",   out_inf1y$results)
}
combined_results <- bind_rows(out_post2014$results, out_inf1y$results)
if (nrow(combined_results) > 0) {
  addWorksheet(wb, "Combined"); writeData(wb, "Combined", combined_results)
}
saveWorkbook(wb, "Outcomes_post2014_infectionDateKnown2.xlsx", overwrite = TRUE)

# ---- ONE PDF (both panels stacked; shared x-axis) ----
plots_to_stack <- list(panel_post2014, panel_inf1y)
plots_to_stack <- plots_to_stack[!vapply(plots_to_stack, is.null, logical(1))]

both <- plot_grid(panel_post2014, panel_inf1y, ncol = 1, rel_heights = c(1, 1))
ggsave("Forest_post2014_infectionDateKnown2.pdf", both, width = 30, height = 8)
```


# 2-years outcome (follow-up timepoint)
```{r}
# ---- ORs_finalplot: 2y outcomes (uses setup chunk objects) ----
setwd(file.path(base_out, "ORs"))

# Data from setup
df <- df_raw %>% prep_r5x4() %>% add_derived_cols()

# Derived columns specific to this chunk
df <- df %>%
  mutate(
    CDC3_INF1y = if_else(HIV_STAGE_CDC == 2 & INF_date_known_1y == 1,
                         1L, 0L, missing = NA_integer_)
  )

# Make MALIG_2Y binary: 1 = Malignancy, 0 = No malignancy
df <- df %>%
  mutate(
    `MALIG_2Y#No malignancy past 2 years` = case_when(
      is.na(`MALIG_2Y#No malignancy past 2 years`) ~ NA_integer_,
      str_trim(`MALIG_2Y#No malignancy past 2 years`) %in% c("Malignancy","malignancy") ~ 1L,
      str_trim(`MALIG_2Y#No malignancy past 2 years`) %in% c("No malignancy","no malignancy") ~ 0L,
      TRUE ~ NA_integer_
    )
  )

# Outcomes -> covariates map
var_covs <- list(
  "MH_TR_END_D_2Y#Hypercholesterolemia" = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","SMOKING_CURRENT"),
  "MH_TR_END_D_2Y#DM2"                  = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","SMOKING_CURRENT"),
  "MH_TR_GE_D_2Y#Liversteatosis"        = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","SMOKING_CURRENT","HT","DM","HC"),
  "CVD2y"                               = c("AGE","SEX_BIRTH","TIMETOCART","BMI_BASELINE","SMOKING_CURRENT","HT","DM","HC"),
  "MALIG_2Y#No malignancy past 2 years" = c("AGE","SEX_BIRTH","TIMETOCART")
)

# Keep only outcomes that exist
var_covs <- var_covs[names(var_covs) %in% names(df)]
if (length(var_covs) == 0L) stop("None of the requested outcomes are present as columns in df.")

# Fit function (kept here; no helper re-definitions)
fit_unadj_adj <- function(y, covs) {

  # Unadjusted
  d_u <- df %>%
    select(all_of(c(y, "Tropism"))) %>%
    mutate("{y}" := coerce_binary01(.data[[y]])) %>%
    filter(!is.na(.data[[y]])) %>%
    na.omit()

  if (!all(sort(unique(d_u[[y]])) %in% c(0,1))) return(NULL)
  if (length(unique(d_u$Tropism)) < 2) return(NULL)

  N_R5_u  <- sum(d_u$Tropism == "R5")
  N_X4_u  <- sum(d_u$Tropism == "X4")
  Ev_R5_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "R5")
  Ev_X4_u <- sum(d_u[[y]] == 1 & d_u$Tropism == "X4")

  f0 <- as.formula(sprintf("`%s` ~ Tropism", y))
  m0 <- suppressWarnings(glm(f0, data = d_u, family = binomial))
  co0 <- summary(m0)$coefficients
  if (!"TropismX4" %in% rownames(co0)) return(NULL)
  b0  <- co0["TropismX4","Estimate"]
  se0 <- co0["TropismX4","Std. Error"]
  p0  <- co0["TropismX4","Pr(>|z|)"]

  # Adjusted
  cols_a <- c(y, "Tropism", covs)
  d_a <- df %>%
    select(any_of(cols_a)) %>%
    mutate("{y}" := coerce_binary01(.data[[y]])) %>%
    filter(!is.na(.data[[y]])) %>%
    na.omit()

  if (length(unique(d_a$Tropism)) < 2) return(NULL)

  N_R5_a  <- sum(d_a$Tropism == "R5")
  N_X4_a  <- sum(d_a$Tropism == "X4")
  Ev_R5_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "R5")
  Ev_X4_a <- sum(d_a[[y]] == 1 & d_a$Tropism == "X4")

  rhs <- paste(c("Tropism", covs), collapse = " + ")
  f1  <- as.formula(sprintf("`%s` ~ %s", y, rhs))
  m1  <- suppressWarnings(glm(f1, data = d_a, family = binomial))
  co1 <- summary(m1)$coefficients
  if (!"TropismX4" %in% rownames(co1)) return(NULL)
  b1  <- co1["TropismX4","Estimate"]
  se1 <- co1["TropismX4","Std. Error"]
  p1  <- co1["TropismX4","Pr(>|z|)"]

  tibble(
    Outcome    = y,
    Model      = c("Unadjusted","Adjusted"),
    OR         = exp(c(b0, b1)),
    CI_Lower   = exp(c(b0 - 1.96*se0, b1 - 1.96*se1)),
    CI_Upper   = exp(c(b0 + 1.96*se0, b1 + 1.96*se1)),
    PValue     = c(p0, p1),
    Covariates = c("none", paste(covs, collapse = ", ")),
    N_R5       = c(N_R5_u, N_R5_a),
    N_X4       = c(N_X4_u, N_X4_a),
    Events_R5  = c(Ev_R5_u, Ev_R5_a),
    Events_X4  = c(Ev_X4_u, Ev_X4_a)
  )
}

# Run models
res <- purrr::imap_dfr(var_covs, ~fit_unadj_adj(.y, .x)) %>%
  mutate(Outcome = factor(Outcome, levels = names(var_covs))) %>%
  mutate(
    Pct_X4 = ifelse(N_X4 > 0, 100 * Events_X4 / N_X4, NA_real_),
    Pct_R5 = ifelse(N_R5 > 0, 100 * Events_R5 / N_R5, NA_real_)
  )

# Save results
wb <- createWorkbook()
addWorksheet(wb, "Results_Unadj_Adj_withCounts")
writeData(wb, "Results_Unadj_Adj_withCounts", res)
saveWorkbook(wb, "Outcomes_Unadj_vs_Adj_CustomCovs_2y.xlsx", overwrite = TRUE)

# Build forest_df
adj <- res %>%
  filter(Model == "Adjusted") %>%
  transmute(
    Outcome,
    adjOR = OR, adj_L = CI_Lower, adj_U = CI_Upper,
    adj_p = PValue,
    Covariates,
    N_R5 = N_R5, N_X4 = N_X4, Ev_R5 = Events_R5, Ev_X4 = Events_X4
  )

raw <- res %>%
  filter(Model == "Unadjusted") %>%
  transmute(
    Outcome,
    rawOR = OR, raw_L = CI_Lower, raw_U = CI_Upper,
    raw_p = PValue
  )

forest_df <- left_join(adj, raw, by = "Outcome") %>%
  mutate(
    Outcome   = factor(Outcome, levels = names(var_covs)),
    raw_p_lab = fmt_p(raw_p),
    adj_p_lab = fmt_p(adj_p),
    Pct_X4    = ifelse(N_X4 > 0, 100 * Ev_X4 / N_X4, NA_real_),
    Pct_R5    = ifelse(N_R5 > 0, 100 * Ev_R5 / N_R5, NA_real_)
  ) %>%
  arrange(Outcome)

# Right-side tables
counts_tbl <- forest_df %>%
  transmute(
    Outcome = as.character(Outcome),
    `X4 n(%)` = ifelse(is.na(Ev_X4) | is.na(Pct_X4), "",
                      paste0(Ev_X4, " (", round(Pct_X4), "%)")),
    `R5 n(%)` = ifelse(is.na(Ev_R5) | is.na(Pct_R5), "",
                      paste0(Ev_R5, " (", round(Pct_R5), "%)"))
  )

tg <- tableGrob(counts_tbl, rows = NULL,
                theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                       colhead=list(fg_params=list(hjust=0, x=0))))

og <- tableGrob(data.frame(`adj OR` = sprintf("%.2f", forest_df$adjOR), check.names = FALSE),
                rows = NULL,
                theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                       colhead=list(fg_params=list(hjust=0, x=0))))

rg <- tableGrob(data.frame(`raw OR` = sprintf("%.2f", forest_df$rawOR),
                           `raw p`  = forest_df$raw_p_lab,
                           check.names = FALSE),
                rows = NULL,
                theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                       colhead=list(fg_params=list(hjust=0, x=0))))

pg <- tableGrob(forest_df %>% transmute(`adj p` = adj_p_lab, Covariates),
                rows = NULL,
                theme = ttheme_minimal(core=list(fg_params=list(hjust=0, x=0)),
                                       colhead=list(fg_params=list(hjust=0, x=0))))

# X-axis clamp + breaks
lo <- 0.04
hi <- 9
cand <- c(0.04, 0.2, 0.5, 1, 3, 6, 9)
br <- cand[cand >= lo & cand <= hi]

# Forest plot (no geom_errorbarh warning)
fp <- ggplot(forest_df, aes(y = Outcome)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(xmin = adj_L, xmax = adj_U),
                width = 0.22, linewidth = 0.8, colour = "brown",
                position = position_nudge(y = -0.15),
                orientation = "y") +
  geom_point(aes(x = adjOR),
             size = 3, colour = "brown",
             position = position_nudge(y = -0.15)) +
  geom_errorbar(aes(xmin = raw_L, xmax = raw_U),
                width = 0.22, linewidth = 0.8, colour = "black",
                position = position_nudge(y = 0.15),
                orientation = "y") +
  geom_point(aes(x = rawOR),
             size = 3, colour = "black",
             position = position_nudge(y = 0.15)) +
  scale_x_log10(breaks = br, limits = c(lo, hi), minor_breaks = NULL) +
  scale_y_discrete(limits = rev(levels(forest_df$Outcome))) +
  labs(x = "OR (95% CI) for Tropism (X4 vs R5)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

# Assemble and save
combined <- cowplot::plot_grid(
  tg, og, rg, fp, pg,
  ncol       = 5,
  rel_widths = c(0.28, 0.10, 0.12, 0.35, 0.15),
  align      = "v"
)

print(combined)

ggsave(
  filename = "Forest_Selected_Outcomes_Unadj_vs_Adj_2y.pdf",
  plot     = combined,
  width    = 30,
  height   = 8
)
```


#Prevalance of variable 
```{r}
# ---- Outcome counts/percentages by Tropism (R5/X4/Mixed) + grouped bar plot ----
# Uses df_raw from setup. Produces:
# 1) Outcome_Counts_Percentages_ByTropism.xlsx
# 2) GroupedBar_Outcomes_by_Tropism_Percent.pdf (or Events if you switch y_var)

setwd(file.path(base_out, "ORs"))

# ---- Filter + factor Tropism (INCLUDE Mixed) ----
df <- df_raw %>%
  filter(Tropism %in% c("R5","X4","Mixed"),
         CONTROLLER_1B != "EC_persistent") %>%
  mutate(Tropism = factor(Tropism, levels = c("R5","X4","Mixed")))

# ---- Derived columns (keep same logic; if your setup already does this, you can delete this block) ----
df <- df %>%
  mutate(
    lo_vs_rest = if_else(ViReTgroup == "LoViReT", 1, 0, missing = NA_real_),
    hi_vs_rest = if_else(ViReTgroup == "HiViReT", 1, 0, missing = NA_real_),
    hi_vs_lo   = case_when(
      ViReTgroup == "HiViReT" ~ 1,
      ViReTgroup == "LoViReT" ~ 0,
      TRUE ~ NA_real_
    ),
    CDC3 = if_else(HIV_STAGE_CDC == 2, 1L, 0L, missing = NA_integer_)
  ) %>%
  mutate(
    Steatosis_yesno = case_when(
      is.na(Steatosis_yesno) ~ NA_character_,
      trimws(Steatosis_yesno) %in% c("S1","S2","S3","S1-S3") ~ "S1-S3",
      trimws(Steatosis_yesno) == "S0" ~ "S0",
      TRUE ~ NA_character_
    ),
    Steatosis_yesno = factor(Steatosis_yesno, levels = c("S0","S1-S3"))
  )

# ---- Outcomes list ----
outcomes <- c(
  "CDC3",
  "OPP_AIDS_DEF",
  "OTHER_AIDS_DEF",
  "INR_Willem",
  "Plaque",
  "MH_TR_END_D.DM2",
  "MH_PSY_D.Other",
  "MED_PSY.Antipsychotics",
  "MED_PSY.Benzodiazepines",
  "MALIG_NONAIDS",
  "CVD_baseline",
  "Steatosis_yesno",
  "hi_vs_rest"
)

# ---- Event rules ----
event_level_map <- list(
  "Steatosis_yesno" = "S1-S3"
)

# ---- Helper: counts/percent per Tropism + Overall ----
summarize_outcome <- function(data, var) {
  if (!var %in% names(data)) return(NULL)

  v <- data[[var]]

  # Determine event vector
  if (is.factor(v) || is.character(v)) {
    if (var %in% names(event_level_map)) {
      event <- v == event_level_map[[var]]
    } else {
      event <- !is.na(v) & !(v %in% c("0","No","None","False","F","S0"))
    }
    denom <- !is.na(v)
  } else {
    event <- !is.na(v) & (v == 1)
    denom <- !is.na(v)
  }

  tmp <- data %>%
    mutate(.event = event, .denom = denom) %>%
    group_by(Tropism, .drop = FALSE) %>%
    summarize(
      N_non_missing = sum(.denom),
      Events = sum(.event, na.rm = TRUE),
      Percent = ifelse(N_non_missing > 0, round(100 * Events / N_non_missing, 1), NA_real_),
      .groups = "drop"
    )

  overall <- tmp %>%
    summarize(
      Tropism = factor("Overall", levels = c("Overall","R5","X4","Mixed")),
      N_non_missing = sum(N_non_missing),
      Events = sum(Events),
      Percent = ifelse(N_non_missing > 0, round(100 * Events / N_non_missing, 1), NA_real_)
    )

  bind_rows(
    overall,
    tmp %>% mutate(Tropism = factor(as.character(Tropism), levels = c("R5","X4","Mixed")))
  ) %>%
    mutate(Outcome = var, .before = 1) %>%
    select(Outcome, Tropism, N_non_missing, Events, Percent)
}

# ---- Run for all outcomes and combine ----
summary_tbl <- purrr::map(outcomes, ~summarize_outcome(df, .x)) %>%
  purrr::compact() %>%
  bind_rows() %>%
  arrange(match(Tropism, c("Overall","R5","X4","Mixed")), Outcome)

print(summary_tbl, n = nrow(summary_tbl))

# ---- Write to Excel ----
out_xlsx <- file.path(getwd(), "Outcome_Counts_Percentages_ByTropism.xlsx")
openxlsx::write.xlsx(summary_tbl, out_xlsx, overwrite = TRUE)
cat("Saved:", out_xlsx, "\n")

# =========================
# PLOT: grouped bars
# =========================
custom_colors <- c("R5" = "aquamarine3", "Mixed" = "steelblue3", "X4" = "coral")

desired_order <- c(
  "CDC3","CVD_baseline","INR_Willem","MALIG_NONAIDS",
  "MED_PSY.Antipsychotics","MED_PSY.Benzodiazepines","MH_PSY_D.Other",
  "MH_TR_END_D.DM2","OPP_AIDS_DEF","OTHER_AIDS_DEF",
  "Plaque","Steatosis_yesno","hi_vs_rest"
)

plot_df <- summary_tbl %>%
  filter(Tropism %in% c("Mixed","X4","R5")) %>%
  mutate(
    Outcome = factor(Outcome, levels = rev(desired_order)),
    Tropism = factor(Tropism, levels = c("Mixed","X4","R5")),
    label   = ifelse(is.na(Percent), NA_character_,
                     paste0(Percent, "% (", Events, "/", N_non_missing, ")"))
  )

y_var <- "Percent"  # change to "Events" if you prefer counts

pd <- position_dodge2(width = 0.8, preserve = "single")

p <- ggplot(plot_df, aes(x = Outcome, y = .data[[y_var]], fill = Tropism)) +
  geom_col(position = pd, width = 0.7, na.rm = TRUE) +
  geom_text(
    aes(y = .data[[y_var]], label = label),
    position = pd,
    hjust = 0,
    vjust = 0.5,
    size = 3,
    na.rm = TRUE
  ) +
  scale_fill_manual(values = custom_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.10))) +
  labs(
    x = NULL,
    y = ifelse(y_var == "Percent", "Percent with outcome (%)", "Number of events"),
    fill = "Tropism",
    title = "Outcome counts/percentages by Tropism"
  ) +
  coord_flip(clip = "off") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0),
    axis.text.y = element_text(size = 10),
    legend.position = "top",
    panel.grid.major.y = element_blank()
  )

print(p)

ggsave(
  filename = file.path(getwd(), paste0("GroupedBar_Outcomes_by_Tropism_", y_var, ".pdf")),
  plot = p,
  width = 10, height = 12
)


```



