---
title: "2000HIV_Tropism_reservoir comparision"
author: "Kavita Mehta"
date: "2026-01-20"
output: html_document
---

#setup chunk
```{r}

suppressPackageStartupMessages({
  library(ggplot2)
  library(readxl)
  library(writexl)
  library(ggsignif)
  library(dplyr)
  library(scales)
  library(purrr)
  library(readr)
  library(tidyr)
  library(rstatix)
})

# -----------------------------
# WORKDIR + INPUT
# -----------------------------
setwd("R:/hcrc/Personal folders/Kavita Mehta/Manuscripts/2000HIV_Tropism/reservoir/nonpara")

file_path <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx"
```


#DNA reservoir Violin plots: Tropism comparisons (R5 vs X4 vs Mixed)_Figure4 & supplementary figure4
EC_persistent were excluded from analysis
No data transformation.
KW with post-hoc test
For each subset and each variable:


```{r}
# ============================================================
# Tropism comparisons + plots
# Reservoir + RNA variables
# FULL, INF_1y, cART_3Years, Post2014 subsets
# Kruskal–Wallis + Dunn’s multiple comparisons (BH-adjusted)
# P-value labels: 2 significant figures or <0.0001
# RNA variables: keep zeros for statistics, remove zeros only for log plots
# ============================================================

# -----------------------------
# READ
# -----------------------------
data <- read_excel(file_path)

# -----------------------------
# VARIABLES
# -----------------------------
columns_to_analyze <- c(
  "Total_million_avg",
  "Rainbow_DSI",
  "Defective_Rainbow",
  "intact_total_ratio",
  "longLTR_cp_Mio",
  "POL_cp_Mio"
)

rna_zero_keep_vars <- c("longLTR_cp_Mio", "POL_cp_Mio")

# -----------------------------
# COERCE NUMERICS
# -----------------------------
data <- data %>%
  mutate(
    TIME_NEG_TO_ART = suppressWarnings(as.numeric(TIME_NEG_TO_ART)),
    EARLY_CART      = suppressWarnings(as.numeric(EARLY_CART)),
    HIV_DATE        = suppressWarnings(as.numeric(HIV_DATE)),
    across(
      any_of(columns_to_analyze),
      ~ suppressWarnings(as.numeric(.))
    )
  )

# -----------------------------
# FILTER BASE COHORT
# -----------------------------
data <- data %>%
  filter(
    CONTROLLER_1B != "EC_persistent",
    Tropism %in% c("X4", "R5", "Mixed")
  )

columns_to_analyze <- intersect(columns_to_analyze, names(data))

if (length(columns_to_analyze) == 0) {
  stop("None of columns_to_analyze found in the dataset.")
}

# -----------------------------
# COLORS + ORDERS
# -----------------------------
custom_colors <- c(
  "R5" = "aquamarine3",
  "X4" = "coral",
  "Mixed" = "steelblue3"
)

tropism_order <- c("R5", "X4", "Mixed")

# ============================================================
# DEFINE SUBSETS
# ============================================================
df_full <- data

subset_conditions <- list(
  FULL        = df_full,
  INF_1y      = df_full %>% filter(INF_date_known_1y == 1),
  cART_3Years = df_full %>% filter(TIME_NEG_TO_ART <= 1095),
  Post2014    = df_full %>% filter(!is.na(HIV_DATE) & HIV_DATE > 2014)
)

subset_order <- c("FULL", "INF_1y", "cART_3Years", "Post2014")

# -----------------------------
# P-VALUE FORMATTER
# -----------------------------
format_p <- function(p) {
  ifelse(
    is.na(p),
    "NA",
    ifelse(
      p < 0.0001,
      "<0.0001",
      formatC(p, format = "g", digits = 2)
    )
  )
}

# -----------------------------
# SUMMARIZE COUNTS
# -----------------------------
summary_df <- bind_rows(
  lapply(names(subset_conditions), function(subset_name) {
    subset_conditions[[subset_name]] %>%
      filter(Tropism %in% tropism_order) %>%
      group_by(Tropism) %>%
      summarise(n = n(), .groups = "drop") %>%
      mutate(Subset = subset_name)
  })
)

print(summary_df)

# ============================================================
# Y-AXIS SCALE FUNCTION
# ============================================================
get_y_scale <- function(marker) {

  if (marker %in% c("Total_million_avg", "Defective_Rainbow", "Rainbow_DSI")) {
    return(
      scale_y_log10(
        limits = c(1e0, 1e5),
        breaks = 10^(0:5),
        labels = trans_format("log10", math_format(10^.x))
      )
    )
  }

  if (marker %in% c("intact_total_ratio")) {
    return(
      scale_y_log10(
        limits = c(1e-3, 1),
        breaks = c(1e-3, 1e-2, 1e-1, 1),
        labels = c("0.001", "0.01", "0.1", "1")
      )
    )
  }

  if (marker %in% c("longLTR_cp_Mio", "POL_cp_Mio")) {
    return(
      list(
        scale_y_log10(
          breaks = 10^(-1:5),
          labels = trans_format("log10", math_format(10^.x))
        ),
        coord_cartesian(ylim = c(0.1, 1e5), clip = "off")
      )
    )
  }

  NULL
}

# ============================================================
# KRUSKAL–WALLIS + DUNN (BH) + PER-SUBSET PLOTS
# ============================================================
p_values_list <- list()

for (subset_name in names(subset_conditions)) {

  current_data <- subset_conditions[[subset_name]] %>%
    filter(Tropism %in% tropism_order) %>%
    mutate(Tropism = factor(Tropism, levels = tropism_order))

  p_values_df <- data.frame()

  for (column_name in columns_to_analyze) {

    # -------------------------
    # FILTER FOR STATISTICS
    # RNA: keep zeros
    # Other reservoir variables: >0 only
    # -------------------------
    if (column_name %in% rna_zero_keep_vars) {
      df_filt <- current_data %>%
        filter(
          !is.na(.data[[column_name]]),
          is.finite(.data[[column_name]])
        )
    } else {
      df_filt <- current_data %>%
        filter(
          !is.na(.data[[column_name]]),
          is.finite(.data[[column_name]]),
          .data[[column_name]] > 0
        )
    }

    df_filt <- df_filt %>%
      mutate(Tropism = droplevels(Tropism))

    if (nrow(df_filt) == 0) next
    if (nlevels(df_filt$Tropism) < 2) next

    # -------------------------
    # GLOBAL KRUSKAL–WALLIS
    # -------------------------
    kw <- kruskal.test(df_filt[[column_name]] ~ df_filt$Tropism)
    kw_p <- kw$p.value

    # -------------------------
    # DUNN POST-HOC BH
    # -------------------------
    dunn_res <- df_filt %>%
      rstatix::dunn_test(
        formula = as.formula(paste0("`", column_name, "` ~ Tropism")),
        p.adjust.method = "BH"
      ) %>%
      mutate(
        Group1 = as.character(group1),
        Group2 = as.character(group2),
        P_value = p,
        Adjusted_P_value = p.adj
      )

    comparisons <- lapply(
      seq_len(nrow(dunn_res)),
      function(i) c(dunn_res$Group1[i], dunn_res$Group2[i])
    )

    raw_p_values <- dunn_res$P_value
    adjusted_p_values <- dunn_res$Adjusted_P_value

    # -------------------------
    # MEDIAN/IQR FROM STATS DATA
    # For RNA this includes zeros
    # -------------------------
    median_iqr <- df_filt %>%
      group_by(Tropism) %>%
      summarise(
        Median = median(.data[[column_name]], na.rm = TRUE),
        Q1 = quantile(.data[[column_name]], 0.25, na.rm = TRUE),
        Q3 = quantile(.data[[column_name]], 0.75, na.rm = TRUE),
        .groups = "drop"
      )

    # -------------------------
    # STORE RESULTS
    # -------------------------
    for (i in seq_along(comparisons)) {

      comp <- comparisons[[i]]

      p_values_df <- rbind(
        p_values_df,
        data.frame(
          Dataset = subset_name,
          Column = column_name,
          Analysis_Note = ifelse(
            column_name %in% rna_zero_keep_vars,
            "RNA stats: NA removed, zeros retained; plot excludes zeros for log scale",
            "Positive values only"
          ),
          Group1 = comp[1],
          Group2 = comp[2],
          N_Group1 = sum(df_filt$Tropism == comp[1]),
          N_Group2 = sum(df_filt$Tropism == comp[2]),
          Zero_Group1 = ifelse(
            column_name %in% rna_zero_keep_vars,
            sum(df_filt$Tropism == comp[1] & df_filt[[column_name]] == 0, na.rm = TRUE),
            NA
          ),
          Zero_Group2 = ifelse(
            column_name %in% rna_zero_keep_vars,
            sum(df_filt$Tropism == comp[2] & df_filt[[column_name]] == 0, na.rm = TRUE),
            NA
          ),
          Median_Group1 = median(df_filt[[column_name]][df_filt$Tropism == comp[1]], na.rm = TRUE),
          Median_Group2 = median(df_filt[[column_name]][df_filt$Tropism == comp[2]], na.rm = TRUE),
          IQR_Group1 = paste0(
            median_iqr$Q1[median_iqr$Tropism == comp[1]],
            " - ",
            median_iqr$Q3[median_iqr$Tropism == comp[1]]
          ),
          IQR_Group2 = paste0(
            median_iqr$Q1[median_iqr$Tropism == comp[2]],
            " - ",
            median_iqr$Q3[median_iqr$Tropism == comp[2]]
          ),
          KW_P_value = kw_p,
          KW_P_Label = format_p(kw_p),
          P_value = raw_p_values[i],
          Raw_P_Label = format_p(raw_p_values[i]),
          Adjusted_P_value = adjusted_p_values[i],
          Adj_P_Label = format_p(adjusted_p_values[i]),
          Test = "Kruskal–Wallis + Dunn (BH)",
          stringsAsFactors = FALSE
        )
      )
    }

    # -------------------------
    # X-AXIS LABELS
    # RNA: show n and zeros
    # Other variables: show n
    # -------------------------
    if (column_name %in% rna_zero_keep_vars) {
      n_labels <- df_filt %>%
        group_by(Tropism) %>%
        summarise(
          n = n(),
          zeros = sum(.data[[column_name]] == 0, na.rm = TRUE),
          .groups = "drop"
        ) %>%
        mutate(label = paste0(Tropism, "\n(n=", n, ", 0s=", zeros, ")"))
    } else {
      n_labels <- df_filt %>%
        group_by(Tropism) %>%
        summarise(n = n(), .groups = "drop") %>%
        mutate(label = paste0(Tropism, "\n(n=", n, ")"))
    }

    label_map <- setNames(n_labels$label, as.character(n_labels$Tropism))
    valid_levels <- label_map[intersect(tropism_order, names(label_map))]

    # -------------------------
    # DATA FOR PLOTTING
    # RNA: remove zeros only for log-axis plot
    # -------------------------
    if (column_name %in% rna_zero_keep_vars) {
      df_plot <- df_filt %>%
        filter(.data[[column_name]] > 0)
    } else {
      df_plot <- df_filt
    }

    df_plot <- df_plot %>%
      mutate(
        Tropism_n = factor(
          label_map[as.character(Tropism)],
          levels = valid_levels
        )
      )

    if (nrow(df_plot) == 0) next

    # -------------------------
    # MEDIAN LABELS
    # RNA: true median includes zeros, placement uses non-zero median
    # -------------------------
    if (column_name %in% rna_zero_keep_vars) {

      median_labels <- df_filt %>%
        group_by(Tropism) %>%
        summarise(
          true_median = median(.data[[column_name]], na.rm = TRUE),
          .groups = "drop"
        ) %>%
        left_join(
          df_plot %>%
            group_by(Tropism) %>%
            summarise(
              y_pos = median(.data[[column_name]], na.rm = TRUE),
              .groups = "drop"
            ),
          by = "Tropism"
        ) %>%
        mutate(
          Tropism_n = factor(
            label_map[as.character(Tropism)],
            levels = valid_levels
          )
        )

    } else {

      median_labels <- df_plot %>%
        group_by(Tropism) %>%
        summarise(
          true_median = median(.data[[column_name]], na.rm = TRUE),
          y_pos = median(.data[[column_name]], na.rm = TRUE),
          .groups = "drop"
        ) %>%
        mutate(
          Tropism_n = factor(
            label_map[as.character(Tropism)],
            levels = valid_levels
          )
        )
    }

    # -------------------------
    # VALID COMPARISONS FOR PLOT
    # -------------------------
    present_tropism <- levels(droplevels(df_plot$Tropism))

    keep_idx <- sapply(
      comparisons,
      function(x) all(x %in% present_tropism)
    )

    valid_comparisons <- comparisons[keep_idx]
    valid_adj_p <- adjusted_p_values[keep_idx]
    valid_p_labels <- format_p(valid_adj_p)

    valid_comparisons_labeled <- lapply(
      valid_comparisons,
      function(comp) unname(label_map[comp])
    )

    if (length(valid_comparisons) == 0) next

    # -------------------------
    # PLOT
    # -------------------------
    p <- ggplot(
      df_plot,
      aes(x = Tropism_n, y = .data[[column_name]], fill = Tropism)
    ) +
      geom_violin(trim = FALSE, alpha = 0.4) +
      geom_jitter(
        aes(color = Tropism),
        width = 0.2,
        alpha = 0.9,
        size = 0.5
      ) +
      geom_boxplot(
        width = 0.2,
        outlier.shape = NA,
        alpha = 0.5,
        color = "black"
      ) +
      geom_text(
        data = median_labels,
        aes(x = Tropism_n, y = y_pos, label = signif(true_median, 3)),
        vjust = -0.7,
        size = 3.5,
        fontface = "bold",
        color = "black",
        inherit.aes = FALSE
      ) +
      scale_fill_manual(values = custom_colors) +
      scale_color_manual(values = custom_colors) +
      get_y_scale(column_name) +
      labs(
        title = paste0(
          column_name,
          " - ",
          subset_name,
          " (KW p=",
          format_p(kw_p),
          ")"
        ),
        x = "",
        y = column_name
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(size = 22, color = "black"),
        axis.text.y = element_text(size = 22, color = "black")
      ) +
      geom_signif(
        comparisons = valid_comparisons_labeled,
        annotations = valid_p_labels,
        tip_length = 0,
        step_increase = 0.04,
        textsize = 6
      )

    ggsave(
      filename = paste0("R5X4_", column_name, "_", subset_name, ".pdf"),
      plot = p,
      width = 8,
      height = 6
    )
  }

  p_values_list[[subset_name]] <- p_values_df
}

# ============================================================
# EXPORT SAME EXCEL
# ============================================================
write_xlsx(
  p_values_list,
  "Tropism_Comparisons_All_Datasets_KW_Dunn.xlsx"
)

# ============================================================
# FACETED PLOTS
# ============================================================
plot_data <- bind_rows(
  lapply(names(subset_conditions), function(nm) {
    subset_conditions[[nm]] %>%
      mutate(Subset = nm)
  })
) %>%
  filter(Tropism %in% tropism_order) %>%
  mutate(
    Tropism = factor(Tropism, levels = tropism_order),
    Subset = factor(Subset, levels = subset_order)
  )

plot_precomputed <- function(marker) {

  if (marker %in% rna_zero_keep_vars) {
    df_stats <- plot_data %>%
      filter(
        !is.na(.data[[marker]]),
        is.finite(.data[[marker]])
      )

    df <- df_stats %>%
      filter(.data[[marker]] > 0)

  } else {
    df_stats <- plot_data %>%
      filter(
        !is.na(.data[[marker]]),
        is.finite(.data[[marker]]),
        .data[[marker]] > 0
      )

    df <- df_stats
  }

  if (nrow(df) == 0) return(NULL)

  if (marker %in% rna_zero_keep_vars) {
    counts <- df_stats %>%
      group_by(Subset, Tropism) %>%
      summarise(
        n = n(),
        zeros = sum(.data[[marker]] == 0, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(Label = paste0(Tropism, " (n=", n, ", 0s=", zeros, ")"))
  } else {
    counts <- df_stats %>%
      group_by(Subset, Tropism) %>%
      summarise(n = n(), .groups = "drop") %>%
      mutate(Label = paste0(Tropism, " (n=", n, ")"))
  }

  df <- df %>%
    left_join(counts, by = c("Subset", "Tropism")) %>%
    group_by(Subset) %>%
    mutate(
      Tropism_label = factor(
        Label,
        levels = Label[match(tropism_order, Tropism)]
      ),
      Tropism_num = as.numeric(factor(Tropism, levels = tropism_order))
    ) %>%
    ungroup()

  signif_df <- bind_rows(p_values_list, .id = "Subset") %>%
    filter(Column == marker) %>%
    mutate(
      P_Label = Adj_P_Label,
      Subset = factor(Subset, levels = subset_order),
      Group1_num = as.numeric(factor(Group1, levels = tropism_order)),
      Group2_num = as.numeric(factor(Group2, levels = tropism_order)),
      midpoint = (Group1_num + Group2_num) / 2
    )

  if (nrow(signif_df) == 0) return(NULL)

  sig_use <- signif_df %>%
    group_by(Subset) %>%
    mutate(
      y_pos = {
        yy <- df[[marker]][df$Subset == first(Subset)]
        max_y <- max(yy, na.rm = TRUE)
        span_y <- diff(range(yy, na.rm = TRUE))
        base_y <- max_y + span_y * 0.15
        step <- span_y * 0.50
        base_y + step * (row_number() - 1)
      }
    ) %>%
    ungroup()

  medians_df <- signif_df %>%
    select(Subset, Group1, Median_Group1) %>%
    rename(Tropism = Group1, Median = Median_Group1) %>%
    bind_rows(
      signif_df %>%
        select(Subset, Group2, Median_Group2) %>%
        rename(Tropism = Group2, Median = Median_Group2)
    ) %>%
    distinct() %>%
    mutate(
      Tropism_num = as.numeric(factor(Tropism, levels = tropism_order)),
      label = signif(Median, 3),
      y = ifelse(Median <= 0, NA, Median)
    ) %>%
    filter(!is.na(y))

  ggplot(
    df,
    aes(x = Tropism_label, y = .data[[marker]], fill = Tropism)
  ) +
    geom_violin(trim = FALSE, alpha = 0.4) +
    geom_jitter(
      aes(color = Tropism),
      width = 0.15,
      size = 0.5,
      alpha = 0.9
    ) +
    geom_boxplot(
      width = 0.2,
      outlier.shape = NA,
      alpha = 0.6,
      color = "black"
    ) +
    geom_segment(
      data = sig_use,
      aes(
        x = Group1_num,
        xend = Group2_num,
        y = y_pos,
        yend = y_pos
      ),
      inherit.aes = FALSE,
      linewidth = 0.4
    ) +
    geom_text(
      data = sig_use,
      aes(
        x = midpoint,
        y = y_pos * 1.03,
        label = P_Label
      ),
      inherit.aes = FALSE,
      size = 4
    ) +
    geom_text(
      data = medians_df,
      aes(
        x = Tropism_num,
        y = y,
        label = label
      ),
      inherit.aes = FALSE,
      size = 3.5,
      fontface = "bold"
    ) +
    facet_wrap(~ Subset, nrow = 1, scales = "free_x") +
    scale_fill_manual(values = custom_colors) +
    scale_color_manual(values = custom_colors) +
    get_y_scale(marker) +
    labs(
      title = marker,
      x = "Tropism",
      y = marker
    ) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 13, color = "black"),
      strip.text = element_text(face = "bold")
    )
}

for (marker in columns_to_analyze) {
  p <- plot_precomputed(marker)
  if (is.null(p)) next

  ggsave(
    filename = paste0("FacetedR5X4_", marker, "_KW_Dunn.pdf"),
    plot = p,
    width = 28,
    height = 8
  )
}

getwd()
```
#distribution check
```{r}
# ============================================================
# DISTRIBUTION CHECK: RAW vs LOG2 (FULL cohort)
# Each Marker × Transform gets its OWN panel with independent axes
# (facet_wrap free scales — raw and log never share an axis)
# Log2 = log2(x) for >0 vars; log2(x+1) for RNA zero-keep vars
# ============================================================

library(ggplot2)
library(dplyr)
library(tidyr)

df_dist <- data %>%
  filter(Tropism %in% tropism_order) %>%
  mutate(Tropism = factor(Tropism, levels = tropism_order))

dist_data <- bind_rows(lapply(columns_to_analyze, function(marker) {

  raw_vals <- df_dist %>%
    filter(!is.na(.data[[marker]])) %>%
    transmute(Marker = marker, Tropism,
              Transform = "Raw",
              value = .data[[marker]])

  if (marker %in% rna_zero_keep_vars) {
    log_vals <- df_dist %>%
      filter(!is.na(.data[[marker]]), .data[[marker]] >= 0) %>%
      transmute(Marker = marker, Tropism,
                Transform = "Log2",
                value = log2(.data[[marker]] + 1))
  } else {
    log_vals <- df_dist %>%
      filter(!is.na(.data[[marker]]), .data[[marker]] > 0) %>%
      transmute(Marker = marker, Tropism,
                Transform = "Log2",
                value = log2(.data[[marker]]))
  }

  bind_rows(raw_vals, log_vals)
}))

dist_data <- dist_data %>%
  mutate(
    Marker    = factor(Marker, levels = columns_to_analyze),
    Transform = factor(Transform, levels = c("Raw", "Log2"))
  )

# ============================================================
# DENSITY — one panel per Marker × Transform, independent axes
# Coloured by Tropism so group overlap is visible
# (drop fill/color = Tropism for a single pooled density)
# ============================================================
p_dist <- ggplot(dist_data, aes(x = value, fill = Tropism, color = Tropism)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~ Marker + Transform, scales = "free", ncol = 2) +
  scale_fill_manual(values = custom_colors) +
  scale_color_manual(values = custom_colors) +
  labs(
    title    = "Distribution check: Raw vs Log2 (FULL cohort)",
    subtitle = "Log2 = log2(x) for >0 variables; log2(x+1) for RNA variables",
    x = "Value", y = "Density"
  ) +
  theme_bw(base_size = 13) +
  theme(strip.text = element_text(face = "bold"))

ggsave("FULL_Distribution_Check_Raw_vs_Log2_density.pdf",
       p_dist, width = 12, height = 14)

# ============================================================
# BOXPLOT — one panel per Marker × Transform, free y
# x = Tropism so you see per-group spread on each scale
# ============================================================
p_dist_box <- ggplot(dist_data, aes(x = Tropism, y = value, fill = Tropism)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.4) +
  facet_wrap(~ Marker + Transform, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Raw vs Log2 distribution by tropism (FULL cohort)",
    x = "", y = "Value"
  ) +
  theme_bw(base_size = 13) +
  theme(strip.text = element_text(face = "bold"))

ggsave("FULL_Distribution_Check_Raw_vs_Log2_boxplot.pdf",
       p_dist_box, width = 12, height = 16)

p_dist
p_dist_box
```

#Parametric test

```{r}
# =====================================================================
# Tropism reservoir comparison: R5 vs X4 vs Mixed
# One-way ANOVA + Tukey HSD on a per-variable transform
#   - intact_total_ratio  -> arcsine-sqrt  asin(sqrt(x))   (bounded proportion)
#   - all other variables -> log2(x + 1)                    (copy-number-like)
# Zeros are RETAINED for every variable (both transforms are defined at 0).
# =====================================================================

# -----------------------------
# READ
# -----------------------------
data <- read_excel(file_path)

setwd("R:/hcrc/Personal folders/Kavita Mehta/Manuscripts/2000HIV_Tropism/reservoir/log_parametric")

# -----------------------------
# VARIABLES
# -----------------------------
columns_to_analyze <- c(
  "Total_million_avg",
  "Rainbow_DSI",
  "Defective_Rainbow",
  "intact_total_ratio",
  "longLTR_cp_Mio",
  "POL_cp_Mio"
)

# Variables treated as bounded proportions (arcsine-sqrt instead of log2)
ratio_vars <- c("intact_total_ratio")

# -----------------------------
# PER-VARIABLE TRANSFORM HELPERS
# -----------------------------
# Transform a value vector according to the variable type.
transform_value <- function(x, marker) {
  if (marker %in% ratio_vars) {
    asin(sqrt(pmin(pmax(x, 0), 1)))   # arcsine-sqrt; clamps to [0,1] as a guard
  } else {
    log2(x + 1)                       # log2(x+1); zeros -> 0
  }
}

# Human-readable description for the export tables.
transform_label <- function(marker) {
  if (marker %in% ratio_vars) {
    "asin(sqrt(x)); proportion, zeros retained"
  } else {
    "log2(x + 1); zeros retained"
  }
}

# Short tag for axis titles / plot labels.
transform_tag <- function(marker) {
  if (marker %in% ratio_vars) "asin(sqrt) " else "log2(x+1) "
}

# Validity filter for the raw values before transforming.
valid_raw <- function(x, marker) {
  if (marker %in% ratio_vars) {
    is.finite(x) & x >= 0 & x <= 1
  } else {
    is.finite(x) & x >= 0
  }
}

# -----------------------------
# COERCE NUMERICS
# -----------------------------
data <- data %>%
  mutate(
    TIME_NEG_TO_ART = suppressWarnings(as.numeric(TIME_NEG_TO_ART)),
    EARLY_CART      = suppressWarnings(as.numeric(EARLY_CART)),
    HIV_DATE        = suppressWarnings(as.numeric(HIV_DATE)),
    across(
      any_of(columns_to_analyze),
      ~ suppressWarnings(as.numeric(.))
    )
  )

# -----------------------------
# FILTER BASE COHORT
# -----------------------------
data <- data %>%
  filter(
    CONTROLLER_1B != "EC_persistent",
    Tropism %in% c("X4", "R5", "Mixed")
  )

columns_to_analyze <- intersect(columns_to_analyze, names(data))

if (length(columns_to_analyze) == 0) {
  stop("None of columns_to_analyze found in the dataset.")
}

# -----------------------------
# COLORS + ORDERS
# -----------------------------
custom_colors <- c(
  "R5" = "aquamarine3",
  "X4" = "coral",
  "Mixed" = "steelblue3"
)

tropism_order <- c("R5", "X4", "Mixed")

# ============================================================
# DEFINE SUBSETS
# ============================================================
df_full <- data

subset_conditions <- list(
  FULL        = df_full,
  INF_1y      = df_full %>% filter(INF_date_known_1y == 1),
  cART_3Years = df_full %>% filter(TIME_NEG_TO_ART <= 1095),
  Post2014    = df_full %>% filter(!is.na(HIV_DATE) & HIV_DATE > 2014)
)

subset_order <- c("FULL", "INF_1y", "cART_3Years", "Post2014")

# -----------------------------
# P-VALUE FORMATTER
# -----------------------------
format_p <- function(p) {
  ifelse(
    is.na(p),
    "NA",
    ifelse(
      p < 0.0001,
      "<0.0001",
      formatC(p, format = "g", digits = 2)
    )
  )
}

# -----------------------------
# SUMMARIZE COUNTS
# -----------------------------
summary_df <- bind_rows(
  lapply(names(subset_conditions), function(subset_name) {
    subset_conditions[[subset_name]] %>%
      filter(Tropism %in% tropism_order) %>%
      group_by(Tropism) %>%
      summarise(n = n(), .groups = "drop") %>%
      mutate(Subset = subset_name)
  })
)

print(summary_df)

# ============================================================
# Y-AXIS SCALE FUNCTIONS
# Proportions (arcsine) get a free axis so the signif brackets fit;
# log2 variables keep the fixed window from before.
# ============================================================
get_y_scale <- function(marker) {
  if (marker %in% ratio_vars) {
    scale_y_continuous()
  } else {
    scale_y_continuous(
      limits = c(-3, 16),
      breaks = c(0, 5, 10, 15),
      expand = expansion(mult = c(0.02, 0.20))
    )
  }
}

get_y_scale_faceted <- function(marker) {
  scale_y_continuous(
    n.breaks = 4
  )
}

# ============================================================
# TRANSFORM + ANOVA + TUKEY HSD (per-variable transform)
# ============================================================
p_values_list   <- list()
assumption_list <- list()

for (subset_name in names(subset_conditions)) {

  current_data <- subset_conditions[[subset_name]] %>%
    filter(Tropism %in% tropism_order) %>%
    mutate(Tropism = factor(Tropism, levels = tropism_order))

  p_values_df   <- data.frame()
  assumption_df <- data.frame()

  for (column_name in columns_to_analyze) {

    trans_column <- paste0(column_name, "_trans")

    # -------------------------
    # FILTER + PER-VARIABLE TRANSFORM (zeros retained)
    # -------------------------
    df_filt <- current_data %>%
      filter(valid_raw(.data[[column_name]], column_name)) %>%
      mutate(
        "{trans_column}" := transform_value(.data[[column_name]], column_name)
      )

    df_filt <- df_filt %>%
      mutate(Tropism = droplevels(Tropism))

    if (nrow(df_filt) == 0) next
    if (nlevels(df_filt$Tropism) < 2) next

    # -------------------------
    # GLOBAL ONE-WAY ANOVA
    # -------------------------
    anova_model <- aov(
      as.formula(paste0("`", trans_column, "` ~ Tropism")),
      data = df_filt
    )

    anova_summary <- summary(anova_model)
    anova_p <- anova_summary[[1]][["Pr(>F)"]][1]
    anova_f <- anova_summary[[1]][["F value"]][1]

    # -------------------------
    # ASSUMPTION CHECKS (Shapiro on residuals, Bartlett for variances)
    # -------------------------
    shapiro_p  <- NA
    bartlett_p <- NA

    if (length(residuals(anova_model)) >= 3 && length(residuals(anova_model)) <= 5000) {
      shapiro_p <- tryCatch(
        shapiro.test(residuals(anova_model))$p.value,
        error = function(e) NA
      )
    }

    bartlett_p <- tryCatch(
      bartlett.test(df_filt[[trans_column]] ~ df_filt$Tropism)$p.value,
      error = function(e) NA
    )

    assumption_df <- rbind(
      assumption_df,
      data.frame(
        Dataset = subset_name,
        Column_original = column_name,
        Column_tested = trans_column,
        Transformation = transform_label(column_name),
        N_total = nrow(df_filt),
        N_groups = nlevels(df_filt$Tropism),
        ANOVA_F_value = anova_f,
        ANOVA_P_value = anova_p,
        ANOVA_P_Label = format_p(anova_p),
        Shapiro_residual_P_value = shapiro_p,
        Shapiro_residual_P_Label = format_p(shapiro_p),
        Bartlett_P_value = bartlett_p,
        Bartlett_P_Label = format_p(bartlett_p),
        stringsAsFactors = FALSE
      )
    )

    # -------------------------
    # TUKEY HSD POST-HOC (on transformed scale)
    # -------------------------
    tukey_res <- df_filt %>%
      rstatix::tukey_hsd(
        formula = as.formula(paste0("`", trans_column, "` ~ Tropism"))
      ) %>%
      mutate(
        Group1 = as.character(group1),
        Group2 = as.character(group2),
        Difference_trans = estimate,
        Lower_CI_trans = conf.low,
        Upper_CI_trans = conf.high,
        P_value = p.adj,
        Adjusted_P_value = p.adj
      )

    comparisons <- lapply(
      seq_len(nrow(tukey_res)),
      function(i) c(tukey_res$Group1[i], tukey_res$Group2[i])
    )

    adjusted_p_values <- tukey_res$Adjusted_P_value

    # -------------------------
    # MEDIAN/IQR ORIGINAL SCALE
    # -------------------------
    median_iqr_original <- df_filt %>%
      group_by(Tropism) %>%
      summarise(
        Median_original = median(.data[[column_name]], na.rm = TRUE),
        Q1_original = quantile(.data[[column_name]], 0.25, na.rm = TRUE),
        Q3_original = quantile(.data[[column_name]], 0.75, na.rm = TRUE),
        Mean_original = mean(.data[[column_name]], na.rm = TRUE),
        SD_original = sd(.data[[column_name]], na.rm = TRUE),
        .groups = "drop"
      )

    # -------------------------
    # MEAN/SD + MEDIAN/IQR TRANSFORMED SCALE
    # -------------------------
    summary_trans <- df_filt %>%
      group_by(Tropism) %>%
      summarise(
        Mean_trans = mean(.data[[trans_column]], na.rm = TRUE),
        SD_trans = sd(.data[[trans_column]], na.rm = TRUE),
        Median_trans = median(.data[[trans_column]], na.rm = TRUE),
        Q1_trans = quantile(.data[[trans_column]], 0.25, na.rm = TRUE),
        Q3_trans = quantile(.data[[trans_column]], 0.75, na.rm = TRUE),
        .groups = "drop"
      )

    # -------------------------
    # STORE RESULTS
    # -------------------------
    for (i in seq_along(comparisons)) {

      comp <- comparisons[[i]]

      p_values_df <- rbind(
        p_values_df,
        data.frame(
          Dataset = subset_name,
          Column_original = column_name,
          Column_tested = trans_column,
          Transformation = transform_label(column_name),
          Transform_scale = ifelse(column_name %in% ratio_vars, "arcsine-sqrt", "log2(x+1)"),
          Test = "One-way ANOVA + Tukey HSD on transformed values",

          Group1 = comp[1],
          Group2 = comp[2],

          N_Group1 = sum(df_filt$Tropism == comp[1]),
          N_Group2 = sum(df_filt$Tropism == comp[2]),

          Zero_Group1 = sum(df_filt$Tropism == comp[1] & df_filt[[column_name]] == 0, na.rm = TRUE),
          Zero_Group2 = sum(df_filt$Tropism == comp[2] & df_filt[[column_name]] == 0, na.rm = TRUE),

          Mean_Group1_trans = summary_trans$Mean_trans[summary_trans$Tropism == comp[1]],
          Mean_Group2_trans = summary_trans$Mean_trans[summary_trans$Tropism == comp[2]],

          SD_Group1_trans = summary_trans$SD_trans[summary_trans$Tropism == comp[1]],
          SD_Group2_trans = summary_trans$SD_trans[summary_trans$Tropism == comp[2]],
          
          Mean_Group1_original = median_iqr_original$Mean_original[median_iqr_original$Tropism == comp[1]],
          Mean_Group2_original = median_iqr_original$Mean_original[median_iqr_original$Tropism == comp[2]],

          SD_Group1_original = median_iqr_original$SD_original[median_iqr_original$Tropism == comp[1]],
          SD_Group2_original = median_iqr_original$SD_original[median_iqr_original$Tropism == comp[2]],
          
          
          Median_Group1_original = median_iqr_original$Median_original[median_iqr_original$Tropism == comp[1]],
          Median_Group2_original = median_iqr_original$Median_original[median_iqr_original$Tropism == comp[2]],

          IQR_Group1_original = paste0(
            median_iqr_original$Q1_original[median_iqr_original$Tropism == comp[1]], " - ",
            median_iqr_original$Q3_original[median_iqr_original$Tropism == comp[1]]
          ),
          IQR_Group2_original = paste0(
            median_iqr_original$Q1_original[median_iqr_original$Tropism == comp[2]], " - ",
            median_iqr_original$Q3_original[median_iqr_original$Tropism == comp[2]]
          ),

          Difference_trans = tukey_res$Difference_trans[i],
          Lower_CI_trans = tukey_res$Lower_CI_trans[i],
          Upper_CI_trans = tukey_res$Upper_CI_trans[i],

          ANOVA_F_value = anova_f,
          ANOVA_P_value = anova_p,
          ANOVA_P_Label = format_p(anova_p),

          Tukey_Adjusted_P_value = adjusted_p_values[i],
          Tukey_Adj_P_Label = format_p(adjusted_p_values[i]),

          Shapiro_residual_P_value = shapiro_p,
          Bartlett_P_value = bartlett_p,

          stringsAsFactors = FALSE
        )
      )
    }

    # -------------------------
    # X-AXIS LABELS (show n and zeros for all variables)
    # -------------------------
    n_labels <- df_filt %>%
      group_by(Tropism) %>%
      summarise(
        n = n(),
        zeros = sum(.data[[column_name]] == 0, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(label = paste0(Tropism, "\n(n=", n, ", 0s=", zeros, ")"))

    label_map <- setNames(n_labels$label, as.character(n_labels$Tropism))
    valid_levels <- label_map[intersect(tropism_order, names(label_map))]

    df_plot <- df_filt %>%
      mutate(
        Tropism_n = factor(label_map[as.character(Tropism)], levels = valid_levels)
      )

    if (nrow(df_plot) == 0) next

    # -------------------------
    # MEDIAN LABELS (original-scale median, placed at transformed median)
    # -------------------------
    median_labels <- df_plot %>%
      group_by(Tropism) %>%
      summarise(
        true_median_original = median(.data[[column_name]], na.rm = TRUE),
        y_pos = median(.data[[trans_column]], na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        Tropism_n = factor(label_map[as.character(Tropism)], levels = valid_levels)
      )

    # -------------------------
    # VALID COMPARISONS FOR PLOT
    # -------------------------
    present_tropism <- levels(droplevels(df_plot$Tropism))

    keep_idx <- sapply(comparisons, function(x) all(x %in% present_tropism))

    valid_comparisons <- comparisons[keep_idx]
    valid_adj_p <- adjusted_p_values[keep_idx]
    valid_p_labels <- format_p(valid_adj_p)

    valid_comparisons_labeled <- lapply(
      valid_comparisons, function(comp) unname(label_map[comp])
    )

    if (length(valid_comparisons) == 0) next

    # -------------------------
    # PLOT ON TRANSFORMED VALUES
    # -------------------------
    p <- ggplot(
      df_plot,
      aes(x = Tropism_n, y = .data[[trans_column]], fill = Tropism)
    ) +
      geom_violin(trim = FALSE, alpha = 0.4) +
      geom_jitter(aes(color = Tropism), width = 0.2, alpha = 0.9, size = 0.5) +
      geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, color = "black") +
      geom_text(
        data = median_labels,
        aes(x = Tropism_n, y = y_pos, label = signif(true_median_original, 3)),
        vjust = -0.7, size = 3.5, fontface = "bold", color = "black", inherit.aes = FALSE
      ) +
      scale_fill_manual(values = custom_colors) +
      scale_color_manual(values = custom_colors) +
      get_y_scale(column_name) +
      labs(
        title = paste0(
          column_name, " - ", subset_name, " ",
          transform_tag(column_name), "ANOVA p=", format_p(anova_p)
        ),
        x = "",
        y = paste0(transform_tag(column_name), column_name)
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(size = 22, color = "black"),
        axis.text.y = element_text(size = 22, color = "black")
      ) +
      geom_signif(
        comparisons = valid_comparisons_labeled,
        annotations = valid_p_labels,
        tip_length = 0, step_increase = 0.04, textsize = 6
      )

    ggsave(
      filename = paste0("R5X4_", column_name, "_", subset_name, "_ANOVA_Tukey.pdf"),
      plot = p, width = 8, height = 6
    )
  }

  p_values_list[[subset_name]]   <- p_values_df
  assumption_list[[subset_name]] <- assumption_df
}

# ============================================================
# EXPORT EXCEL
# ============================================================
write_xlsx(
  c(
    p_values_list,
    list(ANOVA_Assumption_Checks = bind_rows(assumption_list))
  ),
  "Tropism_Comparisons_All_Datasets_ANOVA_Tukey.xlsx"
)

# ============================================================
# FACETED PLOTS
# ============================================================
plot_data <- bind_rows(
  lapply(names(subset_conditions), function(nm) {
    subset_conditions[[nm]] %>% mutate(Subset = nm)
  })
) %>%
  filter(Tropism %in% tropism_order) %>%
  mutate(
    Tropism = factor(Tropism, levels = tropism_order),
    Subset = factor(Subset, levels = subset_order)
  )

plot_precomputed <- function(marker) {

  trans_marker <- paste0(marker, "_trans")

  df <- plot_data %>%
    filter(valid_raw(.data[[marker]], marker)) %>%
    mutate("{trans_marker}" := transform_value(.data[[marker]], marker))

  if (nrow(df) == 0) return(NULL)

  counts <- df %>%
    group_by(Subset, Tropism) %>%
    summarise(
      n = n(),
      zeros = sum(.data[[marker]] == 0, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(Label = paste0(Tropism, " (n=", n, ", 0s=", zeros, ")"))

  df <- df %>%
    left_join(counts, by = c("Subset", "Tropism")) %>%
    group_by(Subset) %>%
    mutate(
      Tropism_label = factor(Label, levels = Label[match(tropism_order, Tropism)]),
      Tropism_num = as.numeric(factor(Tropism, levels = tropism_order))
    ) %>%
    ungroup()

  signif_df <- bind_rows(p_values_list, .id = "Subset") %>%
    filter(Column_original == marker) %>%
    mutate(
      P_Label = Tukey_Adj_P_Label,
      Subset = factor(Subset, levels = subset_order),
      Group1_num = as.numeric(factor(Group1, levels = tropism_order)),
      Group2_num = as.numeric(factor(Group2, levels = tropism_order)),
      midpoint = (Group1_num + Group2_num) / 2
    )

  if (nrow(signif_df) == 0) return(NULL)

  sig_use <- signif_df %>%
    group_by(Subset) %>%
    mutate(
      y_pos = {
        yy <- df[[trans_marker]][df$Subset == first(Subset)]
        max_y <- max(yy, na.rm = TRUE)
        span_y <- diff(range(yy, na.rm = TRUE))
        ifelse(span_y == 0, max_y + 1, max_y + span_y * 0.05) +
          ifelse(span_y == 0, 1, span_y * 0.05) * (row_number() - 1)
      }
    ) %>%
    ungroup()

  medians_df <- df %>%
    group_by(Subset, Tropism) %>%
    summarise(
      Median_original = median(.data[[marker]], na.rm = TRUE),
      Median_trans = median(.data[[trans_marker]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      Tropism_num = as.numeric(factor(Tropism, levels = tropism_order)),
      label = signif(Median_original, 3),
      y = Median_trans
    )

  ggplot(
    df,
    aes(x = Tropism_label, y = .data[[trans_marker]], fill = Tropism)
  ) +
    geom_violin(trim = FALSE, alpha = 0.4) +
    geom_jitter(aes(color = Tropism), width = 0.15, size = 0.5, alpha = 0.9) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.6, color = "black") +
    geom_segment(
      data = sig_use,
      aes(x = Group1_num, xend = Group2_num, y = y_pos, yend = y_pos),
      inherit.aes = FALSE, linewidth = 0.4
    ) +
    geom_text(
      data = sig_use,
      aes(x = midpoint, y = y_pos, label = P_Label),
      inherit.aes = FALSE, size = 4
    ) +
    geom_text(
      data = medians_df,
      aes(x = Tropism_num, y = y, label = label),
      inherit.aes = FALSE, size = 3.5, fontface = "bold"
    ) +
    facet_wrap(~ Subset, nrow = 1, scales = "free_x") +
    scale_fill_manual(values = custom_colors) +
    scale_color_manual(values = custom_colors) +
    get_y_scale_faceted(marker) +
    labs(
      title = paste0(marker, " ", transform_tag(marker), "ANOVA + Tukey"),
      x = "Tropism",
      y = paste0(transform_tag(marker), marker)
    ) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 13, color = "black"),
      strip.text = element_text(face = "bold")
    )
}

for (marker in columns_to_analyze) {
  p <- plot_precomputed(marker)
  if (is.null(p)) next

  ggsave(
    filename = paste0("FacetedR5X4_", marker, "_ANOVA_Tukey.pdf"),
    plot = p, width = 28, height = 8
  )
}

# ============================================================
# DISTRIBUTION CHECK: Raw vs Transformed (per-variable transform)
# log2(x+1) for copy-number variables, arcsine-sqrt for the ratio.
# Built on the FULL cohort, density by Tropism.
# ============================================================
dist_subset <- "FULL"   # change to any name in subset_conditions
 
dist_base <- subset_conditions[[dist_subset]] %>%
  filter(Tropism %in% tropism_order) %>%
  mutate(Tropism = factor(Tropism, levels = tropism_order))
 
dist_long <- bind_rows(lapply(columns_to_analyze, function(marker) {
 
  d <- dist_base %>%
    filter(valid_raw(.data[[marker]], marker)) %>%
    transmute(
      Tropism,
      Marker = marker,
      Raw       = .data[[marker]],
      Transform = transform_value(.data[[marker]], marker)
    )
 
  if (nrow(d) == 0) return(NULL)
 
  # tidy to long: one row per (sample, scale)
  bind_rows(
    d %>% transmute(Tropism, Marker, Scale = "Raw",       Value = Raw),
    d %>% transmute(
      Tropism, Marker,
      Scale = trimws(transform_tag(marker)),   # "log2(x+1)" or "asin(sqrt)"
      Value = Transform
    )
  )
}))
 
# Panel order: each marker's Raw then its Transform, side by side (ncol = 2)
panel_levels <- unlist(lapply(columns_to_analyze, function(m) {
  c(paste0(m, "\nRaw"), paste0(m, "\n", trimws(transform_tag(m))))
}))
 
dist_long <- dist_long %>%
  mutate(
    Panel = factor(paste0(Marker, "\n", Scale), levels = panel_levels)
  )
 
p_dist <- ggplot(dist_long, aes(x = Value, fill = Tropism, color = Tropism)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  facet_wrap(~ Panel, scales = "free", ncol = 2) +
  scale_fill_manual(values = custom_colors) +
  scale_color_manual(values = custom_colors) +
  labs(
    title = paste0("Distribution check: Raw vs Transformed (", dist_subset, " cohort)"),
    subtitle = "Transform: log2(x+1) for copy-number variables; arcsine-sqrt for intact_total_ratio",
    x = "Value",
    y = "Density"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "right"
  )
 
ggsave(
  filename = paste0("FULL_Distribution_Check_Raw_vs_Transformed_", dist_subset, ".pdf"),
  plot = p_dist, width = 10, height = 14
)
 
getwd()
```

#forestplot reservoir
FOREST PLOT: Tropism effect (X4 vs R5) on reservoir outcomes
Outcomes are analyzed on log2 scale:
we looked into total and Intact (Rainbow) reservoir
- log2(Total_million_avg)
- log2(Rainbow_DSI)

For each subset, we fit:
1) Unadjusted model:   log2(outcome) ~ Tropism
2) Adjusted model:     log2(outcome) ~ Tropism + covariates
for Intact reservoir, the total resrvoir size was also corrected for

Then we extract the coefficient for TropismX4 (difference in log2 scale)
and plot betas + 95% CI as a forest plot.
the beta coefficient and other details (based on excel output) in figure were added using Adobe illustrator.
```{r}
# ====== Tropism Effect on log(Total_million_avg) & log(Rainbow_DSI): Unadjusted vs Adjusted ======

# ---- 0) LOAD LIBRARIES ----
suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(purrr)
  library(openxlsx)
  library(tidyr)
  library(lubridate)
})

# ---- 1) READ DATA ----
df_full <- read_excel("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx")
setwd("R:/hcrc/Personal folders/Kavita Mehta/Manuscripts/2000HIV_Tropism/reservoir")

df_full <- df_full %>%
  mutate(
    TIME_NEG_TO_ART   = suppressWarnings(as.numeric(TIME_NEG_TO_ART)),
    AGE               = suppressWarnings(as.numeric(AGE)),
    Total_million_avg = suppressWarnings(as.numeric(Total_million_avg)),
    Rainbow_DSI       = suppressWarnings(as.numeric(Rainbow_DSI))
  )
# ---- 2) DEFINE SUBSETS ----
subset_list <- list(
  FULL        = df_full,
  INF_1y      = df_full %>% dplyr::filter(INF_date_known_1y == 1),
  cART_3Years <- df_full %>% 
  dplyr::filter(suppressWarnings(as.numeric(TIME_NEG_TO_ART)) <= 1095,
                TIME_NEG_DIAG == 1)
  #post2014  = df_full %>% dplyr::filter(HIV_DATE > 2014)
)

# ---- 3) VARIABLE-SPECIFIC COVARIATES ----
covariate_map <- list(
  Total_million_avg = c("AGE", "SEX_BIRTH"),
  Rainbow_DSI       = c("AGE", "SEX_BIRTH", "Total_million_avg")
)

# ---- 4) OUTCOME VARIABLES ----
vars <- c("Total_million_avg", "Rainbow_DSI")

# ---- 5) FIT MODELS & COLLECT RESULTS ----
results <- purrr::imap_dfr(subset_list, function(df0, subset_name) {
  # Filter to R5/X4 only and set factor order
  df <- df0 %>%
    dplyr::filter(Tropism %in% c("R5", "X4"), CONTROLLER_1B != "EC_persistent") %>%
    dplyr::mutate(Tropism = factor(Tropism, levels = c("R5", "X4")))

  purrr::map_dfr(vars, function(var) {

    # Prepare analysis df: numeric, >0 for log2, then log-transform
    df1 <- df %>%
      dplyr::mutate(.y = suppressWarnings(as.numeric(.data[[var]]))) %>%
      dplyr::mutate(.y = ifelse(.y > 0, .y, NA_real_)) %>%
      dplyr::filter(!is.na(.y)) %>%
      dplyr::mutate(log_var = log2(.y))
    
   # --- Usable N by group for this variable (post >0 + non-NA + log filter) ---
n_used <- df1 %>%
  dplyr::count(Tropism) %>%
  tidyr::pivot_wider(names_from = Tropism, values_from = n, values_fill = 0)

N_R5_used    <- n_used[["R5"]]
N_X4_used    <- n_used[["X4"]]
N_total_used <- N_R5_used + N_X4_used

    # Bail if too few rows or only one Tropism level
    if (nrow(df1) < 3 || length(unique(df1$Tropism)) < 2) {
      return(tibble::tibble(
        Subset = subset_name, Variable = var, Model = c("Unadjusted","Adjusted"),
        Estimate = NA_real_, CI_Lower = NA_real_, CI_Upper = NA_real_, P_value = NA_real_,
        N = nrow(df1), Covariates = NA_character_
      ))
    }

    # --- Build covariate set ---
    base_covs <- covariate_map[[var]]
    if (is.null(base_covs)) base_covs <- character(0)

    # Extra covariates only for INF_1y (use those present)
    extra_covs <- character(0)
    if (identical(subset_name, "INF_1y")) {
      candidates <- c("TIMETOCART", "CART_DURATION")
      extra_covs <- intersect(candidates, names(df1))
    }

    covs <- intersect(c(base_covs, extra_covs), names(df1))

    # Coerce character covariates to factors (do NOT coerce to numeric)
    for (cv in covs) {
      if (is.character(df1[[cv]])) df1[[cv]] <- factor(df1[[cv]])
    }

    # --- Fit models ---
    f_unadj <- stats::as.formula("log_var ~ Tropism")
    m_unadj <- stats::lm(f_unadj, data = df1)

    f_adj <- if (length(covs)) {
      stats::as.formula(paste0("log_var ~ Tropism + ", paste(covs, collapse = " + ")))
    } else {
      f_unadj
    }
    m_adj <- stats::lm(f_adj, data = df1)

    # Extractor for the TropismX4 coefficient
    pull_row <- function(fit) {
      cf <- summary(fit)$coefficients
      if (!"TropismX4" %in% rownames(cf)) {
        return(tibble::tibble(Estimate = NA_real_, CI_Lower = NA_real_, CI_Upper = NA_real_, P_value = NA_real_))
      }
      est <- cf["TropismX4", "Estimate"]
      p   <- cf["TropismX4", "Pr(>|t|)"]
      ci  <- suppressWarnings(stats::confint(fit, "TropismX4", level = 0.95))
      tibble::tibble(
        Estimate = est, CI_Lower = ci[1], CI_Upper = ci[2], P_value = p
      )
    }

    out_unadj <- pull_row(m_unadj) %>% dplyr::mutate(Model = "Unadjusted")
    out_adj   <- pull_row(m_adj)   %>% dplyr::mutate(Model = "Adjusted")

dplyr::bind_rows(out_unadj, out_adj) %>%
  dplyr::mutate(
    Subset = subset_name,
    Variable = var,
    N_total_used = N_total_used,
    N_R5_used    = N_R5_used,
    N_X4_used    = N_X4_used,
    Covariates = dplyr::case_when(
      Model == "Adjusted" & length(covs) ~ paste(covs, collapse = ", "),
      Model == "Adjusted"                ~ "none",
      TRUE                               ~ "none"
    )
  ) %>%
  dplyr::select(Subset, Variable, Model,
                Estimate, CI_Lower, CI_Upper, P_value,
                N_total_used, N_R5_used, N_X4_used, Covariates)

  })
})

# ---- 6A) ADD OVERALL R5/X4 COUNTS (pre log-filter) PER SUBSET ----
tropism_stats <- purrr::imap_dfr(subset_list, function(df0, subset_name) {
  df <- df0 %>%
    dplyr::filter(Tropism %in% c("R5", "X4"), CONTROLLER_1B != "EC_persistent")

  counts <- df %>%
    dplyr::count(Tropism) %>%
    dplyr::mutate(Subset = subset_name) %>%
    tidyr::pivot_wider(names_from = Tropism, values_from = n, values_fill = 0) %>%
    dplyr::mutate(
      Total = R5 + X4,
      Pct_R5 = round(100 * R5 / Total, 1),
      Pct_X4 = round(100 * X4 / Total, 1)
    ) %>%
    dplyr::select(
      Subset,
      N_R5 = R5, Pct_R5,
      N_X4 = X4, Pct_X4
    )

  counts
})

results <- dplyr::left_join(results, tropism_stats, by = "Subset")

# ---- 6B) SAVE RESULTS (MODELS ONLY) ----
wb1 <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb1, "Tropism_Log_Results")
openxlsx::writeData(wb1, "Tropism_Log_Results", results)
openxlsx::saveWorkbook(wb1, "Tropism_Log_Total_Rainbow_log2.xlsx", overwrite = TRUE)

# ---- 6C) MEDIANS (and IQR) BY TROPISM, PER SUBSET & VARIABLE ----
median_tbl <- purrr::imap_dfr(subset_list, function(df0, subset_name) {
  df <- df0 %>%
    dplyr::filter(Tropism %in% c("R5","X4"), CONTROLLER_1B != "EC_persistent")

  purrr::map_dfr(vars, function(var) {
    df %>%
      dplyr::filter(!is.na(.data[[var]])) %>%
      dplyr::group_by(Tropism) %>%
      dplyr::summarise(
        N      = dplyr::n(),
        Median = stats::median(.data[[var]], na.rm = TRUE),
        Q1     = stats::quantile(.data[[var]], 0.25, na.rm = TRUE, type = 2),
        Q3     = stats::quantile(.data[[var]], 0.75, na.rm = TRUE, type = 2),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        IQR = Q3 - Q1,
        Subset = subset_name,
        Variable = var
      ) %>%
      dplyr::select(Subset, Variable, Tropism, N, Median, Q1, Q3, IQR)
  })
})

median_wide <- median_tbl %>%
  tidyr::pivot_wider(
    names_from  = Tropism,
    values_from = c(N, Median, Q1, Q3, IQR),
    names_sep   = "_"
  ) %>%
  dplyr::relocate(Subset, Variable)

# ---- 6D) SAVE RESULTS + MEDIANS TO EXCEL ----
wb2 <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb2, "Tropism_Log_Results")
openxlsx::writeData(wb2, "Tropism_Log_Results", results)

openxlsx::addWorksheet(wb2, "Medians_by_Tropism_long")
openxlsx::writeData(wb2, "Medians_by_Tropism_long", median_tbl)

openxlsx::addWorksheet(wb2, "Medians_by_Tropism_wide")
openxlsx::writeData(wb2, "Medians_by_Tropism_wide", median_wide)

openxlsx::saveWorkbook(wb2, "Tropism_Log_Total_Rainbow_log10.xlsx", overwrite = TRUE)

# ---- 7) FORMAT FOR PLOTTING ----
subset_levels <- c("cART_3Years", "INF_1y", "FULL")
results_plot <- results %>%
  dplyr::mutate(
    Subset   = factor(Subset, subset_levels),
    y0       = as.numeric(Subset),
    shift    = if_else(Model == "Unadjusted", 0.2, -0.2),
    y_plot   = y0 + shift,
    Variable = factor(Variable, vars),
    p_lab    = if_else(P_value < 0.001, "<0.001", sprintf("%.3f", P_value))
  )

# Shared x-axis across facets: compute global CI bounds
x_min <- min(results_plot$CI_Lower, na.rm = TRUE)
x_max <- max(results_plot$CI_Upper, na.rm = TRUE)

# ---- 8) FOREST PLOT ----
p <- ggplot(results_plot, aes(x = Estimate, y = y_plot, color = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.1, size = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(label = p_lab), hjust = -0.2, size = 3) +
  scale_y_continuous(
    breaks = seq_along(subset_levels),
    labels = subset_levels
  ) +
  scale_color_manual(values = c(Unadjusted = "black", Adjusted = "brown")) +
  facet_wrap(~ Variable, ncol = 1) +  # shared x-axis across facets
  labs(
    x = expression("β (difference in log"[2]*" scale) for X4 vs R5"),
    y = NULL,
    title = "Unadjusted (black, top) vs Adjusted (brown, bottom)\nTropism Effect on log2(Total_million_avg) & log2(Rainbow_DSI)",
    color = ""
  ) +
  xlim(x_min, x_max) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    panel.spacing = grid::unit(3, "lines")
  )

# ---- 9) SAVE & DISPLAY PLOT ----
print(p)
ggsave("Forest_Total_Rainbow_log2.pdf", p, width = 10, height = 6)
```

#reservoir and different covariates
```{r}
# ============================================================
# CHUNK — Reservoir analysis (FULL cohort only) with confounder sets
# Outcomes: Total_million_avg, Rainbow_DSI
# Compare X4 vs R5 (exclude EC_persistent), log2-transform outcomes (>0)
# Runs lm() for each outcome × confounder-set and exports results + medians
# ============================================================

suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(openxlsx)
})

# ---- 1) READ DATA + SET WD ----
df_full <- read_excel("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx")
setwd("R:/hcrc/Personal folders/Kavita Mehta/Manuscripts/2000HIV_Tropism/reservoir")

# ---- 2) CONFIG ----
outcomes <- c("Total_million_avg", "Rainbow_DSI")

# confounder sets (as you specified)
confounder_sets <- list(
  NONE               = character(0),
  AGE                = c("AGE"),
  SEX                = c("SEX_BIRTH"),
  AGE_SEX            = c("AGE","SEX_BIRTH"),
  AGE_SEX_TIMETOcART = c("AGE","SEX_BIRTH","TIMETOCART"),
  AGE_SEX_CD4NADIR   = c("AGE","SEX_BIRTH","CD4_NADIR"),
  AGE_SEX_VL   = c("AGE","SEX_BIRTH","VL_ZENITH"),
  AGE_SEX_TIMETOCART_CD4NADIR   = c("AGE","SEX_BIRTH","TIMETOCART","CD4_NADIR"),
  AGE_SEX_TIMETOCART_CD4NADIR_VL   = c("AGE","SEX_BIRTH","TIMETOCART","CD4_NADIR","VL_ZENITH")
)

# ---- 3) BASE FILTER (FULL cohort only) ----
df0 <- df_full %>%
  filter(
    Tropism %in% c("R5","X4"),
    CONTROLLER_1B != "EC_persistent"
  ) %>%
  mutate(Tropism = factor(Tropism, levels = c("R5","X4")))

# overall N by tropism (pre log-filter)
tropism_counts_overall <- df0 %>%
  count(Tropism) %>%
  tidyr::pivot_wider(names_from = Tropism, values_from = n, values_fill = 0) %>%
  mutate(
    N_total = R5 + X4,
    Pct_R5 = round(100 * R5 / N_total, 1),
    Pct_X4 = round(100 * X4 / N_total, 1)
  )

# ---- 4) HELPERS ----
to_num <- function(x) suppressWarnings(as.numeric(x))

# clean a covariate in df: handle blanks + NA consistently
clean_covariate <- function(df, cv) {
  x <- df[[cv]]

  # If factor, convert to character first so we can trim blanks
  if (is.factor(x)) x <- as.character(x)

  # Character: trim and set "" to NA, then factor again
  if (is.character(x)) {
    x <- trimws(x)
    x[x == ""] <- NA_character_
    df[[cv]] <- factor(x)
    return(df)
  }

  # Logical: keep as is
  if (is.logical(x)) {
    df[[cv]] <- x
    return(df)
  }

  # Everything else: try numeric coercion (covers "numeric stored as text")
  df[[cv]] <- suppressWarnings(as.numeric(x))
  df
}

pull_tropism_row <- function(fit) {
  cf <- summary(fit)$coefficients
  if (!"TropismX4" %in% rownames(cf)) {
    return(tibble(Estimate = NA_real_, CI_Lower = NA_real_, CI_Upper = NA_real_, P_value = NA_real_))
  }
  est <- cf["TropismX4", "Estimate"]
  p   <- cf["TropismX4", "Pr(>|t|)"]
  ci  <- suppressWarnings(stats::confint(fit, "TropismX4", level = 0.95))
  tibble(Estimate = est, CI_Lower = ci[1], CI_Upper = ci[2], P_value = p)
}

# ---- 5) FIT MODELS: outcome × confounder-set ----
model_results <- purrr::map_dfr(outcomes, function(outcome) {

  # outcome filter: numeric, >0 only, then log2
  df_y <- df0 %>%
    mutate(.y_raw = to_num(.data[[outcome]])) %>%
    mutate(.y_raw = ifelse(.y_raw > 0, .y_raw, NA_real_)) %>%
    filter(!is.na(.y_raw)) %>%
    mutate(log_y = log2(.y_raw))

  # N used for outcome (post >0 + non-NA), BEFORE covariate complete-case
  n_used <- df_y %>%
    count(Tropism) %>%
    pivot_wider(names_from = Tropism, values_from = n, values_fill = 0)

  N_R5_used    <- n_used[["R5"]]
  N_X4_used    <- n_used[["X4"]]
  N_total_used <- N_R5_used + N_X4_used

  purrr::imap_dfr(confounder_sets, function(covs, set_name) {

    covs <- intersect(covs, names(df_y))

    # clean covariates: remove blanks/whitespace, coerce numeric where needed
    df1 <- df_y
    for (cv in covs) {
      df1 <- clean_covariate(df1, cv)
    }

    # complete-case for the model (this WILL drop NA + blanks we turned into NA)
    df1 <- df1 %>%
      filter(if (length(covs)) if_all(all_of(covs), ~ !is.na(.)) else TRUE)

    # N actually used in model (after complete-case)
    n_model <- df1 %>%
      count(Tropism) %>%
      pivot_wider(names_from = Tropism, values_from = n, values_fill = 0)

    N_model_R5    <- n_model[["R5"]]
    N_model_X4    <- n_model[["X4"]]
    N_model_total <- N_model_R5 + N_model_X4

    # bail if too small / single group
    if (N_model_total < 10 || n_distinct(df1$Tropism) < 2) {
      return(tibble(
        Subset = "FULL",
        Variable = outcome,
        Confounder_Set = set_name,
        Covariates = if (length(covs)) paste(covs, collapse = ", ") else "none",
        Estimate = NA_real_, CI_Lower = NA_real_, CI_Upper = NA_real_, P_value = NA_real_,
        N_total_used = N_total_used,
        N_model_total = N_model_total, N_model_R5 = N_model_R5, N_model_X4 = N_model_X4
      ))
    }

    fml <- if (length(covs)) {
      as.formula(paste0("log_y ~ Tropism + ", paste(covs, collapse = " + ")))
    } else {
      as.formula("log_y ~ Tropism")
    }

    fit <- lm(fml, data = df1)

    pull_tropism_row(fit) %>%
      mutate(
        Subset = "FULL",
        Variable = outcome,
        Confounder_Set = set_name,
        Covariates = if (length(covs)) paste(covs, collapse = ", ") else "none",
        N_total_used = N_total_used,
        N_model_total = N_model_total, N_model_R5 = N_model_R5, N_model_X4 = N_model_X4
      ) %>%
      select(Subset, Variable, Confounder_Set, Covariates,
             Estimate, CI_Lower, CI_Upper, P_value,
             N_total_used, N_model_total, N_model_R5, N_model_X4)
  })
})


# attach overall pre-filter counts (same for all rows)
model_results <- model_results %>%
  mutate(
    N_R5_overall  = tropism_counts_overall$R5,
    N_X4_overall  = tropism_counts_overall$X4,
    N_total_overall = tropism_counts_overall$N_total,
    Pct_R5_overall = tropism_counts_overall$Pct_R5,
    Pct_X4_overall = tropism_counts_overall$Pct_X4
  )

# ---- 6) MEDIANS/IQR (raw scale) in FULL cohort, per outcome & tropism ----
median_tbl <- purrr::map_dfr(outcomes, function(outcome) {
  df0 %>%
    mutate(.y = to_num(.data[[outcome]])) %>%
    filter(!is.na(.y)) %>%
    group_by(Tropism) %>%
    summarise(
      N      = n(),
      Median = median(.y, na.rm = TRUE),
      Q1     = quantile(.y, 0.25, na.rm = TRUE, type = 2),
      Q3     = quantile(.y, 0.75, na.rm = TRUE, type = 2),
      .groups = "drop"
    ) %>%
    mutate(
      IQR = Q3 - Q1,
      Subset = "FULL",
      Variable = outcome
    ) %>%
    select(Subset, Variable, Tropism, N, Median, Q1, Q3, IQR)
})

median_wide <- median_tbl %>%
  pivot_wider(
    names_from  = Tropism,
    values_from = c(N, Median, Q1, Q3, IQR),
    names_sep   = "_"
  ) %>%
  relocate(Subset, Variable)

# ---- 7) SAVE TO EXCEL ----
wb <- createWorkbook()
addWorksheet(wb, "Models_log2_X4vsR5")
writeData(wb, "Models_log2_X4vsR5", model_results)

addWorksheet(wb, "Medians_IQR_long")
writeData(wb, "Medians_IQR_long", median_tbl)

addWorksheet(wb, "Medians_IQR_wide")
writeData(wb, "Medians_IQR_wide", median_wide)

saveWorkbook(wb, "Reservoir_FULL_confounder_sets_log2.xlsx", overwrite = TRUE)

# ---- 8) QUICK VIEW ----
model_results %>%
  arrange(Variable, match(Confounder_Set, names(confounder_sets))) %>%
  print(n = 200)

# ============================================================
# FOREST PLOT — FULL cohort, all confounder sets (ALL BLACK)
# ============================================================

library(ggplot2)

plot_df <- model_results %>%
  mutate(
    Confounder_Set = factor(
      Confounder_Set,
      levels = rev(names(confounder_sets))  # reverse so NONE appears on top
    ),
    Variable = factor(Variable, levels = outcomes)
  )

p <- ggplot(plot_df,
            aes(x = Estimate,
                y = Confounder_Set)) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "black") +
  geom_errorbarh(aes(xmin = CI_Lower,
                     xmax = CI_Upper),
                 height = 0.2,
                 size = 0.8,
                 color = "black") +
  geom_point(size = 3,
             color = "black") +
  facet_wrap(~ Variable, ncol = 1) +
  labs(
    x = expression("β (log"[2]*" difference X4 vs R5)"),
    y = "Confounder Set",
    title = "Reservoir Analysis — FULL Cohort\nX4 vs R5 across different confounder sets"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text  = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

print(p)

ggsave("Forest_FULL_confounder_sets_log2.pdf",
       p,
       width = 8,
       height = 6)
```


#TimetoART pre and post 2015
```{r}
# ============================================================
# Time-to-ART comparison: R5 vs X4
# Primary test : Wilcoxon rank-sum (raw days)            -> median (IQR)
# Sensitivity  : Welch t-test on log2(days + 1)          -> parametric check
# Outputs: counts + median/IQR + both tests + plots + distribution check + Excel
# ============================================================

# ============================================================
# FUNCTION: DxEra (<2015 vs >=2015) time-to-ART comparison
# ============================================================
run_dxera_time_to_art <- function(dat,
                                  dataset_name = "FULL",
                                  era_cutoff = 2015,
                                  era_order  = c("Pre2015", "Post2014"),
                                  tx_order   = c("R5", "X4"),
                                  time_var   = "TIMETOCART",
                                  year_var   = "HIV_DATE",
                                  out_prefix = "DxEra_TimeToART_R5vsX4") {

  df_era <- dat %>%
    filter(Tropism %in% tx_order) %>%
    mutate(
      Tropism = factor(Tropism, levels = tx_order),

      DxEra = dplyr::case_when(
        !is.na(.data[[year_var]]) & .data[[year_var]] <  era_cutoff ~ "Pre2015",
        !is.na(.data[[year_var]]) & .data[[year_var]] >= era_cutoff ~ "Post2014",
        TRUE ~ NA_character_
      ),
      DxEra = factor(DxEra, levels = era_order),

      TIMEVAR = suppressWarnings(as.numeric(.data[[time_var]]))
    ) %>%
    filter(
      !is.na(DxEra),
      !is.na(TIMEVAR),
      is.finite(TIMEVAR),
      TIMEVAR >= 0
    )

  era_counts <- df_era %>%
    count(DxEra, Tropism, name = "n") %>%
    arrange(DxEra, Tropism)

  era_summary <- df_era %>%
    group_by(DxEra, Tropism) %>%
    summarise(
      n = n(),
      Median = median(TIMEVAR, na.rm = TRUE),
      Q1 = quantile(TIMEVAR, 0.25, na.rm = TRUE),
      Q3 = quantile(TIMEVAR, 0.75, na.rm = TRUE),
      # geometric-mean time (back-transformed log2 mean) for the parametric side
      GeoMean_days = 2^(mean(log2(TIMEVAR + 1), na.rm = TRUE)) - 1,
      .groups = "drop"
    ) %>%
    mutate(
      Median_IQR = paste0(
        round(Median, 1), " (",
        round(Q1, 1), " - ",
        round(Q3, 1), ")"
      )
    )

  # -------------------------
  # TESTS: Wilcoxon (raw, primary) + Welch t-test on log2(x+1) (sensitivity)
  # -------------------------
  era_tests <- df_era %>%
    group_by(DxEra) %>%
    summarise(
      wilcox_p = {
        x <- TIMEVAR[Tropism == "R5"]
        y <- TIMEVAR[Tropism == "X4"]
        if (length(x) >= 5 && length(y) >= 5) {
          wilcox.test(x, y, exact = FALSE)$p.value
        } else NA_real_
      },
      t_log2_p = {
        x <- TIMEVAR[Tropism == "R5"]
        y <- TIMEVAR[Tropism == "X4"]
        if (length(x) >= 5 && length(y) >= 5) {
          t.test(log2(x + 1), log2(y + 1), var.equal = FALSE)$p.value
        } else NA_real_
      },
      .groups = "drop"
    ) %>%
    mutate(
      wilcox_p_BH = p.adjust(wilcox_p, method = "BH"),
      t_log2_p_BH = p.adjust(t_log2_p, method = "BH"),
      signif = case_when(
        is.na(wilcox_p_BH) ~ "NA",
        wilcox_p_BH < 0.001 ~ "***",
        wilcox_p_BH < 0.01  ~ "**",
        wilcox_p_BH < 0.05  ~ "*",
        TRUE ~ "ns"
      ),
      p_label = ifelse(
        is.na(wilcox_p_BH),
        "Wilcoxon BH p=NA",
        paste0(
          "Wilcoxon BH p=", signif(wilcox_p_BH, 3), " (", signif, ")",
          "\nt(log2) BH p=", ifelse(is.na(t_log2_p_BH), "NA", signif(t_log2_p_BH, 3))
        )
      )
    )

  # -------------------------
  # PLOT (raw days)
  # -------------------------
  p_time_to_art <- ggplot(
    df_era,
    aes(x = Tropism, y = TIMEVAR, fill = Tropism)
  ) +
    geom_violin(trim = FALSE, alpha = 0.4) +
    geom_jitter(aes(color = Tropism), width = 0.15, size = 0.5, alpha = 0.9) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.6, color = "black") +
    facet_wrap(~ DxEra, nrow = 1, scales = "free_y") +
    scale_x_discrete(limits = tx_order) +
    scale_fill_manual(values = custom_colors) +
    scale_color_manual(values = custom_colors) +
    labs(
      title = paste0("Time to ART by diagnosis era (R5 vs X4) - ", dataset_name),
      x = "",
      y = paste0(time_var, " (days)")
    ) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 13, color = "black"),
      strip.text = element_text(face = "bold"),
      legend.position = "none"
    )

  # n labels at y = 0
  n_label_df <- era_counts %>%
    mutate(y = 0, label = paste0("n=", n))

  p_time_to_art <- p_time_to_art +
    geom_text(
      data = n_label_df,
      aes(x = Tropism, y = y, label = label),
      inherit.aes = FALSE, size = 4, vjust = 1.5
    )

  # p-value labels (Wilcoxon primary + t(log2) sensitivity)
  ypos_df <- df_era %>%
    group_by(DxEra) %>%
    summarise(y_pos = max(TIMEVAR, na.rm = TRUE) * 1.08, .groups = "drop") %>%
    left_join(era_tests, by = "DxEra") %>%
    mutate(x = "R5")

  p_time_to_art <- p_time_to_art +
    geom_text(
      data = ypos_df,
      aes(x = x, y = y_pos, label = p_label),
      inherit.aes = FALSE, hjust = 0, vjust = 1, size = 3.8
    )

  pdf_file  <- paste0(out_prefix, "_", dataset_name, ".pdf")

  ggsave(pdf_file, p_time_to_art, width = 14, height = 6)

  list(
    df_era = df_era,
    counts = era_counts,
    summary = era_summary,
    tests = era_tests,
    plot = p_time_to_art,
    pdf_file = pdf_file
  )
}


# ============================================================
# FUNCTION: INF_1y subset only - R5 vs X4 time-to-ART (no era split)
# ============================================================
run_inf1y_time_to_art_noera <- function(dat,
                                        dataset_name = "INF_1y_NoEra",
                                        tx_order = c("R5", "X4"),
                                        time_var = "TIMETOCART",
                                        out_prefix = "DxEra_TimeToART_R5vsX4") {

  df <- dat %>%
    filter(Tropism %in% tx_order) %>%
    mutate(
      Tropism = factor(Tropism, levels = tx_order),
      TIMEVAR = suppressWarnings(as.numeric(.data[[time_var]]))
    ) %>%
    filter(!is.na(TIMEVAR), is.finite(TIMEVAR), TIMEVAR >= 0)

  counts <- df %>%
    count(Tropism, name = "n") %>%
    mutate(label = paste0("n=", n))

  summary_tbl <- df %>%
    group_by(Tropism) %>%
    summarise(
      n = n(),
      Median = median(TIMEVAR, na.rm = TRUE),
      Q1 = quantile(TIMEVAR, 0.25, na.rm = TRUE),
      Q3 = quantile(TIMEVAR, 0.75, na.rm = TRUE),
      GeoMean_days = 2^(mean(log2(TIMEVAR + 1), na.rm = TRUE)) - 1,
      .groups = "drop"
    ) %>%
    mutate(
      Median_IQR = paste0(
        round(Median, 1), " (",
        round(Q1, 1), " - ",
        round(Q3, 1), ")"
      )
    )

  x <- df$TIMEVAR[df$Tropism == "R5"]
  y <- df$TIMEVAR[df$Tropism == "X4"]

  ok <- length(x) >= 5 && length(y) >= 5

  wilcox_p <- if (ok) wilcox.test(x, y, exact = FALSE)$p.value else NA_real_
  t_log2_p <- if (ok) t.test(log2(x + 1), log2(y + 1), var.equal = FALSE)$p.value else NA_real_

  tests <- tibble::tibble(
    Dataset = dataset_name,
    wilcox_p = wilcox_p,
    t_log2_p = t_log2_p,
    p_label = paste0(
      "Wilcoxon p=", ifelse(is.na(wilcox_p), "NA", signif(wilcox_p, 3)),
      "\nt(log2) p=", ifelse(is.na(t_log2_p), "NA", signif(t_log2_p, 3))
    )
  )

  p <- ggplot(df, aes(x = Tropism, y = TIMEVAR, fill = Tropism)) +
    geom_violin(trim = FALSE, alpha = 0.4) +
    geom_jitter(aes(color = Tropism), width = 0.15, size = 0.5, alpha = 0.9) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.6, color = "black") +
    scale_x_discrete(limits = tx_order) +
    scale_fill_manual(values = custom_colors) +
    scale_color_manual(values = custom_colors) +
    labs(
      title = paste0("Time to ART (R5 vs X4) - ", dataset_name),
      x = "",
      y = paste0(time_var, " (days)")
    ) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 13, color = "black"),
      legend.position = "none"
    )

  p <- p +
    geom_text(
      data = counts,
      aes(x = Tropism, y = 0, label = label),
      inherit.aes = FALSE, size = 4, vjust = 1.5
    )

  y_pos <- max(df$TIMEVAR, na.rm = TRUE) * 1.08

  p <- p +
    geom_text(
      data = data.frame(x = "R5", y = y_pos, label = tests$p_label),
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE, hjust = 0, vjust = 1, size = 3.8
    )

  pdf_file  <- paste0(out_prefix, "_", dataset_name, ".pdf")

  ggsave(pdf_file, p, width = 8, height = 6)

  list(
    plot = p,
    counts = counts,
    summary = summary_tbl,
    tests = tests,
    pdf_file = pdf_file
  )
}


# ============================================================
# FUNCTION: DISTRIBUTION CHECK
# Raw vs log2(x+1), density by Tropism, faceted by diagnosis era.
# Shows why Wilcoxon (raw) is primary and the parametric test is on log2.
# ============================================================
plot_time_to_art_distribution <- function(dat,
                                          dataset_name = "FULL",
                                          era_cutoff = 2015,
                                          era_order  = c("Pre2015", "Post2014"),
                                          tx_order   = c("R5", "X4"),
                                          time_var   = "TIMETOCART",
                                          year_var   = "HIV_DATE",
                                          out_prefix = "DxEra_TimeToART_R5vsX4") {

  df_era <- dat %>%
    filter(Tropism %in% tx_order) %>%
    mutate(
      Tropism = factor(Tropism, levels = tx_order),
      DxEra = dplyr::case_when(
        !is.na(.data[[year_var]]) & .data[[year_var]] <  era_cutoff ~ "Pre2015",
        !is.na(.data[[year_var]]) & .data[[year_var]] >= era_cutoff ~ "Post2014",
        TRUE ~ NA_character_
      ),
      DxEra = factor(DxEra, levels = era_order),
      TIMEVAR = suppressWarnings(as.numeric(.data[[time_var]]))
    ) %>%
    filter(!is.na(DxEra), !is.na(TIMEVAR), is.finite(TIMEVAR), TIMEVAR >= 0)

  if (nrow(df_era) == 0) return(NULL)

  dist_long <- bind_rows(
    df_era %>% transmute(Tropism, DxEra, Scale = "Raw (days)", Value = TIMEVAR),
    df_era %>% transmute(Tropism, DxEra, Scale = "log2(x+1)",  Value = log2(TIMEVAR + 1))
  ) %>%
    mutate(Scale = factor(Scale, levels = c("Raw (days)", "log2(x+1)")))

  # Independent axes for every era x scale panel (Raw and log2 must NOT share an axis,
  # and Pre/Post eras must not share one either). facet_wrap + scales="free" does this;
  # facet_grid would force a common x within each column and y within each row.
  panel_levels <- unlist(lapply(
    era_order,
    function(e) paste0(e, "\n", c("Raw (days)", "log2(x+1)"))
  ))
  dist_long <- dist_long %>%
    mutate(Panel = factor(paste0(DxEra, "\n", Scale), levels = panel_levels))

  p_dist <- ggplot(dist_long, aes(x = Value, fill = Tropism, color = Tropism)) +
    geom_density(alpha = 0.35, linewidth = 0.4) +
    facet_wrap(~ Panel, scales = "free", ncol = 2) +
    scale_fill_manual(values = custom_colors) +
    scale_color_manual(values = custom_colors) +
    labs(
      title = paste0("Time-to-ART distribution (R5 vs X4) - ", dataset_name),
      subtitle = "Raw is right-skewed (Wilcoxon, primary); log2(x+1) used for the parametric sensitivity test",
      x = "Time to ART", y = "Density"
    ) +
    theme_bw(base_size = 13) +
    theme(strip.text = element_text(face = "bold"))

  pdf_file <- paste0(out_prefix, "_", dataset_name, "_DistributionCheck.pdf")
  ggsave(pdf_file, p_dist, width = 11, height = 7)

  list(plot = p_dist, data = dist_long, pdf_file = pdf_file)
}


# ============================================================
# RUN ANALYSES
# ============================================================

# --- Subsets ---
data_INF_1y <- data %>%
  filter(INF_date_known_1y == 1)

# --- Distribution checks FIRST (justify test choice) ---
dist_full  <- plot_time_to_art_distribution(data,        dataset_name = "FULL")
dist_inf1y <- plot_time_to_art_distribution(data_INF_1y, dataset_name = "INF_1y")

# --- Comparisons: Wilcoxon (primary) + Welch t on log2 (sensitivity) ---
res_full <- run_dxera_time_to_art(
  dat = data,
  dataset_name = "FULL"
)

res_inf1y <- run_dxera_time_to_art(
  dat = data_INF_1y,
  dataset_name = "INF_1y"
)

res_inf1y_noera <- run_inf1y_time_to_art_noera(
  dat = data_INF_1y,
  dataset_name = "INF_1y_NoEra"
)

# --- Show plots ---
dist_full$plot
dist_inf1y$plot
res_full$plot
res_inf1y$plot
res_inf1y_noera$plot

# ============================================================
# SINGLE CONSOLIDATED EXCEL
# One workbook; one sheet per table, stacked across analyses
# and tagged by an Analysis column.
# ============================================================
tag <- function(df, analysis) dplyr::mutate(df, Analysis = analysis, .before = 1)

writexl::write_xlsx(
  list(
    Summary = dplyr::bind_rows(
      tag(res_full$summary,        "FULL_era"),
      tag(res_inf1y$summary,       "INF_1y_era"),
      tag(res_inf1y_noera$summary, "INF_1y_noera")
    ),
    Counts = dplyr::bind_rows(
      tag(res_full$counts,        "FULL_era"),
      tag(res_inf1y$counts,       "INF_1y_era"),
      tag(res_inf1y_noera$counts, "INF_1y_noera")
    ),
    Tests = dplyr::bind_rows(
      tag(res_full$tests,        "FULL_era"),
      tag(res_inf1y$tests,       "INF_1y_era"),
      tag(res_inf1y_noera$tests, "INF_1y_noera")
    )
  ),
  "DxEra_TimeToART_R5vsX4_AllResults.xlsx"
)
```
