---
title: "Olink DEP Analysis_Tropism (R5 vs X4)"
author: "Kavita Mehta (modified from Vicky's and Victoria's script)"
date: "2025-09-18"
output:
  html_document:
    toc: true
    toc_float: true
---

This script performs differential protein expression (DEP) analysis comparing R5- and X4-tropic HIV-1 in the 2000HIV cohort, using Olink Explore 3072 proximity extension assay data. The analysis follows confounder identification (conducted separately) and proceeds through data integration, PCA-based quality assessment, limma-based DEP testing in discovery and validation cohorts, and targeted violin plot visualisation of pre-selected immune proteins. All input file paths should be verified before running.
```{r setup-chunk, include=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      cache = TRUE)
```


# Load Required Libraries
```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(car)
  library(qqplotr)
  library(GGally)
  library(ggfortify)
  library(psych)
  library(corrplot)
  library(ggplot2)
  library(ggrepel)
  library(randomForest)
  library(caret)
  #library(vip)
  library(FactoMineR)
  library(factoextra)
  #library(clusterProfiler)
  library(enrichplot)
  library(org.Hs.eg.db)
  #library(ReactomePA)
  #library(gage)
 #library(KEGGREST)
  library(knitr)
})  

```

#load and merge datasets
Three input files are merged: the Olink NPX matrix (post-bridging normalisation and QC), the 2000HIV clinical export, and the tropism classification file. Participants classified as neither R5 nor X4 (including EC_persistent controllers) are excluded. Seasonality is encoded as sine and cosine terms of days since 1 January 2019 to model continuous seasonal variation without assuming a linear trend. Ethnicity and recruitment centre are converted to binary dummy variables for use as covariates in downstream linear models.
```{r}
## Create one dataset with OLINK data and clinical data
# Import OLINK data from the 2000HIV
library(dplyr)
library(readxl)
library(openxlsx)

# Read and clean the dataset
cov_dat_de_master <- openxlsx::read.xlsx("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/updated_OLINK_Explore3072_1910samples_2367proteins_after_bridging_normalization_QCed_21nov2022.xlsx") %>%
  na.omit()

clinical_df <- read.xlsx("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx") 

out_dir <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/Coreceptor"
# Compute days since Jan 1, 2019
clinical_df <- clinical_df %>%
  mutate(
    dayssincejan2019 = as.numeric(as.POSIXct(DATE_VISIT) - as.POSIXct("2019-01-01")),
    season_sin = sin((2 * pi * dayssincejan2019) / 365.2425),
    season_cos = cos((2 * pi * dayssincejan2019) / 365.2425),
    ETHNICITY = as.factor(ETHNICITY)
  )

# ================================
# Ethnicity and Center Dummies
# ================================
clinical_df <- clinical_df %>%
  mutate(
    ASIAN    = as.numeric(ETHNICITY == "Asian"),
    BLACK    = as.numeric(ETHNICITY == "Black"),
    HISPANIC = as.numeric(ETHNICITY == "Hispanic"),
    MIXED    = as.numeric(ETHNICITY == "Mixed"),
    WHITE    = as.numeric(ETHNICITY == "White"),
    CENTER_RUMC = as.numeric(CENTER == "RUMC")
  )

# Create dummy variable for CENTER
clinical_df <- clinical_df %>%
  mutate(CENTER_RUMC = as.integer(CENTER == "RUMC"))

# Merge datasets
cov_dat_de <- clinical_df %>%
  inner_join(cov_dat_de_master, by = c("ID" = "ID"))


# Subset to X4 and R5, remove EC_persistent, and set factor levels
cov_dat_de <- cov_dat_de %>%
  filter(Tropism %in% c("X4", "R5")) %>%
  filter(CONTROLLER_1B != "EC_persistent") %>%
  mutate(condition = factor(Tropism, levels = c("R5", "X4")))


# Handle -Inf values
cov_dat_de[cov_dat_de == -Inf] <- NA

```

#Import Genetic PCs
Genetic principal components (PCs) are included as covariates to control for population stratification in the Olink models. The PCA was performed externally on genome-wide SNP data after imputation and QC. The first five genetic PCs are merged into the analysis dataset by participant ID. 
```{r}

# Function to process genetic pcs data
process_genetic_pcs <- function(file_path, sample_table, pcs_to_select = 5) {
  # Read the data
  genetic_pcs <- read.table(file_path, row.names = 1)
  
  # Match the IDs and row names
  genetic_pcs <- genetic_pcs[match(sample_table$`Study ID`, as.character(genetic_pcs$V2)), ]
  genetic_pcs$V2 <- sample_table$`Study ID`
  row.names(genetic_pcs) <- row.names(sample_table)
  
  # Replace the NAs with 0
  genetic_pcs[is.na(genetic_pcs)] <- 0
  
  # Choose the first 5 PCs
  pcs_to_select <- pcs_to_select + 2
  genetic_pcs_covariates <- genetic_pcs[, 2:pcs_to_select]
  
  # Verify the order
  head(cbind(genetic_pcs$V2, sample_table$`Study ID`))
  head(genetic_pcs)
  head(genetic_pcs_covariates)
  
  return(genetic_pcs_covariates)
}

genetic_pcs <-openxlsx::read.xlsx("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/2000HIV_preimputed_GRCh38_2000HIV_all_ethnicities_clean_final_complete_cohort_wo_outliers.pca.xlsx") %>% as.data.frame()



# Function to add principal components to sample table
add_pcs_to_sample_table <- function(sample_table, genetic_pcs_covariates) {
  sample_table$PC1_gen <- genetic_pcs_covariates[, 1]
  sample_table$PC2_gen <- genetic_pcs_covariates[, 2]
  sample_table$PC3_gen <- genetic_pcs_covariates[, 3]
  sample_table$PC4_gen <- genetic_pcs_covariates[, 4]
  sample_table$PC5_gen <- genetic_pcs_covariates[, 5]
  
  return(sample_table)
}

# Add PCs to All samples sample table
cov_dat_de <- cov_dat_de %>%
  left_join(genetic_pcs, by = "ID")

#SUBSET COHORT
cov_dat_de_complete <- cov_dat_de
cov_dat_de_validation <- subset(cov_dat_de,COHORT == "VALIDATION")
cov_dat_de_discovery <- subset(cov_dat_de,COHORT == "DISCOVERY")
```



#PCA regression analysis
PCA is run on the full Olink protein matrix (discovery cohort) to assess whether technical and biological covariates explain systematic variance in the proteome. Adjusted R² from univariate regression of each PC score against each covariate is visualised as a heatmap. PCA is then repeated on the residual matrix after regressing out the selected confounders, to confirm that the correction attenuates technical variance without removing biological signal. The protein matrix is identified by column position between ADA_Inflammation and ZPR1_Neurology_II, which are the first and last proteins in the Olink panel block in this dataset; verify these anchors if the column order changes.
```{r}
# ================================
# PCA on OLINK Data: Before and After Correction
# Heatmap = adjusted R2, pheatmap format
# ================================

suppressPackageStartupMessages({
  library(openxlsx)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(FactoMineR)
  library(factoextra)
  library(pheatmap)
  library(corrplot)
})

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

# ================================
# Dataset
# ================================

pca_df <- cov_dat_de_discovery

pca_df[pca_df == -Inf] <- NA
pca_df[pca_df == Inf]  <- NA

# ================================
# Required covariates, exact order
# ================================

required_cols <- c(
  "ID", "AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", "CENTER",
  "season_sin", "season_cos", "CENTER_RUMC", "ASIAN", "BLACK",
  "HISPANIC", "MIXED", "WHITE", "PANDEMIC_BEFOREAFTER", "COVID19", "COVID_VACC",
  "TIMETOLAB", "MUTHIV", "CD4_NADIR", "CD4CD8_PRECART", "VL_ZENITH", "VL_PRECART",
  "HEPC", "HEPA_BASELINE", "HEPB_BASELINE", "CART_DURATION", "HIV_STAGE_CDC",
  "EARLY_CART", "HIV_DURATION", paste0("PC", 1:5)
)

missing_cols <- setdiff(required_cols, colnames(pca_df))
if (length(missing_cols) > 0) {
  message("Missing columns: ", paste(missing_cols, collapse = ", "))
}

required_cols_present <- required_cols[required_cols %in% colnames(pca_df)]
heatmap_cols          <- required_cols_present[required_cols_present != "ID"]

# ================================
# Olink protein columns
# ================================

start_index <- which(colnames(pca_df) == "ADA_Inflammation")
end_index   <- which(colnames(pca_df) == "ZPR1_Neurology_II")

if (length(start_index) == 0 || length(end_index) == 0) {
  stop("Could not find ADA_Inflammation or ZPR1_Neurology_II in pca_df.")
}

selected_columns <- pca_df[, start_index:end_index] %>%
  mutate(across(everything(), as.numeric))

selected_columns <- selected_columns[, colSums(!is.na(selected_columns)) > 0, drop = FALSE]
selected_columns <- selected_columns[,
  apply(selected_columns, 2, function(x) sd(x, na.rm = TRUE) > 0),
  drop = FALSE
]

complete_rows        <- complete.cases(selected_columns)
selected_columns_pca <- selected_columns[complete_rows, , drop = FALSE]
pca_meta             <- pca_df[complete_rows, , drop = FALSE]

# ================================
# Custom scree plot function
# ================================

plot_scree_custom <- function(pca_obj, title, n_dims = 25, pc_cutoff = 5) {

  eig_vals  <- pca_obj$eig[1:n_dims, ]
  dims      <- 1:n_dims
  pct       <- eig_vals[, 2]
  cum_pct   <- eig_vals[, 3]

  cutoff_cum <- cum_pct[pc_cutoff]

  df <- data.frame(
    Dimension  = dims,
    Variance   = pct,
    Cumulative = cum_pct
  )

  ggplot(df, aes(x = Dimension)) +
    geom_bar(aes(y = Variance), stat = "identity",
             fill = "steelblue3", alpha = 0.8) +
    geom_line(aes(y = Variance), color = "black", linewidth = 0.6) +
    geom_point(aes(y = Variance), color = "black", size = 2) +
    geom_line(aes(y = Cumulative), color = "grey50", linewidth = 0.8) +
    geom_point(aes(y = Cumulative), color = "grey50", size = 2) +
    geom_vline(xintercept = pc_cutoff, linetype = "dashed",
               color = "black", linewidth = 0.6) +
    geom_hline(yintercept = cutoff_cum, linetype = "dotted",
               color = "grey30", linewidth = 0.5) +
    annotate("text",
             x     = pc_cutoff + 0.5,
             y     = max(pct) * 0.95,
             label = paste0("PC", pc_cutoff, "\n(cum. ", round(cutoff_cum, 1), "%)"),
             hjust = 0, size = 3.5, color = "black") +
    geom_text(aes(y = Variance,
                  label = paste0(round(Variance, 1), "%")),
              vjust = -0.5, size = 3, color = "black") +
    scale_x_continuous(breaks = dims) +
    scale_y_continuous(
      name      = "Percentage of explained variance",
      sec.axis  = sec_axis(~ ., name = "Cumulative variance (%)")
    ) +
    labs(title = title, x = "Dimensions") +
    theme_classic(base_size = 13) +
    theme(
      axis.text.x  = element_text(size = 9),
      axis.title.y = element_text(size = 11),
      plot.title   = element_text(face = "bold")
    )
}

# ================================
# PCA + adjusted R2 pheatmap function
# ================================

plot_PCA_regression <- function(pca, meta_data, meta_variables,
                                title = "PC contribution") {

  pca_res <- as.data.frame(pca$ind$coord[, 1:5, drop = FALSE])
  colnames(pca_res) <- paste0("PC", 1:ncol(pca_res))

  meta_variables_present <- meta_variables[meta_variables %in% colnames(meta_data)]

  pca_adj_r_sq <- matrix(
    NA_real_,
    nrow = length(meta_variables_present),
    ncol = ncol(pca_res)
  )
  rownames(pca_adj_r_sq) <- meta_variables_present
  colnames(pca_adj_r_sq) <- colnames(pca_res)

  for (i in seq_len(ncol(pca_res))) {
    for (j in seq_along(meta_variables_present)) {

      var_name <- meta_variables_present[j]

      tmp <- data.frame(
        pc_value = as.numeric(pca_res[[i]]),
        covar    = meta_data[[var_name]]
      )
      tmp <- tmp[complete.cases(tmp), , drop = FALSE]

      if (nrow(tmp) < 10 || length(unique(tmp$covar)) < 2) {
        pca_adj_r_sq[j, i] <- NA_real_
        next
      }

      if (is.character(tmp$covar) || is.factor(tmp$covar)) {
        tmp$covar <- as.factor(tmp$covar)
      } else {
        tmp$covar <- suppressWarnings(as.numeric(tmp$covar))
      }

      tmp <- tmp[complete.cases(tmp), , drop = FALSE]

      if (nrow(tmp) < 10 || length(unique(tmp$covar)) < 2) {
        pca_adj_r_sq[j, i] <- NA_real_
        next
      }

      fit <- tryCatch(
        lm(pc_value ~ covar, data = tmp),
        error = function(e) NULL
      )

      if (!is.null(fit)) {
        pca_adj_r_sq[j, i] <- as.numeric(summary(fit)$adj.r.squared)
      }
    }
  }

  storage.mode(pca_adj_r_sq) <- "numeric"

  pheatmap::pheatmap(
    pca_adj_r_sq,
    cluster_rows   = FALSE,
    cluster_cols   = FALSE,
    show_rownames  = TRUE,
    show_colnames  = TRUE,
    scale          = "none",
    main           = title,
    display_numbers = TRUE,
    number_format  = "%.2f",
    fontsize        = 12,
    fontsize_number = 10,
    color  = colorRampPalette(c("blue", "white", "red"))(50),
    breaks = seq(-0.5, 0.5, length.out = 51)
  )
}

# ================================
# PCA before correction
# ================================

mypca <- PCA(selected_columns_pca, ncp = 5, graph = FALSE)

# Scree plot before correction
pca_scree_plot <- plot_scree_custom(
  mypca,
  title     = "Scree plot: Olink PCA before correction",
  n_dims    = 25,
  pc_cutoff = 5
)

ggsave(
  file.path(out_dir, "Olink_PCA_Scree_Plot_before_correction.pdf"),
  plot = pca_scree_plot, width = 12, height = 6
)
ggsave(
  file.path(out_dir, "Olink_PCA_Scree_Plot_before_correction.tiff"),
  plot = pca_scree_plot, width = 12, height = 6,
  dpi = 600, units = "in", device = "tiff"
)

# PCA scatter plot before correction
pca_before_df     <- as.data.frame(mypca$ind$coord)
pca_before_df$ID  <- pca_meta$ID
group_var         <- "Tropism"

pca_before_df$Group <- if (group_var %in% colnames(pca_meta)) {
  as.factor(pca_meta[[group_var]])
} else {
  "All samples"
}

var_pc1 <- round(mypca$eig[1, 2], 1)
var_pc2 <- round(mypca$eig[2, 2], 1)

pca_before_plot <- ggplot(
  pca_before_df, aes(x = Dim.1, y = Dim.2, color = Group)
) +
  geom_point(size = 2.2, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    title = "Olink PCA before correction",
    x     = paste0("PC1 (", var_pc1, "%)"),
    y     = paste0("PC2 (", var_pc2, "%)"),
    color = group_var
  )

ggsave(
  file.path(out_dir, "Olink_PCA_before_correction.pdf"),
  plot = pca_before_plot, width = 7, height = 6
)
ggsave(
  file.path(out_dir, "Olink_PCA_before_correction.tiff"),
  plot = pca_before_plot, width = 7, height = 6,
  dpi = 600, units = "in", device = "tiff"
)

# Adjusted R2 heatmap before correction
pdf(file.path(out_dir, "PCA_Regression_Heatmap_before_correction.pdf"),
    width = 10, height = 12)
plot_PCA_regression(mypca, pca_meta, heatmap_cols,
                    title = "PCA regression heatmap before correction")
dev.off()

tiff(file.path(out_dir, "PCA_Regression_Heatmap_before_correction.tiff"),
     width = 10, height = 12, units = "in", res = 600)
plot_PCA_regression(mypca, pca_meta, heatmap_cols,
                    title = "PCA regression heatmap before correction")
dev.off()

# ================================
# Correlation matrix of covariates
# ================================

meta_data_cor <- pca_meta %>%
  dplyr::select(all_of(heatmap_cols)) %>%
  mutate(across(everything(), ~ as.numeric(as.factor(.))))

cor_matrix <- cor(meta_data_cor, use = "pairwise.complete.obs")

pdf(file.path(out_dir, "Correlation_Matrix_Plot.pdf"), width = 12, height = 10)
corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 50,
  tl.cex      = 0.6,
  addCoef.col = "black",
  number.cex  = 0.4,
  col         = colorRampPalette(c("blue", "white", "red"))(200)
)
dev.off()

# ================================
# Residuals PCA after correction
# ================================

covariates_for_correction <- c(
  "AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT",
  "CENTER_RUMC", "PC1", "season_cos", "season_sin",
  "TIMETOLAB", "CD4CD8_PRECART", "PANDEMIC_BEFOREAFTER"
)

covariates_for_correction <- covariates_for_correction[
  covariates_for_correction %in% colnames(pca_meta)
]

residual_matrix <- matrix(
  NA_real_,
  nrow = nrow(pca_meta),
  ncol = ncol(selected_columns_pca)
)
colnames(residual_matrix) <- colnames(selected_columns_pca)
rownames(residual_matrix) <- pca_meta$ID

for (marker_col in colnames(selected_columns_pca)) {

  model_df <- data.frame(
    row_id = seq_len(nrow(pca_meta)),
    marker = selected_columns_pca[[marker_col]],
    pca_meta[, covariates_for_correction, drop = FALSE]
  )
  model_df <- model_df[complete.cases(model_df), , drop = FALSE]

  if (nrow(model_df) < 10 || sd(model_df$marker, na.rm = TRUE) == 0) next

  fit <- tryCatch(
    lm(
      as.formula(paste("marker ~",
                       paste(covariates_for_correction, collapse = " + "))),
      data = model_df
    ),
    error = function(e) NULL
  )

  if (!is.null(fit)) {
    residual_matrix[model_df$row_id, marker_col] <- residuals(fit)
  }
}

residuals_df <- as.data.frame(residual_matrix)
residuals_df <- residuals_df[, colSums(!is.na(residuals_df)) > 0, drop = FALSE]
residuals_df <- residuals_df[,
  apply(residuals_df, 2, function(x) sd(x, na.rm = TRUE) > 0),
  drop = FALSE
]

complete_rows_res <- complete.cases(residuals_df)
residuals_pca     <- residuals_df[complete_rows_res, , drop = FALSE]
pca_meta_res      <- pca_meta[complete_rows_res, , drop = FALSE]

mypcares <- PCA(residuals_pca, ncp = 5, graph = FALSE)

# Scree plot after correction
pca_scree_plot_after <- plot_scree_custom(
  mypcares,
  title     = "Scree plot: Olink PCA after correction",
  n_dims    = 25,
  pc_cutoff = 5
)

ggsave(
  file.path(out_dir, "Olink_PCA_Scree_Plot_after_correction.pdf"),
  plot = pca_scree_plot_after, width = 12, height = 6
)
ggsave(
  file.path(out_dir, "Olink_PCA_Scree_Plot_after_correction.tiff"),
  plot = pca_scree_plot_after, width = 12, height = 6,
  dpi = 600, units = "in", device = "tiff"
)

# PCA scatter plot after correction
pca_after_df     <- as.data.frame(mypcares$ind$coord)
pca_after_df$ID  <- pca_meta_res$ID

pca_after_df$Group <- if (group_var %in% colnames(pca_meta_res)) {
  as.factor(pca_meta_res[[group_var]])
} else {
  "All samples"
}

var_pc1_after <- round(mypcares$eig[1, 2], 1)
var_pc2_after <- round(mypcares$eig[2, 2], 1)

pca_after_plot <- ggplot(
  pca_after_df, aes(x = Dim.1, y = Dim.2, color = Group)
) +
  geom_point(size = 2.2, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    title = "Olink PCA after correction",
    x     = paste0("PC1 (", var_pc1_after, "%)"),
    y     = paste0("PC2 (", var_pc2_after, "%)"),
    color = group_var
  )

ggsave(
  file.path(out_dir, "Olink_PCA_after_correction.pdf"),
  plot = pca_after_plot, width = 7, height = 6
)
ggsave(
  file.path(out_dir, "Olink_PCA_after_correction.tiff"),
  plot = pca_after_plot, width = 7, height = 6,
  dpi = 600, units = "in", device = "tiff"
)

# Adjusted R2 heatmap after correction
heatmap_cols_res <- heatmap_cols[heatmap_cols %in% colnames(pca_meta_res)]

pdf(file.path(out_dir, "PCA_Regression_Heatmap_after_correction.pdf"),
    width = 10, height = 12)
plot_PCA_regression(mypcares, pca_meta_res, heatmap_cols_res,
                    title = "PCA regression heatmap after correction")
dev.off()

tiff(file.path(out_dir, "PCA_Regression_Heatmap_after_correction.tiff"),
     width = 10, height = 12, units = "in", res = 600)
plot_PCA_regression(mypcares, pca_meta_res, heatmap_cols_res,
                    title = "PCA regression heatmap after correction")
dev.off()
```

#linearity check
Before using PCA-derived scores as the outcome in regression models, three assumptions are checked for each PC-covariate pair: normality of residuals (Q-Q plots), homoscedasticity (residuals versus fitted values with Breusch-Pagan test), and linearity for continuous covariates (natural cubic spline smoother compared to linear fit by ANOVA). Results are saved as an Excel file and diagnostic plots are saved as PDF and TIFF. Only PC1 to PC5 are shown in plots for readability; all ten PCs are included in the saved table.
```{r}
# ================================
# PCA regression assumption checks
# Reviewer response:
# Normality, homoscedasticity, and linearity
#
# Models: PC score ~ covariate
#
# Outputs:
# 1) Q-Q plots: normality
# 2) Residuals vs fitted: homoscedasticity
# 3) Residuals vs fitted for continuous covariates only: linearity
#    with natural cubic spline smoother, not LOESS
# ================================

suppressPackageStartupMessages({
  library(dplyr)
  library(FactoMineR)
  library(lmtest)
  library(splines)
  library(openxlsx)
  library(ggplot2)
})

out_dir <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/2"

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

# ================================
# Covariates used for PCA-based confounder assessment
# ================================

all_covariates_to_check <- c(
  "AGE",
  "SEX_BIRTH",
  "BMI_BASELINE",
  "SMOKING_CURRENT",
  "season_sin",
  "season_cos",
  "CENTER_RUMC",
  "TIMETOLAB",
  "CD4CD8_PRECART"
)

# ================================
# Continuous covariates only
# Linearity is assessed only for these
# ================================

continuous_covariates <- c(
  "AGE",
  "BMI_BASELINE",
  "season_sin",
  "season_cos",
  "TIMETOLAB",
  "CD4CD8_PRECART"
)

# ================================
# PCA on original Olink data
# Requires selected_columns_pca and pca_meta already created
# ================================

mypca <- PCA(selected_columns_pca, ncp = 10, graph = FALSE)

n_pc <- min(10, ncol(mypca$ind$coord))

pca_scores <- as.data.frame(mypca$ind$coord[, 1:n_pc, drop = FALSE])
colnames(pca_scores) <- paste0("PC", 1:n_pc)

analysis_df <- cbind(pca_scores, pca_meta)

all_covariates_to_check <- all_covariates_to_check[
  all_covariates_to_check %in% colnames(analysis_df)
]

continuous_covariates <- continuous_covariates[
  continuous_covariates %in% colnames(analysis_df)
]

# ================================
# Run assumption checks
# ================================

results <- list()
diagnostic_data <- list()

for (pc in paste0("PC", 1:n_pc)) {
  
  for (covar in all_covariates_to_check) {
    
    tmp <- analysis_df[, c(pc, covar)]
    tmp <- tmp[complete.cases(tmp), , drop = FALSE]
    
    if (nrow(tmp) < 20) next
    
    colnames(tmp) <- c("pc", "covar")
    
    tmp$pc <- as.numeric(tmp$pc)
    
    if (covar %in% continuous_covariates) {
      tmp$covar <- suppressWarnings(as.numeric(tmp$covar))
    } else {
      tmp$covar <- as.factor(tmp$covar)
    }
    
    tmp <- tmp[complete.cases(tmp), , drop = FALSE]
    
    if (nrow(tmp) < 20 || length(unique(tmp$covar)) < 2) next
    
    lm_fit <- tryCatch(
      lm(pc ~ covar, data = tmp),
      error = function(e) NULL
    )
    
    if (is.null(lm_fit)) next
    
    # Breusch-Pagan test for heteroscedasticity
    bp_p <- tryCatch(
      lmtest::bptest(lm_fit)$p.value,
      error = function(e) NA_real_
    )
    
    # Linearity check only for continuous covariates:
    # Compare linear model with natural spline model
    linearity_spline_p <- NA_real_
    
    if (covar %in% continuous_covariates && length(unique(tmp$covar)) >= 4) {
      
      spline_fit <- tryCatch(
        lm(pc ~ splines::ns(covar, df = 3), data = tmp),
        error = function(e) NULL
      )
      
      if (!is.null(spline_fit)) {
        linearity_spline_p <- tryCatch(
          anova(lm_fit, spline_fit)$`Pr(>F)`[2],
          error = function(e) NA_real_
        )
      }
    }
    
    results[[length(results) + 1]] <- data.frame(
      PC = pc,
      Covariate = covar,
      Variable_Type = ifelse(
        covar %in% continuous_covariates,
        "Continuous",
        "Categorical"
      ),
      N = nrow(tmp),
      Adj_R2 = summary(lm_fit)$adj.r.squared,
      BreuschPagan_P = bp_p,
      Linearity_Spline_P = linearity_spline_p
    )
    
    diagnostic_data[[length(diagnostic_data) + 1]] <- data.frame(
      PC = pc,
      Covariate = covar,
      Variable_Type = ifelse(
        covar %in% continuous_covariates,
        "Continuous",
        "Categorical"
      ),
      Fitted = fitted(lm_fit),
      Residual = residuals(lm_fit),
      Std_Residual = rstandard(lm_fit)
    )
  }
}

results_df <- bind_rows(results) %>%
  arrange(PC, Covariate)

diagnostics_df <- bind_rows(diagnostic_data)

# ================================
# Save assumption test results
# ================================

write.xlsx(
  results_df,
  file = file.path(out_dir, "PCA_Regression_Assumption_Checks.xlsx"),
  sheetName = "Assumption_Checks",
  overwrite = TRUE
)

# ================================
# Diagnostic plots
# Representative plots: PC1-PC5 only
# ================================

pcs_for_plots <- paste0("PC", 1:min(5, n_pc))

diagnostics_plot <- diagnostics_df %>%
  filter(PC %in% pcs_for_plots)

# ================================
# 1) Q-Q plots
# Answer: normality of residuals
# ================================

qq_plot <- ggplot(diagnostics_plot, aes(sample = Std_Residual)) +
  stat_qq(size = 0.35, alpha = 0.5) +
  stat_qq_line(linewidth = 0.4, color = "red") +
  facet_grid(PC ~ Covariate, scales = "free_y") +
  theme_bw(base_size = 8) +
  theme(
    strip.text.x = element_text(size = 6, angle = 90),
    strip.text.y = element_text(size = 8),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8),
    plot.title = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 9)
  ) +
  labs(
    title = "Q-Q plots of standardized residuals",
    subtitle = "Assesses normality; PCA regression models: PC score ~ covariate",
    x = "Theoretical quantiles",
    y = "Standardized residuals"
  )

ggsave(
  filename = file.path(out_dir, "PCA_Regression_QQ_Plots_Normality_PC1_to_PC5.pdf"),
  plot = qq_plot,
  width = 18,
  height = 10
)

ggsave(
  filename = file.path(out_dir, "PCA_Regression_QQ_Plots_Normality_PC1_to_PC5.tiff"),
  plot = qq_plot,
  width = 18,
  height = 10,
  dpi = 600,
  units = "in",
  device = "tiff"
)

# ================================
# 2) Residuals vs fitted plots
# Answer: homoscedasticity
# Include continuous + categorical covariates
# ================================

homoscedasticity_plot <- ggplot(diagnostics_plot, aes(x = Fitted, y = Residual)) +
  geom_point(size = 0.35, alpha = 0.5) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.4,
    color = "red"
  ) +
  facet_grid(PC ~ Covariate, scales = "free_y") +
  theme_bw(base_size = 8) +
  theme(
    strip.text.x = element_text(size = 6, angle = 90),
    strip.text.y = element_text(size = 8),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8),
    plot.title = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 9)
  ) +
  labs(
    title = "Residuals versus fitted values",
    subtitle = "Assesses homoscedasticity; PCA regression models: PC score ~ covariate",
    x = "Fitted values",
    y = "Residuals"
  )

ggsave(
  filename = file.path(out_dir, "PCA_Regression_Residuals_vs_Fitted_Homoscedasticity_PC1_to_PC5.pdf"),
  plot = homoscedasticity_plot,
  width = 18,
  height = 10
)

ggsave(
  filename = file.path(out_dir, "PCA_Regression_Residuals_vs_Fitted_Homoscedasticity_PC1_to_PC5.tiff"),
  plot = homoscedasticity_plot,
  width = 18,
  height = 10,
  dpi = 600,
  units = "in",
  device = "tiff"
)

# ================================
# 3) Linearity plots
# Answer: linearity
# Continuous covariates only
# Natural cubic spline smoother shown, not LOESS
# ================================

linearity_plot_df <- diagnostics_plot %>%
  filter(Covariate %in% continuous_covariates)

linearity_plot <- ggplot(linearity_plot_df, aes(x = Fitted, y = Residual)) +
  geom_point(size = 0.35, alpha = 0.5) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.4,
    color = "red"
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ splines::ns(x, df = 3),
    se = FALSE,
    linewidth = 0.5,
    color = "blue"
  ) +
  facet_grid(PC ~ Covariate, scales = "free_y") +
  theme_bw(base_size = 8) +
  theme(
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 8),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8),
    plot.title = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 9)
  ) +
  labs(
    title = "Residual trends for continuous covariates",
    subtitle = "Assesses linearity; natural cubic spline smoother shown",
    x = "Fitted values",
    y = "Residuals"
  )

ggsave(
  filename = file.path(out_dir, "PCA_Regression_Linearity_Continuous_Covariates_PC1_to_PC5.pdf"),
  plot = linearity_plot,
  width = 14,
  height = 10
)

ggsave(
  filename = file.path(out_dir, "PCA_Regression_Linearity_Continuous_Covariates_PC1_to_PC5.tiff"),
  plot = linearity_plot,
  width = 14,
  height = 10,
  dpi = 600,
  units = "in",
  device = "tiff"
)

# ================================
# Print results
# ================================

results_df
```

#DEPs in complete Cohort
Differential protein expression between R5 and X4 is tested using limma with empirical Bayes moderation, in the complete cohort first and then separately in discovery and validation subsets. The design matrix includes age, sex at birth, BMI, smoking status, seasonality (sine and cosine), genetic PC1, time to blood processing, pre-ART CD4:CD8 ratio, and recruitment centre. CD4CD8_PRECART has 258 missing values; after confirming a right-skewed distribution, missing values are imputed using k-nearest neighbours (k = 5) to avoid excluding approximately 17% of participants for a single covariate. The correlation between season terms and PANDEMIC_BEFOREAFTER exceeded 0.5; PANDEMIC_BEFOREAFTER was therefore dropped from the final model. Proteins are identified by column index between ADA_Inflammation and ZPR1_Neurology_II — verify these anchors if the dataset is updated.
```{r}
library(VIM)
library(ggplot2)
library(reshape2)
library(openxlsx)
library(stringr)
library(limma)



# Design matrix with covariates
cov_dat_de_complete_pr <- cov_dat_de_complete$condition
condition <-
  factor(cov_dat_de_complete_pr,
         levels = c("R5", "X4"))
#colnames(cov_dat_de_complete)

# Before removing the cases with NAs in any of the covariates let's check the number of samples we get before and after the removal of non-complete cases
dim(cov_dat_de_complete)
# [1] 1480 2665
cov_dat_de_complete[complete.cases(cov_dat_de_complete[, c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", "PANDEMIC_BEFOREAFTER", "season_sin", "season_cos", "CENTER_RUMC","PC1", "CD4CD8_PRECART",  "condition")]), ] %>% dim()
# [1] 1222 2665

# Let's check which variables contribute to the missingness
selected_vars <- c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", 
                   "PANDEMIC_BEFOREAFTER", "season_sin", "season_cos", 
                   "CENTER_RUMC", "PC1", "CD4CD8_PRECART", "condition")

na_counts_selected <- colSums(is.na(cov_dat_de_complete[selected_vars]))
print(na_counts_selected)
         #         AGE            SEX_BIRTH         BMI_BASELINE      SMOKING_CURRENT PANDEMIC_BEFOREAFTER           season_sin           season_cos 
         #           0                    0                    0                    0                    0                    0                    0 
         # CENTER_RUMC              PC1       CD4CD8_PRECART            condition 
         #           0                    0                  258                    0 

# All the missing values come from CD4CD8_PRECART, it might be not worth it to lose so many samples for just one variable, let's try to impute it.
# To impute it let's first check its distribution
ggplot(cov_dat_de_complete, aes(x = CD4CD8_PRECART)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density(aes(y = ..density.. * max(..count..)), color = "red", linewidth = 1) +
  theme_minimal() +
  labs(title = "Distribution of CD4CD8_PRECART",
       x = "CD4/CD8 Pre-CART",
       y = "Count") +
  theme(text = element_text(size = 14))
# Based on the histogram and density plot, the CD4CD8_PRECART variable appears to have a right-skewed distribution, with most values concentrated near zero and a long tail extending to higher values.
# Recommended imputations in these cases are: Median imputation (less sensitive to outliers) or kNN (k-Nearest Neighbors)
# Let's use kNN as it is a state of the ART
cov_dat_de_complete <- kNN(cov_dat_de_complete, variable = "CD4CD8_PRECART", k = 5)

# Let's check again for complete cases
cov_dat_de_complete[complete.cases(cov_dat_de_complete[, c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", "PANDEMIC_BEFOREAFTER", "season_sin", "season_cos", "CENTER_RUMC","PC1", "CD4CD8_PRECART", "condition")]), ] %>% dim()
# [1] 1480 2666

# Nice, we did not lose any samples as all the variables have now no missingness, let's continue.

#complete cases for covariate
cov_dat_de_complete <- cov_dat_de_complete[complete.cases(cov_dat_de_complete[, c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", "PANDEMIC_BEFOREAFTER", "season_sin", "season_cos", "CENTER_RUMC","PC1", "CD4CD8_PRECART",  "condition")]), ]

# Maybe before running the model let's check for multicollinearity, as a quick check we could check the correlation across variables
# Load necessary libraries

# Convert factor "condition" into numeric
cov_dat_de_complete$condition_numeric <- as.numeric(cov_dat_de_complete$condition)

cov_dat_de_complete |>
  dplyr::select(AGE, SEX_BIRTH, BMI_BASELINE, SMOKING_CURRENT,
                PANDEMIC_BEFOREAFTER, season_sin, season_cos,
                PC1, CD4CD8_PRECART, CENTER_RUMC, condition_numeric) |>
  sapply(class)

cov_dat_de |>
  dplyr::select(AGE, SEX_BIRTH, BMI_BASELINE, SMOKING_CURRENT,
                PANDEMIC_BEFOREAFTER, season_sin, season_cos,
                PC1, CD4CD8_PRECART, CENTER_RUMC) |>
  sapply(class)

# Calculate the correlation matrix for the specified variables
correlation_matrix <- cor(cov_dat_de_complete[, c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", 
                                                    "PANDEMIC_BEFOREAFTER", "season_sin", 
                                                    "season_cos", "PC1", "CD4CD8_PRECART", 
                                                    "CENTER_RUMC", "condition_numeric")], 
                          use = "pairwise.complete.obs")

# Melt the correlation matrix to long format
correlation_melted <- melt(correlation_matrix)

# Create the heatmap using ggplot2
ggplot(data = correlation_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), 
                       name = "Correlation") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) + # Add this line
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = "Variables", y = "Variables")

# It seems like seasonality and pandemic before and after are correlated (>50% correlation), we could try removing one (PANDEMIC_BEFOREAFTER), but after running the model
# it seems to make no difference after multiple testing correction, but is often preferable to choose the most simple model to avoid overcorrection
design_cov <-
  model.matrix(
    ~ AGE + SEX_BIRTH + BMI_BASELINE + SMOKING_CURRENT + season_sin + season_cos + PC1 + TIMETOLAB + CD4CD8_PRECART + CENTER_RUMC  + condition,
    data = cov_dat_de_complete
  )

dim(design_cov)
print(colnames(design_cov))
print(rownames(contrasts))
# Generate the protein expression matrix 
# Transform NPX data for proteins to represent rows - no group names
start_index <- which(colnames(cov_dat_de_complete) == "ADA_Inflammation")
end_index <- which(colnames(cov_dat_de_complete) == "ZPR1_Neurology_II")


cov_dat_proteins <- cov_dat_de_complete[, start_index:end_index]
cov_dat_de_complete_tr <- as.matrix(t(cov_dat_proteins))
dim(cov_dat_de_complete_tr)
#Fitting models for differentially expressed genes
cont.matrix_cov <- makeContrasts("conditionX4", levels = design_cov)
fit1_cov <- lmFit(cov_dat_de_complete_tr, design_cov)
fit2_cov <- contrasts.fit(fit1_cov, cont.matrix_cov)
fit2_cov <- eBayes(fit2_cov)

# Display "N" number of differentially expressed genes 
# Use this object for volcano plot 
DEP_cov <- topTable(fit2_cov, coef = "conditionX4", adjust = "BH", number = dim(cov_dat_de_complete_tr)[1]) 
DEP_cov$Cohort <- "Complete"

# Let's create a column with the protein names
DEP_cov$Protein <- str_split_i(rownames(DEP_cov), "_", i = 1)
DEP_cov$Panel <- str_split_i(rownames(DEP_cov), "_", i = 2)

#Display number of proteins significantly expressed in two groups
results_cov <- decideTests(fit2_cov)

# Display the number of UP and Down
summary(results_cov)
#  DiseasetvsHealthy
#Down                   4
#NotSig                20
#Up                    61

# Extract significant proteins
DEP_cov <- topTable(fit2_cov, coef = "conditionX4", adjust = "BH", number = dim(cov_dat_de_complete_tr)[1])

# Add cohort label
DEP_cov$Cohort <- "Complete"

# Extract protein name and panel
DEP_cov$Protein <- str_split_i(rownames(DEP_cov), "_", i = 1)  # Extract protein name
DEP_cov$Panel <- str_split_i(rownames(DEP_cov), "_", i = 2)    # Extract panel name

# Select relevant columns for export
DEP_cov_export <- DEP_cov[, c("Protein", "Panel", "P.Value", "adj.P.Val", "logFC", "t", "Cohort")]

# Rename columns for clarity
colnames(DEP_cov_export) <- c("Protein", "Short_Name", "P_Value", "Adj_P_Value", "Fold_Change", "t_stat" ,"Cohort")

# Print the **top 20 significant proteins**
print(head(DEP_cov_export, 20))

#Save as Excel file
write.xlsx(DEP_cov_export, file = "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Significant_Proteins_CompleteCohort.xlsx", rowNames = FALSE)


# Load necessary libraries
library(ggplot2)
library(ggrepel)

# Define significance thresholds
pval_threshold <- 0.05
adj_pval_threshold <- 0.05

# Volcano Plot using P.Value
volcano_pval <- ggplot(DEP_cov, aes(x = logFC, y = -log10(P.Value), label = Protein)) +
  geom_point(aes(color = P.Value < pval_threshold), size = 3) +
  scale_color_manual(values = c("black", "red")) +  # Red = Significant, Black = Not Significant
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(DEP_cov, P.Value < pval_threshold), size = 3, max.overlaps = Inf) +  
  theme_bw() +
  labs(
    title = "Volcano Plot - complete Cohort",
    x = "Log Fold Change (logFC)",
    y = "-log10(P.Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Display the plot
print(volcano_pval)

# Save the P.Value volcano plot
ggsave("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Volcano_Plot_PValue_CompleteCohort.pdf", plot = volcano_pval, width = 10, height = 10, dpi = 300)

# Volcano Plot using Adjusted P.Value
volcano_adj_pval <- ggplot(DEP_cov, aes(x = logFC, y = -log10(adj.P.Val), label = Protein)) +
  geom_point(aes(color = adj.P.Val < adj_pval_threshold), size = 3) +
  scale_color_manual(values = c("black", "red")) +
  geom_hline(yintercept = -log10(adj_pval_threshold), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(DEP_cov, adj.P.Val < adj_pval_threshold), size = 5, max.overlaps = Inf) +  
  theme_bw() +
  labs(
    title = "Volcano Plot - CompleteCohort Adjusted P.Value",
    x = "Log Fold Change (logFC)",
    y = "-log10(Adj.P.Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Display the plot
print(volcano_adj_pval)

# Save the Adjusted P.Value volcano plot
ggsave("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Volcano_Plot_AdjPValue_CompleteCohort.pdf", plot = volcano_adj_pval, width = 10, height = 10, dpi = 300)

# Select relevant participant-level data
participant_data <- cov_dat_de_complete[, c("ID", "Tropism",  "SEX_BIRTH", "COHORT")]

# Rename columns for clarity
colnames(participant_data) <- c("Participant_ID", "Tropism", "Sex_Birth", "cohort")

# Save participant data to Excel
write.xlsx(participant_data, 
           file = "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/CompleteCohort_Participants.xlsx", 
            rowNames = FALSE)

#Display number of proteins significantly expressed in two groups
results_cov <- decideTests(fit2_cov)

# Display the number of UP and Down
summary(results_cov)
#  DiseasetvsHealthy
#Down                   4
#NotSig                20
#Up                    61

```



#DEPs in Discovery cohort
```{r}
# Design matrix with covariates
cov_dat_de_discovery_pr <- cov_dat_de_discovery$condition
condition <-
  factor(cov_dat_de_discovery_pr,
         levels = c("R5", "X4"))
colnames(cov_dat_de_discovery)

# Let's use kNN as it is a state of the ART
cov_dat_de_discovery <- kNN(cov_dat_de_discovery, variable = "CD4CD8_PRECART", k = 5)

cov_dat_de_discovery <- cov_dat_de_discovery[complete.cases(cov_dat_de_discovery[, c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT","PANDEMIC_BEFOREAFTER", "season_sin", "season_cos",  "PC1", "CD4CD8_PRECART", "CENTER_RUMC", "condition")]), ]

design_cov <-
  model.matrix(
    ~ AGE+ SEX_BIRTH + BMI_BASELINE + SMOKING_CURRENT + season_sin + season_cos + PC1 + CD4CD8_PRECART + CENTER_RUMC + TIMETOLAB + condition,
    data = cov_dat_de_discovery
  )

dim(design_cov)
colnames(design_cov)

# Generate the protein expression matrix 
# Transform NPX data for proteins to represent rows - no group names

# Get indices from cov_dat_de_discovery)
start_index <- which(colnames(cov_dat_de_discovery) == "ADA_Inflammation")
end_index <- which(colnames(cov_dat_de_discovery) == "ZPR1_Neurology_II")


cov_dat_proteins <- cov_dat_de_discovery[, start_index:end_index]
cov_dat_de_discovery_tr <- as.matrix(t(cov_dat_proteins))
dim(cov_dat_de_discovery_tr)
#Fitting models for differentially expressed genes
fit1_cov <- lmFit(cov_dat_de_discovery_tr, design_cov)
cont.matrix_cov <- makeContrasts("conditionX4", levels = design_cov)
fit2_cov <- contrasts.fit(fit1_cov, cont.matrix_cov)
fit2_cov <- eBayes(fit2_cov)

# Display "N" number of differentially expressed genes 
# Use this object for volcano plot 
DEP_cov_discovery <- topTable(fit2_cov, coef = "conditionX4", adjust = "BH", number = dim(cov_dat_de_discovery_tr)[1]) 
DEP_cov_discovery$Cohort <- "Discovery"

# Let's create a column with the protein names
DEP_cov_discovery$Protein <- str_split_i(rownames(DEP_cov_discovery), "_", i = 1)
DEP_cov_discovery$Panel <- str_split_i(rownames(DEP_cov_discovery), "_", i = 2)

#Display number of proteins significantly expressed in two groups
results_cov <- decideTests(fit2_cov)

# Display the number of UP and Down
summary(results_cov)
#  DiseasetvsHealthy
#Down                   4
#NotSig                20
#Up                    61


library(openxlsx)
library(stringr)
library(limma)

# Extract significant proteins
DEP_cov_discovery <- topTable(fit2_cov, coef = "conditionX4", adjust = "BH", number = dim(cov_dat_de_discovery_tr)[1])

# Add cohort label
DEP_cov_discovery$Cohort <- "Discovery"

# Extract protein name and panel
DEP_cov_discovery$Protein <- str_split_i(rownames(DEP_cov_discovery), "_", i = 1)  # Extract protein name
DEP_cov_discovery$Panel <- str_split_i(rownames(DEP_cov_discovery), "_", i = 2)    # Extract panel name

# Select relevant columns for export
DEP_cov_discovery_export <- DEP_cov_discovery[, c("Protein", "Panel", "P.Value", "adj.P.Val", "logFC", "t", "Cohort")]

# Rename columns for clarity
colnames(DEP_cov_discovery_export) <- c("Protein", "Short_Name", "P_Value", "Adj_P_Value", "Fold_Change", "t_stat", "Cohort")

# Print the **top 20 significant proteins**
print(head(DEP_cov_discovery_export, 20))

# Save as Excel file
write.xlsx(DEP_cov_discovery_export, file = "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Significant_Proteins_DiscoveryCohort.xlsx", rowNames = FALSE)


# Load necessary libraries
library(ggplot2)
library(ggrepel)

# Define significance thresholds
pval_threshold <- 0.05
adj_pval_threshold <- 0.05

# Volcano Plot using P.Value
volcano_pval <- ggplot(DEP_cov_discovery, aes(x = logFC, y = -log10(P.Value), label = Protein)) +
  geom_point(aes(color = P.Value < pval_threshold), size = 3) +
  scale_color_manual(values = c("black", "red")) +  # Red = Significant, Black = Not Significant
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(DEP_cov_discovery, P.Value < pval_threshold), size = 3, max.overlaps = Inf) +  
  theme_bw() +
  labs(
    title = "Volcano Plot - DiscoveryCohort",
    x = "Log Fold Change (logFC)",
    y = "-log10(P.Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Display the plot
print(volcano_pval)

# Save the P.Value volcano plot
ggsave("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Volcano_Plot_PValue_DiscoveryCohort.pdf", plot = volcano_pval, width = 10, height = 10, dpi = 300)

# Volcano Plot using Adjusted P.Value
volcano_adj_pval <- ggplot(DEP_cov_discovery, aes(x = logFC, y = -log10(adj.P.Val), label = Protein)) +
  geom_point(aes(color = adj.P.Val < adj_pval_threshold), size = 3) +
  scale_color_manual(values = c("black", "red")) +
  geom_hline(yintercept = -log10(adj_pval_threshold), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(DEP_cov_discovery, adj.P.Val < adj_pval_threshold), size = 5, max.overlaps = Inf) +  
  theme_bw() +
  labs(
    title = "Volcano Plot - DiscoveryCohort Adjusted P.Value",
    x = "Log Fold Change (logFC)",
    y = "-log10(Adj.P.Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Display the plot
print(volcano_adj_pval)

# Save the Adjusted P.Value volcano plot
ggsave("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Volcano_Plot_AdjPValue_DiscoveryCohort.pdf", plot = volcano_adj_pval, width = 10, height = 10, dpi = 300)

# Select relevant participant-level data
participant_data <- cov_dat_de_discovery[, c("ID", "Tropism", "SEX_BIRTH", "COHORT")]

# Rename columns for clarity
colnames(participant_data) <- c("Participant_ID", "Tropism", "Sex_Birth", "cohort")

# Save participant data to Excel
write.xlsx(participant_data, 
           file = "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/DiscoveryCohort_Participants.xlsx", 
           rowNames = FALSE)

#Display number of proteins significantly expressed in two groups
results_cov <- decideTests(fit2_cov)

# Display the number of UP and Down
summary(results_cov)
#  DiseasetvsHealthy
#Down                   4
#NotSig                20
#Up                    61

```

# DEPs in Validation cohort
The same limma model is applied in the validation cohort independently. Significant proteins from the discovery cohort are evaluated for directional consistency and nominal significance in the validation cohort.
```{r}
# Design matrix with covariates
cov_dat_de_validation_pr <- cov_dat_de_validation$condition
condition <-
  factor(cov_dat_de_validation_pr,
         levels = c("R5", "X4"))
colnames(cov_dat_de_validation)

# Let's use kNN as it is a state of the ART
cov_dat_de_validation <- kNN(cov_dat_de_validation, variable = "CD4CD8_PRECART", k = 5)

cov_dat_de_validation <- cov_dat_de_validation[complete.cases(cov_dat_de_validation[, c("AGE", "SEX_BIRTH", "BMI_BASELINE", "SMOKING_CURRENT", "PANDEMIC_BEFOREAFTER", "season_sin", "season_cos", "CD4CD8_PRECART" , "CENTER_RUMC","PC1", "condition")]), ]

design_cov <-
  model.matrix(
    ~ AGE+ SEX_BIRTH + BMI_BASELINE + SMOKING_CURRENT + season_sin + season_cos + PC1  + CD4CD8_PRECART + TIMETOLAB  + condition,
    data = cov_dat_de_validation
  )

dim(design_cov)
colnames(design_cov)

# Generate the protein expression matrix 
# Transform NPX data for proteins to represent rows - no group names

# Get indices from cov_dat_de_validation
start_index <- which(colnames(cov_dat_de_validation) == "ADA_Inflammation")
end_index <- which(colnames(cov_dat_de_validation) == "ZPR1_Neurology_II")

cov_dat_proteins <- cov_dat_de_validation[, start_index:end_index]
cov_dat_de_validation_tr <- as.matrix(t(cov_dat_proteins))
dim(cov_dat_de_validation_tr)
#Fitting models for differentially expressed genes
fit1_cov <- lmFit(cov_dat_de_validation_tr, design_cov)
cont.matrix_cov <- makeContrasts("conditionX4", levels = design_cov)
fit2_cov <- contrasts.fit(fit1_cov, cont.matrix_cov)
fit2_cov <- eBayes(fit2_cov)

# Display "N" number of differentially expressed genes 
# Use this object for volcano plot 
DEP_cov_validation <- topTable(fit2_cov, coef = "conditionX4", adjust = "BH", number = dim(cov_dat_de_validation_tr)[1]) 
DEP_cov_validation$Cohort <- "Validation"

# Let's create a column with the protein names
DEP_cov_validation$Protein <- str_split_i(rownames(DEP_cov_validation), "_", i = 1)
DEP_cov_validation$Panel <- str_split_i(rownames(DEP_cov_validation), "_", i = 2)

#Display number of proteins significantly expressed in two groups
results_cov <- decideTests(fit2_cov)

# Display the number of UP and Down
summary(results_cov)
#  DiseasetvsHealthy
#Down                   4
#NotSig                20
#Up                    61


library(openxlsx)
library(stringr)
library(limma)

# Extract significant proteins
DEP_cov_validation <- topTable(fit2_cov, coef = "conditionX4", adjust = "BH", number = dim(cov_dat_de_validation_tr)[1])

# Add cohort label
DEP_cov_validation$Cohort <- "Validation"

# Extract protein name and panel
DEP_cov_validation$Protein <- str_split_i(rownames(DEP_cov_validation), "_", i = 1)  # Extract protein name
DEP_cov_validation$Panel <- str_split_i(rownames(DEP_cov_validation), "_", i = 2)    # Extract panel name



# Select relevant columns for export
DEP_cov_validation_export <- DEP_cov_validation[, c("Protein", "Panel", "P.Value", "adj.P.Val", "logFC", "t" , "Cohort")]

# Rename columns for clarity
colnames(DEP_cov_validation_export) <- c("Protein", "Short_Name", "P_Value", "Adj_P_Value", "Fold_Change", "t_stat", "Cohort")

# Print the **top 20 significant proteins**
print(head(DEP_cov_validation_export, 20))

# Save as Excel file
write.xlsx(DEP_cov_validation_export, file = "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Significant_Proteins_ValidationCohort.xlsx", rowNames = FALSE)


# Load necessary libraries
library(ggplot2)
library(ggrepel)

# Define significance thresholds
pval_threshold <- 0.05
adj_pval_threshold <- 0.05

# Volcano Plot using P.Value
volcano_pval <- ggplot(DEP_cov_validation, aes(x = logFC, y = -log10(P.Value), label = Protein)) +
  geom_point(aes(color = P.Value < pval_threshold), size = 3) +
  scale_color_manual(values = c("black", "red")) +  # Red = Significant, Black = Not Significant
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(DEP_cov_validation, P.Value < pval_threshold), size = 3, max.overlaps = Inf) +  
  theme_bw() +
  labs(
    title = "Volcano Plot - ValidationCohort",
    x = "Log Fold Change (logFC)",
    y = "-log10(P.Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Display the plot
print(volcano_pval)

# Save the P.Value volcano plot
ggsave("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Volcano_Plot_PValue_ValidationCohort.pdf", plot = volcano_pval, width = 10, height = 10, dpi = 300)

# Volcano Plot using Adjusted P.Value
volcano_adj_pval <- ggplot(DEP_cov_validation, aes(x = logFC, y = -log10(adj.P.Val), label = Protein)) +
  geom_point(aes(color = adj.P.Val < adj_pval_threshold), size = 3) +
  scale_color_manual(values = c("black", "red")) +
  geom_hline(yintercept = -log10(adj_pval_threshold), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(DEP_cov_validation, adj.P.Val < adj_pval_threshold), size = 5, max.overlaps = Inf) +  
  theme_bw() +
  labs(
    title = "Volcano Plot - ValidationCohort Adjusted P.Value",
    x = "Log Fold Change (logFC)",
    y = "-log10(Adj.P.Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Display the plot
print(volcano_adj_pval)

# Save the Adjusted P.Value volcano plot
ggsave("R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Volcano_Plot_AdjPValue_ValidationCohort.pdf", plot = volcano_adj_pval, width = 10, height = 10, dpi = 300)

# Select relevant participant-level data
participant_data <- cov_dat_de_validation[, c("ID", "Tropism", "SEX_BIRTH", "COHORT")]

# Rename columns for clarity
colnames(participant_data) <- c("Participant_ID", "Tropism", "Sex_Birth", "cohort")

# Save participant data to Excel
write.xlsx(participant_data, 
           file = "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/ValidationCohort_Participants.xlsx", 
           rowNames = FALSE)

#Display number of proteins significantly expressed in two groups
results_cov <- decideTests(fit2_cov)

# Display the number of UP and Down
summary(results_cov)
#  DiseasetvsHealthy
#Down                   4
#NotSig                20
#Up                    61


```



#common proteins in both cohorts
Discovery and validation DEP results are merged on protein name and classified into six groups based on directional consistency and significance thresholds: significant in both cohorts with consistent direction, significant in both with opposite direction, nominally significant in discovery with consistent direction in validation, discovery-only, validation-only, and not significant. The biplot displays log fold change in discovery (x-axis) against validation (y-axis), with the top 20 proteins from the "significant in discovery, same direction" category labelled. The full protein lists for this category, split by direction of effect, are printed to the console.
```{r}
# ===== Biplot: focus on "Significant in discovery (Same Direction)" =====
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggrepel)
  library(readxl)
})

disc_path <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Significant_Proteins_DiscoveryCohort.xlsx"
vali_path <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Significant_Proteins_ValidationCohort.xlsx"

DEP_cov_discovery  <- read_excel(disc_path)
DEP_cov_validation <- read_excel(vali_path)

# --- CONFIG ---
p_thresh <- 0.05
out_file <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/ManuscriptTropism/Biplot_logFC_Disc_vs_Val_DISC_SameDir_top30.pdf"

# --- Merge limma outputs (raw p only) ---
# Requires DEP_cov_discovery and DEP_cov_validation (topTable outputs)
logfc_comparison <- merge(
  DEP_cov_discovery[, c("Protein","Fold_Change","P_Value","Adj_P_Value")],
  DEP_cov_validation[, c("Protein","Fold_Change","P_Value", "Adj_P_Value")],
  by = "Protein",
  suffixes = c("_Discovery", "_Validation")
)

# --- Direction consistency flag ---
logfc_comparison <- logfc_comparison %>%
  mutate(Direction_Consistent = sign(Fold_Change_Discovery) == sign(Fold_Change_Validation))

# --- Grouping (raw p only) ---
logfc_comparison <- logfc_comparison %>%
  mutate(
    Group = case_when(
     Adj_P_Value_Discovery < p_thresh & P_Value_Validation < p_thresh & Direction_Consistent ~
        "Significant in Both (Same Direction)",
      Adj_P_Value_Discovery < p_thresh & P_Value_Validation < p_thresh & !Direction_Consistent ~
        "Significant in Both (Opposite Direction)",
      P_Value_Discovery < p_thresh & Direction_Consistent ~
        "Significant in discovery (Same Direction)",
      Adj_P_Value_Discovery < p_thresh & !(P_Value_Validation < p_thresh) ~
        "Discovery Only",
      !(Adj_P_Value_Discovery < p_thresh) & P_Value_Validation < p_thresh ~
        "Validation Only",
      TRUE ~ "Not Significant"
    ),
    Group = factor(
      Group,
      levels = c(
        "Significant in Both (Same Direction)",
        "Significant in Both (Opposite Direction)",
        "Significant in discovery (Same Direction)",
        "Discovery Only",
        "Validation Only",
        "Not Significant"
      )
    )
  )

# ===== Focus category =====
focus_cat <- "Significant in discovery (Same Direction)"

disc_same_df <- logfc_comparison %>%
  filter(as.character(Group) == focus_cat) %>%
  mutate(abs_fc = abs(Fold_Change_Discovery))

# Print ALL proteins in the category (and UP/DOWN splits)
cat("\n===== ALL proteins in category: ", focus_cat, " =====\n", sep = "")
if (nrow(disc_same_df)) {
  print(disc_same_df$Protein)
} else {
  cat("None\n")
}

all_up_disc_same <- disc_same_df %>%
  filter(Fold_Change_Discovery > 0) %>%
  arrange(P_Value_Discovery, desc(abs_fc)) %>%
  pull(Protein)

all_down_disc_same <- disc_same_df %>%
  filter(Fold_Change_Discovery < 0) %>%
  arrange(P_Value_Discovery, desc(abs_fc)) %>%
  pull(Protein)

cat("\n----- UP (Discovery FC > 0) -----\n")
if (length(all_up_disc_same)) print(all_up_disc_same) else cat("None\n")

cat("\n----- DOWN (Discovery FC < 0) -----\n")
if (length(all_down_disc_same)) print(all_down_disc_same) else cat("None\n")

# ===== Labels: TOP 30 from the focus category (by p then |FC|, any direction) =====
label_df <- disc_same_df %>%
  arrange(P_Value_Discovery, desc(abs_fc)) %>%
  slice_head(n = 20) %>%
  mutate(Label = Protein)

cat("\n===== Labeled (Top 30 within category) =====\n")
if (nrow(label_df)) print(label_df$Protein) else cat("None (category empty)\n")

# ===== Plot =====
biplot_dynamic <- ggplot(logfc_comparison, aes(x = Fold_Change_Discovery, y = Fold_Change_Validation, color = Group)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  # label only the selected Top 30 from the focus category
  geom_text_repel(
    data = label_df,
    inherit.aes = FALSE,
    aes(x = Fold_Change_Discovery, y = Fold_Change_Validation, label = Label),
    size = 5, color = "black",
    max.overlaps = Inf, box.padding = 0.4, min.segment.length = 0
  ) +
  scale_color_manual(values = c(
    "Significant in Both (Same Direction)"       = "hotpink3",
    "Significant in Both (Opposite Direction)"   = "darkorange",
    "Significant in discovery (Same Direction)"  = "red",
    "Discovery Only"                             = "salmon",
    "Validation Only"                            = "dodgerblue",
    "Not Significant"                            = "gray70"
  )) +
  labs(
    title = "Biplot of Fold_Change: Discovery vs Validation",
    subtitle = paste0("Labels: Top 30 within category \"", focus_cat, "\" (ranked by p then |FC|); printing ALL in category + UP/DOWN lists"),
    x = "Log Fold Change (Discovery)",
    y = "Log Fold Change (Validation)",
    color = "Significance Group"
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.line  = element_line(),      
    axis.ticks = element_line())       

print(biplot_dynamic)

ggsave(filename = out_file, plot = biplot_dynamic, width = 10, height = 10, dpi = 600)



```

#voilin plots known inflammation markers
Targeted visualisation of 15 pre-selected immune-relevant proteins (CCL5, CD14, CD163, CXCL10, CXCL9, GZMB, IFNG, IL1B, IL6, OSM, TNF, CCL17, CCL22, IL10, TGFB1) using raw NPX values. Note that this chunk reloads the three input files independently rather than reusing the merged dataset from the first chunk; this is intentional to ensure the raw NPX values are used without any prior filtering or imputation applied to the main analysis dataset. Pairwise comparisons use Wilcoxon rank-sum test with global BH correction across all proteins within each cohort.
```{r}
# ===== RAW NPX: R5 vs X4 only — Medians + Global BH across proteins; ONE Excel sheet =====
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(openxlsx)
  library(rstatix)    # pairwise_wilcox_test
  library(ggpubr)     # stat_pvalue_manual
  library(colorspace)
})

# --------- CONFIG ---------
out_dir        <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/Coreceptor"
excel_out      <- file.path(out_dir, "RAW_NPX_R5vsX4_Medians_and_Pairwise.xlsx")
plot_disc_file <- file.path(out_dir, "Plots_NPX_R5vsX4_Discovery.pdf")
plot_vali_file <- file.path(out_dir, "Plots_NPX_R5vsX4_Validation.pdf")
plot_both_file <- file.path(out_dir, "Plots_NPX_R5vsX4_BothCohorts.pdf")

# Exact colors
custom_colors <- c("R5" = "aquamarine3", "X4" = "coral")

# Proteins of interest (deduplicated; keep order)
protein_symbols <- c(
  "CCL5", "CD14", "CD163", "CXCL10", "CXCL9", "GZMB", "IFNG", "IL1B", "IL6",  "OSM",  "TNF",  "CCL17",  "CCL22",  "IL10",  "TGFB1"
)
# --------- LOAD DATA ---------
cov_dat_de_master <- openxlsx::read.xlsx(
  "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/updated_OLINK_Explore3072_1910samples_2367proteins_after_bridging_normalization_QCed_21nov2022.xlsx"
) %>% na.omit()

clinical_df <- openxlsx::read.xlsx(
  "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx"
)

# --------- BUILD RAW DF (R5/X4 only; drop Mixed/NA); convert -Inf -> NA ---------
raw_df <- clinical_df %>%
  inner_join(cov_dat_de_master, by = "ID") %>%  # Olink NPX
  mutate(
    SMOKING_CURRENT = as.numeric(SMOKING_CURRENT),
    Tropism = factor(Tropism, levels = c("R5", "X4")),
    COHORT  = factor(COHORT,  levels = c("DISCOVERY", "VALIDATION"))
  ) %>%
  filter(!is.na(Tropism), Tropism %in% c("R5", "X4"))

raw_df[raw_df == -Inf] <- NA
# --------- MAP SYMBOLS -> NPX COLUMNS (handles symbols with multiple panels) ---------
find_protein_cols <- function(symbols, df) {
  npx_cols <- colnames(df)
  out <- vapply(symbols, function(sym){
    hits <- grep(paste0("^", sym, "_"), npx_cols, value = TRUE)
    if (length(hits) == 0) return(NA_character_)
    paste(hits, collapse = ";")
  }, character(1))
  tibble(Protein = symbols, NPX_cols = out)
}

map_tbl <- find_protein_cols(protein_symbols, raw_df)
if (any(is.na(map_tbl$NPX_cols))) {
  message("Missing NPX columns for: ",
          paste(map_tbl$Protein[is.na(map_tbl$NPX_cols)], collapse = ", "))
}
map_tbl_expanded <- map_tbl %>%
  tidyr::separate_rows(NPX_cols, sep = ";") %>%
  filter(!is.na(NPX_cols) & NPX_cols != "")

# --------- LONG RAW NPX TABLE FOR TARGET PROTEINS ---------
long_raw <- raw_df %>%
  dplyr::select(ID, Tropism, COHORT, dplyr::all_of(unique(map_tbl_expanded$NPX_cols))) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(unique(map_tbl_expanded$NPX_cols)),
    names_to = "Protein_Panel",
    values_to = "NPX"
  ) %>%
  dplyr::mutate(
    Protein = sub("_.*$", "", Protein_Panel),
    Panel   = sub("^[^_]+_", "", Protein_Panel)
  ) %>%
  dplyr::filter(Protein %in% protein_symbols, !is.na(NPX))

# Split by cohort
disc_df <- long_raw %>% filter(COHORT == "DISCOVERY")
vali_df <- long_raw %>% filter(COHORT == "VALIDATION")

# --------- MEDIANS + PAIRWISE (R5 vs X4 only) ---------
compute_medians <- function(df_cohort) {
  df_cohort %>%
    group_by(Protein, Tropism) %>%
    summarise(
      n = dplyr::n(),
      median_NPX = median(NPX, na.rm = TRUE),
      IQR_NPX    = IQR(NPX, na.rm = TRUE),
      .groups = "drop"
    )
}

pairwise_R5_X4 <- function(df_cohort) {
  present <- df_cohort %>%
    count(Protein, Tropism) %>%
    filter(n > 0) %>%
    group_by(Protein) %>%
    filter(all(c("R5","X4") %in% Tropism)) %>%
    ungroup() %>%
    distinct(Protein)

  df_ok <- df_cohort %>% semi_join(present, by = "Protein")
  if (nrow(df_ok) == 0) {
    return(tibble(
      Protein = character(), group1 = character(), group2 = character(),
      p = numeric(), p_adj = numeric(), p_label = character()
    ))
  }

  pw_raw <- df_ok %>%
    group_by(Protein) %>%
    rstatix::pairwise_wilcox_test(NPX ~ Tropism, p.adjust.method = "BH") %>%
    ungroup() %>%
    filter(group1 %in% "R5" & group2 %in% "X4")

  # Standardize within-protein adjusted p column name -> p_adj
  std_padj <- function(d) {
    cand <- c("p_adj","p.adj","padj","Adjusted_P_value","adj_P.Val","adj_P","p_adjust")
    hit  <- cand[cand %in% names(d)]
    if (length(hit) == 0) d$p_adj <- NA_real_ else d$p_adj <- d[[hit[1]]]
    d
  }

  std_padj(pw_raw) %>%
    transmute(
      Protein,
      group1 = as.character(group1),
      group2 = as.character(group2),
      p      = p,
      p_adj  = p_adj
    )
}

disc_medians <- compute_medians(disc_df)
vali_medians <- compute_medians(vali_df)
disc_pw      <- pairwise_R5_X4(disc_df)
vali_pw      <- pairwise_R5_X4(vali_df)

# --------- GLOBAL BH across proteins (per cohort) ---------
disc_pw <- disc_pw %>%
  mutate(
    Global_FDR = p.adjust(p, method = "BH"),
    p_label    = formatC(Global_FDR, format = "f", digits = 3)
  )

vali_pw <- vali_pw %>%
  mutate(
    Global_FDR = p.adjust(p, method = "BH"),
    p_label    = formatC(Global_FDR, format = "f", digits = 3)
  )

# --------- ONE-SHEET SUMMARY (includes Global_FDR) ---------
make_export_table <- function(cohort_name, med_tbl, pw_tbl) {
  need_cols <- c("Protein","Tropism","n","median_NPX","IQR_NPX")
  stopifnot(all(need_cols %in% names(med_tbl)))
  need_pw <- c("Protein","group1","group2","p","p_adj","Global_FDR","p_label")
  stopifnot(all(need_pw %in% names(pw_tbl)))

  med_tbl <- tibble::as_tibble(med_tbl)
  pw_tbl  <- tibble::as_tibble(pw_tbl)

  gsum <- med_tbl %>%
    transmute(
      Protein,
      Group  = Tropism,
      N      = n,
      Median = median_NPX,
      IQR    = IQR_NPX
    )

  out <- pw_tbl %>%
    dplyr::left_join(gsum, by = c("Protein" = "Protein", "group1" = "Group")) %>%
    dplyr::rename(N_Group1 = N, Median_Group1 = Median, IQR_Group1 = IQR) %>%
    dplyr::left_join(gsum, by = c("Protein" = "Protein", "group2" = "Group")) %>%
    dplyr::rename(N_Group2 = N, Median_Group2 = Median, IQR_Group2 = IQR) %>%
    mutate(
      Dataset          = cohort_name,
      Column           = Protein,
      P_value          = p,
      Adjusted_P_value = p_adj,        # within-protein (trivial here)
      Global_FDR       = Global_FDR,   # across proteins in this cohort
      Test             = "Wilcoxon rank-sum; BH across proteins"
    ) %>%
    dplyr::select(
      Dataset, Column, group1, group2,
      N_Group1, N_Group2, Median_Group1, Median_Group2,
      IQR_Group1, IQR_Group2, P_value, Adjusted_P_value,
      Global_FDR, Test
    ) %>%
    arrange(Column, group1, group2)

  colnames(out) <- c(
    "Dataset","Column","Group1","Group2",
    "N_Group1","N_Group2","Median_Group1","Median_Group2",
    "IQR_Group1","IQR_Group2","P_value","Adjusted_P_value",
    "Global_FDR","Test"
  )
  out
}

disc_export <- make_export_table("Discovery",  disc_medians, disc_pw)
vali_export <- make_export_table("Validation", vali_medians, vali_pw)
one_sheet   <- bind_rows(disc_export, vali_export)

# Write ONE sheet
wb <- createWorkbook()
addWorksheet(wb, "Medians_Pairwise_R5vsX4")
writeData(wb, "Medians_Pairwise_R5vsX4", one_sheet)
addStyle(wb, "Medians_Pairwise_R5vsX4", createStyle(textDecoration = "Bold"),
         rows = 1, cols = 1:ncol(one_sheet), gridExpand = TRUE)
saveWorkbook(wb, excel_out, overwrite = TRUE)

# --------- BRACKET POSITIONS (per Protein; R5–X4 only) + LABELS USE GLOBAL FDR ---------
make_pairwise_anno <- function(df_cohort, pw_tbl) {
  if (nrow(df_cohort) == 0 || nrow(pw_tbl) == 0) return(tibble())

  rng <- df_cohort %>%
    group_by(Protein) %>%
    summarise(
      y_top = as.numeric(quantile(NPX, probs = 0.98, na.rm = TRUE)),
      y_min = min(NPX, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(step = pmax(0.06 * (y_top - y_min + 1e-6), 0.15))

  pos <- rng %>%
    transmute(Protein, group1 = "R5", group2 = "X4", y.position = y_top + step*1)

  # add label from pw_tbl (Global_FDR)
  pos %>%
    dplyr::left_join(
  dplyr::select(as.data.frame(pw_tbl), Protein, group1, group2, Global_FDR),
  by = c("Protein", "group1", "group2")
) %>%
    mutate(label = formatC(Global_FDR, format = "f", digits = 3))
}

disc_pw_anno <- make_pairwise_anno(disc_df, disc_pw)
vali_pw_anno <- make_pairwise_anno(vali_df, vali_pw)

# --------- PLOTS (labels show Global_FDR) ---------
make_violin_plot <- function(df_plot, title_txt, anno_df, strip_size = 18) {
  ggplot(df_plot, aes(x = Tropism, y = NPX, fill = Tropism)) +
    geom_violin(trim = FALSE, alpha = 0.4) +
    geom_jitter(aes(color = Tropism), width = 0.2, alpha = 0.9, size = 0.5) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, color = "black") +
    stat_summary(
      fun = median, geom = "text",
      aes(label = round(..y.., 1)),
      vjust = -0.7, size = 3, color = "black"
    ) +
    scale_fill_manual(values = custom_colors, drop = FALSE) +
    scale_color_manual(values = custom_colors, drop = FALSE) +
    labs(title = title_txt, x = "", y = "NPX (raw)") +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(size = 14, color = "black"),
      axis.text.y = element_text(size = 14, color = "black"),
      strip.text  = element_text(size = strip_size, face = "bold"),
      axis.line.x = element_line(color = "black", linewidth = 0.6),
      axis.line.y = element_line(color = "black", linewidth = 0.6),
      axis.ticks  = element_line(color = "black")
    ) +
    ggpubr::stat_pvalue_manual(
      data         = anno_df,
      label        = "label",            # numeric Global_FDR
      y.position   = "y.position",
      xmin         = "group1",
      xmax         = "group2",
      tip.length   = 0,
      bracket.size = 0.35,
      size         = 6,
      hide.ns      = FALSE,
      inherit.aes  = FALSE
    ) +
    facet_wrap(~ Protein, scales = "free_y", ncol = 5)
}

plot_discovery <- make_violin_plot(
  df_plot   = disc_df,
  title_txt = "RAW NPX by Tropism — Discovery cohort (R5 vs X4)",
  anno_df   = disc_pw_anno,
  strip_size = 18
)

plot_validation <- make_violin_plot(
  df_plot   = vali_df,
  title_txt = "RAW NPX by Tropism — Validation cohort (R5 vs X4)",
  anno_df   = vali_pw_anno,
  strip_size = 18
)

# --- Combined plot (no brackets), then enlarge strip text ---
plot_both <- ggplot(long_raw, aes(x = Tropism, y = NPX, fill = Tropism)) +
  geom_violin(trim = FALSE, alpha = 0.4) +
  geom_jitter(aes(color = Tropism), width = 0.2, alpha = 0.9, size = 0.5) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, color = "black") +
  stat_summary(
    fun = median, geom = "text",
    aes(label = round(..y.., 1)),
    vjust = -0.7, size = 3, color = "black"
  ) +
  scale_fill_manual(values = custom_colors, drop = FALSE) +
  scale_color_manual(values = custom_colors, drop = FALSE) +
  labs(
    title = "RAW NPX by Tropism — Discovery vs Validation (R5 vs X4)",
    x = "", y = "NPX (raw)"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    strip.text  = element_text(face = "bold"),
    axis.line.x = element_line(color = "black", linewidth = 0.6),
    axis.line.y = element_line(color = "black", linewidth = 0.6),
    axis.ticks  = element_line(color = "black")
  ) +
  facet_grid(COHORT ~ Protein, scales = "free_y")

plot_both <- plot_both +
  theme(
    strip.text.x = element_text(size = 18, face = "bold"),
    strip.text.y = element_text(size = 18, face = "bold")
  )

# --------- SAVE ---------
ggsave(filename = plot_disc_file, plot = plot_discovery, width = 14, height = 9, dpi = 300)
ggsave(filename = plot_vali_file, plot = plot_validation, width = 14, height = 9, dpi = 300)
ggsave(filename = plot_both_file, plot = plot_both,     width = 16, height = 10, dpi = 300)

message("Excel (one sheet) -> ", excel_out)
message("Plots ->\n  ", plot_disc_file, "\n  ", plot_vali_file, "\n  ", plot_both_file)


```

#Sensitivity analysis using parametric test
As a sensitivity analysis, the Wilcoxon tests above are repeated using Welch two-sample t-tests to assess robustness of findings under a parametric assumption. Olink NPX values are approximately normally distributed by design, making the parametric test defensible. The mean difference is computed as X4 minus R5 throughout; the t.test confidence interval direction depends on factor level ordering in R and is retained as returned. Global BH correction is applied across all proteins within each cohort separately.
```{r}
# ===== RAW NPX: R5 vs X4 only — PARAMETRIC Welch t-test + Global BH =====

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(openxlsx)
  library(ggpubr)
  library(colorspace)
})

# --------- CONFIG ---------
out_dir <- "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/Coreceptor"

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

excel_out      <- file.path(out_dir, "RAW_NPX_R5vsX4_WelchTtest_GlobalBH.xlsx")
plot_disc_file <- file.path(out_dir, "Plots_NPX_R5vsX4_Discovery_WelchTtest.pdf")
plot_vali_file <- file.path(out_dir, "Plots_NPX_R5vsX4_Validation_WelchTtest.pdf")
plot_both_file <- file.path(out_dir, "Plots_NPX_R5vsX4_BothCohorts_WelchTtest.pdf")

custom_colors <- c(
  "R5" = "aquamarine3",
  "X4" = "coral"
)

protein_symbols <- c(
  "CCL5", "CD14", "CD163", "CXCL10", "CXCL9",
  "GZMB", "IFNG", "IL1B", "IL6", "OSM",
  "TNF", "CCL17", "CCL22", "IL10", "TGFB1"
)


# --------- LOAD DATA ---------
cov_dat_de_master <- openxlsx::read.xlsx(
  "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/2000HIV_VROLINK2/updated_OLINK_Explore3072_1910samples_2367proteins_after_bridging_normalization_QCed_21nov2022.xlsx"
)

clinical_df <- openxlsx::read.xlsx(
  "R:/hcrc/Personal folders/Kavita Mehta/2000HIV/clinical data/KAVITA_IMPUT_DATAFRAME_SMOKING_CURRENT_RESOLVED.xlsx"
)

# --------- BUILD RAW DF ---------
raw_df <- clinical_df %>%
  dplyr::inner_join(cov_dat_de_master, by = "ID") %>%
  dplyr::mutate(
    SMOKING_CURRENT = as.numeric(SMOKING_CURRENT),
    Tropism = factor(Tropism, levels = c("R5", "X4")),
    COHORT  = factor(COHORT,  levels = c("DISCOVERY", "VALIDATION"))
  ) %>%
  dplyr::filter(
    !is.na(Tropism),
    Tropism %in% c("R5", "X4")
  ) %>%
  dplyr::mutate(
    dplyr::across(
      where(is.numeric),
      ~ replace(., is.infinite(.), NA)
    )
  )

# --------- MAP SYMBOLS TO NPX COLUMNS ---------
find_protein_cols <- function(symbols, df) {

  npx_cols <- colnames(df)

  out <- vapply(symbols, function(sym) {
    hits <- grep(paste0("^", sym, "_"), npx_cols, value = TRUE)
    if (length(hits) == 0) return(NA_character_)
    paste(hits, collapse = ";")
  }, character(1))

  tibble::tibble(
    Protein = symbols,
    NPX_cols = out
  )
}

map_tbl <- find_protein_cols(protein_symbols, raw_df)

if (any(is.na(map_tbl$NPX_cols))) {
  message(
    "Missing NPX columns for: ",
    paste(map_tbl$Protein[is.na(map_tbl$NPX_cols)], collapse = ", ")
  )
}

map_tbl_expanded <- map_tbl %>%
  tidyr::separate_rows(NPX_cols, sep = ";") %>%
  dplyr::filter(!is.na(NPX_cols), NPX_cols != "")

# --------- LONG RAW NPX TABLE ---------
long_raw <- raw_df %>%
  dplyr::select(
    ID,
    Tropism,
    COHORT,
    dplyr::all_of(unique(map_tbl_expanded$NPX_cols))
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(unique(map_tbl_expanded$NPX_cols)),
    names_to = "Protein_Panel",
    values_to = "NPX"
  ) %>%
  dplyr::mutate(
    Protein = sub("_.*$", "", Protein_Panel),
    Panel   = sub("^[^_]+_", "", Protein_Panel),
    NPX     = suppressWarnings(as.numeric(NPX))
  ) %>%
  dplyr::filter(
    Protein %in% protein_symbols,
    !is.na(NPX),
    !is.infinite(NPX)
  )

disc_df <- long_raw %>%
  dplyr::filter(COHORT == "DISCOVERY")

vali_df <- long_raw %>%
  dplyr::filter(COHORT == "VALIDATION")

# --------- SUMMARY STATISTICS ---------
compute_summary <- function(df_cohort) {

  df_cohort %>%
    dplyr::group_by(Protein, Tropism) %>%
    dplyr::summarise(
      n = dplyr::n(),
      mean_NPX   = mean(NPX, na.rm = TRUE),
      sd_NPX     = sd(NPX, na.rm = TRUE),
      median_NPX = median(NPX, na.rm = TRUE),
      IQR_NPX    = IQR(NPX, na.rm = TRUE),
      .groups = "drop"
    )
}

# --------- PARAMETRIC TEST: WELCH T-TEST ---------
pairwise_R5_X4_parametric <- function(df_cohort) {

  present <- df_cohort %>%
    dplyr::count(Protein, Tropism) %>%
    dplyr::filter(n > 1) %>%
    dplyr::group_by(Protein) %>%
    dplyr::filter(all(c("R5", "X4") %in% Tropism)) %>%
    dplyr::ungroup() %>%
    dplyr::distinct(Protein)

  df_ok <- df_cohort %>%
    dplyr::semi_join(present, by = "Protein")

  if (nrow(df_ok) == 0) {
    return(tibble::tibble(
      Protein = character(),
      group1 = character(),
      group2 = character(),
      p = numeric(),
      estimate_diff_mean = numeric(),
      conf_low = numeric(),
      conf_high = numeric()
    ))
  }

  df_ok %>%
    dplyr::group_by(Protein) %>%
    dplyr::group_modify(~ {

      dat <- .x

      mean_R5 <- mean(dat$NPX[dat$Tropism == "R5"], na.rm = TRUE)
      mean_X4 <- mean(dat$NPX[dat$Tropism == "X4"], na.rm = TRUE)

      tt <- tryCatch(
        t.test(NPX ~ Tropism, data = dat, var.equal = FALSE),
        error = function(e) NULL
      )

      if (is.null(tt)) {
        return(tibble::tibble(
          group1 = "R5",
          group2 = "X4",
          p = NA_real_,
          estimate_diff_mean = mean_X4 - mean_R5,
          conf_low = NA_real_,
          conf_high = NA_real_
        ))
      }

      # R's t.test CI direction can depend on group order.
      # Here p-value is reliable; mean difference is manually forced as X4 - R5.
      tibble::tibble(
        group1 = "R5",
        group2 = "X4",
        p = tt$p.value,
        estimate_diff_mean = mean_X4 - mean_R5,
        conf_low = tt$conf.int[1],
        conf_high = tt$conf.int[2]
      )
    }) %>%
    dplyr::ungroup()
}

disc_summary <- compute_summary(disc_df)
vali_summary <- compute_summary(vali_df)

disc_pw <- pairwise_R5_X4_parametric(disc_df)
vali_pw <- pairwise_R5_X4_parametric(vali_df)

# --------- GLOBAL BH ACROSS PROTEINS WITHIN EACH COHORT ---------
disc_pw <- disc_pw %>%
  dplyr::mutate(
    Global_FDR = p.adjust(p, method = "BH"),
    p_label = dplyr::case_when(
      is.na(Global_FDR) ~ "NA",
      Global_FDR < 0.0001 ~ "<0.0001",
      TRUE ~ formatC(Global_FDR, format = "g", digits = 2)
    )
  )

vali_pw <- vali_pw %>%
  dplyr::mutate(
    Global_FDR = p.adjust(p, method = "BH"),
    p_label = dplyr::case_when(
      is.na(Global_FDR) ~ "NA",
      Global_FDR < 0.0001 ~ "<0.0001",
      TRUE ~ formatC(Global_FDR, format = "g", digits = 2)
    )
  )

# --------- EXPORT TABLE ---------
make_export_table <- function(cohort_name, sum_tbl, pw_tbl) {

  gsum <- sum_tbl %>%
    dplyr::transmute(
      Protein,
      Group = Tropism,
      N = n,
      Mean = mean_NPX,
      SD = sd_NPX,
      Median = median_NPX,
      IQR = IQR_NPX
    )

  out <- pw_tbl %>%
    dplyr::left_join(
      gsum,
      by = c("Protein" = "Protein", "group1" = "Group")
    ) %>%
    dplyr::rename(
      N_Group1 = N,
      Mean_Group1 = Mean,
      SD_Group1 = SD,
      Median_Group1 = Median,
      IQR_Group1 = IQR
    ) %>%
    dplyr::left_join(
      gsum,
      by = c("Protein" = "Protein", "group2" = "Group")
    ) %>%
    dplyr::rename(
      N_Group2 = N,
      Mean_Group2 = Mean,
      SD_Group2 = SD,
      Median_Group2 = Median,
      IQR_Group2 = IQR
    ) %>%
    dplyr::mutate(
      Dataset = cohort_name,
      Column = Protein,
      P_value = p,
      Global_FDR = Global_FDR,
      P_label = p_label,
      Test = "Welch two-sample t-test; BH FDR across proteins within cohort"
    ) %>%
    dplyr::select(
      Dataset, Column, group1, group2,
      N_Group1, N_Group2,
      Mean_Group1, SD_Group1,
      Mean_Group2, SD_Group2,
      Median_Group1, IQR_Group1,
      Median_Group2, IQR_Group2,
      estimate_diff_mean, conf_low, conf_high,
      P_value, Global_FDR, P_label, Test
    ) %>%
    dplyr::arrange(Column)

  colnames(out) <- c(
    "Dataset", "Column", "Group1", "Group2",
    "N_Group1", "N_Group2",
    "Mean_Group1", "SD_Group1",
    "Mean_Group2", "SD_Group2",
    "Median_Group1", "IQR_Group1",
    "Median_Group2", "IQR_Group2",
    "Mean_difference_X4_minus_R5",
    "CI_low_from_ttest", "CI_high_from_ttest",
    "P_value", "Global_FDR", "P_label", "Test"
  )

  out
}

disc_export <- make_export_table("Discovery", disc_summary, disc_pw)
vali_export <- make_export_table("Validation", vali_summary, vali_pw)

one_sheet <- dplyr::bind_rows(disc_export, vali_export)

wb <- openxlsx::createWorkbook()
openxlsx::addWorksheet(wb, "WelchTtest_R5vsX4")
openxlsx::writeData(wb, "WelchTtest_R5vsX4", one_sheet)

openxlsx::addStyle(
  wb,
  "WelchTtest_R5vsX4",
  openxlsx::createStyle(textDecoration = "Bold"),
  rows = 1,
  cols = 1:ncol(one_sheet),
  gridExpand = TRUE
)

openxlsx::saveWorkbook(wb, excel_out, overwrite = TRUE)

# --------- BRACKET POSITIONS ---------
make_pairwise_anno <- function(df_cohort, pw_tbl) {

  if (nrow(df_cohort) == 0 || nrow(pw_tbl) == 0) {
    return(tibble::tibble())
  }

  pw_tbl <- as.data.frame(pw_tbl)

  rng <- df_cohort %>%
    dplyr::group_by(Protein) %>%
    dplyr::summarise(
      y_top = as.numeric(quantile(NPX, probs = 0.98, na.rm = TRUE)),
      y_min = min(NPX, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      step = pmax(0.06 * (y_top - y_min + 1e-6), 0.15)
    )

  rng %>%
    dplyr::transmute(
      Protein,
      group1 = "R5",
      group2 = "X4",
      y.position = y_top + step
    ) %>%
    dplyr::left_join(
      dplyr::select(
        pw_tbl,
        Protein, group1, group2, Global_FDR, p_label
      ),
      by = c("Protein", "group1", "group2")
    ) %>%
    dplyr::mutate(
      label = p_label
    )
}

disc_pw_anno <- make_pairwise_anno(disc_df, disc_pw)
vali_pw_anno <- make_pairwise_anno(vali_df, vali_pw)

# --------- PLOTS ---------
make_violin_plot <- function(df_plot, title_txt, anno_df, strip_size = 18) {

  base_plot <- ggplot(df_plot, aes(x = Tropism, y = NPX, fill = Tropism)) +
    geom_violin(trim = FALSE, alpha = 0.4) +
    geom_jitter(
      aes(color = Tropism),
      width = 0.2,
      alpha = 0.9,
      size = 0.5
    ) +
    geom_boxplot(
      width = 0.2,
      outlier.shape = NA,
      alpha = 0.5,
      color = "black"
    ) +
    stat_summary(
      fun = mean,
      geom = "text",
      aes(label = round(after_stat(y), 1)),
      vjust = -0.7,
      size = 3,
      color = "black"
    ) +
    scale_fill_manual(values = custom_colors, drop = FALSE) +
    scale_color_manual(values = custom_colors, drop = FALSE) +
    labs(
      title = title_txt,
      x = "",
      y = "NPX"
    ) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(size = 14, color = "black"),
      axis.text.y = element_text(size = 14, color = "black"),
      strip.text = element_text(size = strip_size, face = "bold"),
      axis.line.x = element_line(color = "black", linewidth = 0.6),
      axis.line.y = element_line(color = "black", linewidth = 0.6),
      axis.ticks = element_line(color = "black")
    ) +
    facet_wrap(~ Protein, scales = "free_y", ncol = 5)

  if (nrow(anno_df) > 0) {
    base_plot <- base_plot +
      ggpubr::stat_pvalue_manual(
        data = anno_df,
        label = "label",
        y.position = "y.position",
        xmin = "group1",
        xmax = "group2",
        tip.length = 0,
        bracket.size = 0.35,
        size = 6,
        hide.ns = FALSE,
        inherit.aes = FALSE
      )
  }

  base_plot
}

plot_discovery <- make_violin_plot(
  df_plot = disc_df,
  title_txt = "NPX by Tropism — Discovery cohort, Welch t-test R5 vs X4",
  anno_df = disc_pw_anno,
  strip_size = 18
)

plot_validation <- make_violin_plot(
  df_plot = vali_df,
  title_txt = "NPX by Tropism — Validation cohort, Welch t-test R5 vs X4",
  anno_df = vali_pw_anno,
  strip_size = 18
)

plot_both <- ggplot(long_raw, aes(x = Tropism, y = NPX, fill = Tropism)) +
  geom_violin(trim = FALSE, alpha = 0.4) +
  geom_jitter(
    aes(color = Tropism),
    width = 0.2,
    alpha = 0.9,
    size = 0.5
  ) +
  geom_boxplot(
    width = 0.2,
    outlier.shape = NA,
    alpha = 0.5,
    color = "black"
  ) +
  stat_summary(
    fun = mean,
    geom = "text",
    aes(label = round(after_stat(y), 1)),
    vjust = -0.7,
    size = 3,
    color = "black"
  ) +
  scale_fill_manual(values = custom_colors, drop = FALSE) +
  scale_color_manual(values = custom_colors, drop = FALSE) +
  labs(
    title = "NPX by Tropism — Discovery vs Validation",
    x = "",
    y = "NPX"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    strip.text = element_text(face = "bold"),
    axis.line.x = element_line(color = "black", linewidth = 0.6),
    axis.line.y = element_line(color = "black", linewidth = 0.6),
    axis.ticks = element_line(color = "black")
  ) +
  facet_grid(COHORT ~ Protein, scales = "free_y") +
  theme(
    strip.text.x = element_text(size = 18, face = "bold"),
    strip.text.y = element_text(size = 18, face = "bold")
  )

# --------- SAVE ---------
ggsave(
  filename = plot_disc_file,
  plot = plot_discovery,
  width = 14,
  height = 9,
  dpi = 300
)

ggsave(
  filename = plot_vali_file,
  plot = plot_validation,
  width = 14,
  height = 9,
  dpi = 300
)

ggsave(
  filename = plot_both_file,
  plot = plot_both,
  width = 16,
  height = 10,
  dpi = 300
)

message("Excel -> ", excel_out)
message("Plots ->\n  ", plot_disc_file, "\n  ", plot_vali_file, "\n  ", plot_both_file)
```



